### 2013-05-30 praglearn_d1.tex
draft from Mike

\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}
%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09

\title{Learning words through recursive pragmatic reasoning about other agents}

\author{
Nathaniel Smith\thanks{...} \\
School of Informatics\\
University of Edinburgh\\
Edinburgh, Scotland, UK\\
\texttt{nsmith23@inf.ed.ac.uk} \\
\AND
Noah D. Goodman \\
Department of Psychology\\
Stanford University \\
Stanford, CA, USA \\
\texttt{ngoodman@stanford.edu} \\
\And
Michael C. Frank \\
Department of Psychology \\
Stanford University \\
Stanford, CA, USA \\
\texttt{mcfrank@stanford.edu}}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

% \nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\begin{abstract}
Adult language users are remarkably good at making inferences about speakers' intentions in context, and children learning their native language also display substantial pragmatic skill in acquiring the meanings of unknown words. While pragmatic inference and word learning have both been characterized in probabilistic terms, no current models make appropriate inferences about the meanings of words in pragmatically-rich inferential contexts. We describe a model in which language learners assume that they jointly approximate a shared, external lexicon and reason recursively about the goals of others in using this lexicon. This model captures phenomena in word learning and pragmatic inference; it additionally leads to insights about the emergence of communicative systems in conversation and the mechanisms by which pragmatic inferences become incorporated into word meanings. 
\end{abstract}

\section{Introduction}

Natural language is a powerful tool for conveying meaning in context. Our utterances are pervasively underspecified. Consider ``it's raining,'' ``I ate some of the cookies,'' or ``can you close the window?'' In each, a listener must go beyond the literal meaning of the words to fill in contextual details (``it's raining here and now''), infer that a stronger alternative is not true (``I ate some but not all of the cookies''), or more generally infer the speaker's communicative goal (``I want you to close the window right now because I'm cold.'').

Following this intuition, theories of pragmatic communication frame the process of language comprehension as inference about the generating goal of an utterance given a rational speaker \cite{grice1975,dale1995,frank2012}. For example, a listener might reason, ``if she had wanted me to think `all' of the cookies, she would have said `all'---but she didn't. Hence `all' must not be true and she must have eaten some {\it but not all} of the cookies.'' This kind of counterfactual reasoning has a recursive character to it: listener models speaker (who may in turn be modeling a listener).

Even before they are proficient language users, children learn the meanings of words from their pattern of use in context. For example, in demonstrations of the ``disambiguation effect,'' a novel word (e.g. ``dax'') is used in a context containing both a novel and a familiar object. From quite young, children tend to make the inference that ``dax'' refers to the novel object \cite{markman1988}. In another demonstration, even young toddlers can extract mappings between words and objects from a pattern of individually-ambiguous co-occurrences between pairs of words and pairs of objects \cite{smith2008}. Taking inspiration from children's learning prowess, recent work in the field of grounded language learning has explored the possibility that even relatively complex word meanings can be inferred from corpus data (e.g. \cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).

Recursive pragmatic inferences such as those described above could be an important part of language learning.\footnote{Very young children make inferences that are often labeled as ``pragmatic'' in that they involve reasoning about context \cite{clark1988,baldwin1993}, though they also show weaknesses in specific inferences (e.g. strengthening ``some'' to ``some but not all'' \cite{papafragou2003}). Here we remain agnostic about the age at which children are able to make such inferences robustly, as it may vary depending on the linguistic materials being used in the inference \cite{barner2011}.} For example, the disambiguation inference described above can emerge from such reasoning (``why would the speaker have said ``dax'' if there was a word for the known object that she knew I would know'') \cite{clark1988}. In addition, there are a variety of intriguing cases in which listeners seem to create situation- and task-specific ways of referring to particular objects. For example, when asked to refer to idiosyncratic geometric shapes, over the course of an experimental session, participants create conventionalized descriptions that allow them to perform accurately even though they do not begin with a shared label \cite{krauss1964,clark1986}. Yet models of grounded language learning have failed to take pragmatic factors into account.

The goal of our current work is to investigate the possibilities for integrating models of recursive pragmatic reasoning with models of language learning, with the hope of capturing phenomena in both domains. We begin by describing previous models in each domain as well as proposals for bringing the two together. We next simulate findings from the domains of pragmatic inference, word learning, and their intersection (summarized in Table \ref{tab:results}). We end by discussing the implications of our results for work on language change.

\section{Models}

Basic idea of iterated pragmatic reasoning models. $S_i$ defines a causal model for utterances described by a simple Bayes net. $L_{i+1}$ does standard rational inference over this network. But the likelihood function in $S_i$'s model is very complicated, and includes equations derived from considering how a hypothetical $L_{i-1}$ would interpret utterances if they were doing rational inference about another hypothetical causal agent $S_{i-2}$.

To simplify matters we envision a simple world that has $w$ words and $b$ objects. The \textit{lexicon} defines the literal meaning of each word, ...

\newcommand{\word}{\text{word}}
\newcommand{\obj}{\text{object}}
\begin{align*}
P_{L_0}(\obj | \word) &= \text{lexicon}(\word, \obj) P_{\text{prior}}(\obj | \text{context}). \\
P_{S_i}(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_{i - 1}}(\obj | \word) - \text{cost}(\word)\big)\Big) \\
P_{L_i}(\obj | \word) &\propto P_{S_{i-1}}(\word | \obj) P_{\text{prior}}(\obj | \text{context})
\end{align*}

For the rest of this paper we somewhat arbitrarily fix the softmax parameter to $\lambda = 3$.

\subsection{Naive literal learner model}

To see how 
Make $L_0$ uncertain about values in $\text{lexicon}(\word, \obj)$, impose a prior, and then

Needs feedback

can't switch roles from speaker to listener

knocks out the lexical uncertainty/horn implicature results (because you can only have uncertainty at one level)

something wrong about taking evidence from the pragmatic speaker and using it to train the literal listener

\subsection{"Social anxiety" model}

\section{Experiments}

\begin{table}[t]
\caption{Table of empirical results and references}
\label{tab:results}
\begin{center}
\begin{tabular}{llllll}
\hline \\
Phenomenon & Ref. & Word learning \cite{frank2009} & Pragmatics \cite{frank2012,bergen2012}  & Model 1 & Model 2 \\
\hline
Scalar implicature & \cite{stiller2011} & & & & \\
Horn implicature & & & & & \\
Cross-situational learning & \cite{yu2007} & & & & \\
Disambiguation & \cite{markman1988} & & & & \\
Communicative equilibria & \cite{galantucci2007} & & & & \\
\end{tabular}
\end{center}
\end{table}

\subsection{Pragmatic reasoning}

\subsubsection{Scalar implicature}

\subsubsection{Horn implicature}

\subsection{Word learning}

\subsubsection{Statistical cross-situational word learning}

\subsubsection{Disambiguation using known words}

\subsection{Lexical learning in pragmatic contexts}

\subsubsection{Emergence of communicative equilibria}

\begin{figure}
\centering
\includegraphics[width=5.5in]{figures/galantucci3.png}
\caption{\label{fig:glantucci} Simulations....} 
\end{figure}

\subsubsection{Lexicalization of Horn implicatures}

\begin{figure}
\centering
\includegraphics[width=5.5in]{figures/horn.png}
\caption{\label{fig:horn} Simulations.} 
\end{figure}

\subsubsection{Lexicaliation and preservation of scalar implicature}

\begin{figure}
\centering
\includegraphics[width=2in]{figures/pragmatic_scalar.png}
\includegraphics[width=2in]{figures/nonpragmatic_scalar.png}
\caption{\label{fig:scalar} Simulations..} 
\end{figure}

\section{Conclusion}

%\subsubsection*{Acknowledgments}

% Thanks to ONR Grant #blahblah

\bibliographystyle{plain}
\bibliography{pragmatics}

\end{document}

### 2013-05-30 figures/ME.py
first pass at adding proper figures code

@@ -1,0 +1,68 @@
+# Mutual exclusivity
+
+# 2 words, 2 objects.
+#
+# Word 0 is known to refer to object 0. Word 1's meaning is totally unknown.
+
+from nips import *
+
+d = dom(adjectives=2, objects=2)
+lexicon_prior = [[5, 1], [1, 1]]
+listener = conpact2.FixedSupportImportanceSampler(d, 0, PARTICLES * 10,
+                                                  lexicon_prior=lexicon_prior)
+def show_ME_posterior(listener):
+    show_lex_posterior(listener.weighted_lexicons(), (0, 0), (1, 0),
+                       xlabel="P(Familiar word literally means familiar object)",
+                       ylabel="P(Novel word literally means familiar object)")
+show_ME_posterior(listener)
+
+# Then we hear someone use the novel word. On average we think this word might
+# mean anything, so on average it might refer to either, so on average it acts
+# like "some" or "glasses" and gets pragmatically strengthened into referring
+# to the novel object.
+
+print "Novel word means: Familiar object: %0.1f%%. Novel object: %0.1f%%." % tuple(100 * np.exp(listener.marginal_dist_n(LISTENER_DEPTH)[1, :]))
+
+# *But* we don't know what was meant. And if it meant the novel object, then
+# *that could be because the literal meaning is ambiguous; and if they meant
+# *the dog, then that would imply that this novel word does refer to dogs. So
+# *on net, we actually end up slightly more convinced that the novel word
+# *refers to the familiar object!
+
+show_ME_posterior(listener.with_data([conpact2.LDataUtt(LISTENER_DEPTH - 1, 1)]))
+
+# That's for hearing one example. If we hear both words said multiple times in
+# this same context, though, then the best explanation is that they're being
+# used contrastively, and so we start believing that actually the novel word
+# does literally refer to the novel object. Critically, though, it's hearing
+# the other, *familiar* word that provides the crucial information here:
+
+show_ME_posterior(listener.with_data([conpact2.LDataUtt(LISTENER_DEPTH - 1, 1), conpact2.LDataUtt(LISTENER_DEPTH - 1, 0)] * 10))
+
+# That's for if we have absolutely no prior bias towards or away from
+# sparsity. Either is plausible, though -- intuitively if the novel object is
[…]

### 2013-05-30 figures/conpact2.py

@@ -1,0 +1,978 @@
+# SNAPSHOT of the conpact2.py model code used for NIPS 2013 submission
+# See pragmods repository for the "official" version
+
+# Second attempt to model conceptual pacts via bayesian pragmatics.
+
+# We imagine a scenario in which a speaker describes an object using a
+# sequence of adjectives. The literal speaker S(-1) just samples from a simple
+# multinomial lexicon. Specifically, they:
+#  - pick an object in the world to describe, according to some prior.
+#  - decide how many adjectives to use. This is always observed, so we could
+#    put a prior over it but there would be no point.
+#  - sample that many times from a multinomial P(word|object). This
+#    distribution is called the "lexicon", and is assumed to vary with the
+#    context.
+# This serves as the base case for the standard probabilistic-game-theoretic
+# recursion, with S-1 defined as above, Ln doing rational bayesian inference
+# on S(n-1), and Sn sampling utterances from a cost-penalized softmax of
+# L(n-1). (NB: this code uses the convention of disjoint indexes for speakers
+# and listeners; odd indices are listeners and even indices are speakers.)
+#
+# We label the above S-1, and make the optimal listener L0, because in
+# practice it's a lot easier to reason about words as having meanings than
+# about objects as having words. One can write down the literal lexicon in
+# either form P(word|obj) or P(obj|word); we choose the latter
+# representation. This means that our lexicon actually contains some poorly
+# interpreted quantity like "lexical affinity" or "appropriateness" or
+# something once we start dealing with multiword utterances, and it's sort of
+# arbitrary that we require they sum to one. But whatever; the above provides
+# a nice justification, and this works.
+#
+# So that's our lexicon. Then, we have uncertainty about the lexicon. The
+# above determines the joint distribution for utterances and objects for Si/Li
+# who *know* what lexicon to use. In fact, the listener is uncertain, so to
+# interpret their input, they marginalize over all the ways that some listener
+# Ln might have interpreted the input
+#    sum_lex Pn(obj|utt, lex) P(lex|data)
+# And the speaker is uncertain, so to choose what to say, they marginalize
+# over all the ways that some listener might interpret their input
+#    sum_lex P(n-1)(obj|utt, lex) P(lex|data)
+# and they do a softmax from the resulting marginalized distribution.
+#
+# Both speaker and listener use a dirichlet prior on each word's meaning, the
[…]

### 2013-05-30 figures/cross-sit.py

@@ -1,0 +1,50 @@
+# Cross-situational learning
+
+# Data available to learning:
+#   Two objects are present in each situation, one is always missing
+#   Either, one of the two objects is named, by producing its word; or else, the distractor is produced
+
+# Word 0 -> object 0
+# Word 1 -> object 1
+# Word 2 -> object 2
+# and Word 3 is a distractor.
+#
+# Our learning is invariant to the order of presentation, so there's no point
+# in simulating lots of long dialogues... the only thing that matters is the
+# relative proportions of different examples, and there the law of large
+# numbers will kick in. So we just evaluate the posterior once.
+#
+# NOTE: this example really pushes the limits of the importance sampling
+# (notice that the below graph used 100,000 particles, and still has visible
+# asymmetries caused by the sampling; when I tried it with 10,000 particles it
+# got the wrong answer! 100,000 particles takes a few minutes on my
+# laptop.). This is a perfect place to use MCMC instead.
+
+from nips import *
+
+CS_OBJECTS = 3
+CS_WORDS = 4
+data_points = []
+for missing_obj in xrange(CS_OBJECTS):
+    present_objs = range(CS_OBJECTS)
+    present_objs.remove(missing_obj)
+    object_prior = np.empty(CS_OBJECTS)
+    object_prior[missing_obj] = 0.01
+    object_prior[present_objs] = (1 - 0.01) / 2
+    for target_obj in present_objs:
+        data_points.append(conpact2.LDataUtt(LISTENER_DEPTH - 1,
+                                             target_obj,
+                                             np.log(object_prior)))
+    for distractor_word in xrange(CS_OBJECTS, CS_WORDS):
+        data_points.append(conpact2.LDataUtt(LISTENER_DEPTH - 1,
+                                             distractor_word,
+                                             np.log(object_prior)))
+
+d = dom(adjectives=4, objects=3)
+learner = conpact2.FixedSupportImportanceSampler(d, 23, PARTICLES * 1,
+                                                 data=data_points * 10)
+matshow(np.exp(learner.marginal_dist_n(LISTENER_DEPTH)), cmap="binary",
+        vmin=0, vmax=1)
+xlabel("Objects")
+ylabel("Words")
+gcf.savefig("cross-sit.pdf")

### 2013-05-30 figures/emergence.py

@@ -1,0 +1,23 @@
+from nips import *
+
+# Galantucci emergence
+
+# XX TODO: calculate some statistics on P(convergence) by running a few
+# hundred dialogues
+#
+# XX mention: speaker model of listener being a bit noisy is fine, because
+# speaker does an extra layer of pragmatic recursion to clean things up (which
+# is why accuracy is so high even when speaker model is noiser than listener
+# model)
+#
+# XX mention: point out that the speaker and listener actually have totally
+# different, possibly contradictory data that they're learning from -- but it
+# works out in the end
+
+d = dom(adjectives=2, objects=2)
+dialogues = simulate_dialogues(4, 10, d)
+show_dialogues("emergence2x2-%i.pdf", dialogues)
+
+d = dom(adjectives=3, objects=3)
+dialogues = simulate_dialogues(4, 10, d, particle_count=PARTICLES * 10)
+show_dialogues("emergence3x3-%i.pdf", dialogues)

### 2013-05-30 figures/horn.py

@@ -1,0 +1,29 @@
+from nips import *
+
+# Horn implicature alone
+
+# A trivial version of Leon's common/rare cheap/expensive domain
+bgl_domain = dom(adjectives=2, objects=2,
+                 word_cost=[0.5, 1.0],
+                 default_object_prior=[0.8, 0.2])
+dialogue = conpact2.Dialogue(bgl_domain, LISTENER_DEPTH, 0, PARTICLES)
+
+print "How speaker refers to \"common\" and \"rare\" objects:"
+print np.exp(dialogue.uncertain_s_dist())
+print "How listener interprets \"cheap\" and \"expensive\" words:"
+print np.exp(dialogue.uncertain_l_dist())
+
+# Emergence + Horn implicature
+
+# XX TODO: calculate some statistics on P(convergence to good system) by
+# running a few hundred dialogues
+#
+# XX mention: this is a very general result, can speculate about it as a
+# partial cause of things like common words being shorter, rise of useful
+# social conventions in general
+#
+# XX mention: Horn implicature can't survive as implicatures. And connect to
+# scalar implicature, which *can* survive.
+
+bgl_dialogues = simulate_dialogues(4, 20, bgl_domain)
+show_dialogues("horn-emergence-%i.pdf", bgl_dialogues)

### 2013-05-30 figures/nips.py

@@ -1,0 +1,124 @@
+# -*- coding: utf-8 -*-
+
+import conpact2
+import numpy as np
+# We need ~this as a weight for the Horn implicature emergence to work, so let's just use it everywhere
+SOFTMAX_WEIGHT = 3
+# Resolution of the importance sampler
+PARTICLES = 1000
+# Listener is L2-with-uncertainty, speaker is S3-with-uncertainty
+LISTENER_DEPTH = 2
+CONVERGENCE_FIG_WIDTH = 20
+
+import matplotlib
+matplotlib.use("Cairo")
+matplotlib.rcParams["xtick.direction"] = "out"
+matplotlib.rcParams["ytick.direction"] = "out"
+from matplotlib.pyplot import *
+import matplotlib.pyplot as plt
+
+# Use a fixed max_utterance_length and softmax weight for everything
+# With length-1 utterances we could get away with using actual sampling for posteriors...
+def dom(**kwargs):
+    return conpact2.MemoizedDomain(max_utterance_length=1, softmax_weight=SOFTMAX_WEIGHT, **kwargs)
+
+def show_lex_posterior(path, weights_particles, idx1, idx2,
+                       labels=True, xlabel=None, ylabel=None):
+    weights, particles = weights_particles
+    masses, x_edges, y_edges = np.histogram2d([p[idx1] for p in particles],
+                                              [p[idx2] for p in particles],
+                                              range=[[0, 1], [0, 1]],
+                                              weights=np.exp(weights))
+    f = figure()
+    imshow(masses.T, aspect="equal",
+           extent=(x_edges[0], x_edges[-1], y_edges[0], y_edges[-1]),
+           origin="lower", vmin=0, cmap="binary", interpolation="nearest")
+    if labels:
+        if xlabel is None:
+            plt.xlabel("P(word %s means obj %s | data)" % idx1)
+        else:
+            plt.xlabel(xlabel)
+        if ylabel is None:
+            plt.ylabel("P(word %s means obj %s | data)" % idx2)
+        else:
+            plt.ylabel(ylabel)
+    savefig(path)
+    close(f)
+
+def simulate_dialogues(count, turns_per_dialogue, domain, lexicon_prior=None,
+                       listener_depth=LISTENER_DEPTH, particle_seed_base=0, particle_count=PARTICLES):
[…]

### 2013-05-30 figures/scalar-implic.py

@@ -1,0 +1,46 @@
+from nips import *
+
+# Demonstrating scalar implicature
+
+some_all_domain = dom(adjectives=2, objects=2)
+some_all_prior = [[10, 10], [1, 10]]
+dialogue = conpact2.Dialogue(some_all_domain, LISTENER_DEPTH, 0, PARTICLES, lexicon_prior=some_all_prior)
+
+print "How speaker refers to \"some\" and \"all\" objects:"
+print np.exp(dialogue.uncertain_s_dist())
+print "How listener interprets \"some\" and \"all\" words:"
+print np.exp(dialogue.uncertain_l_dist())
+
+# Learning scalar implicature
+
+# two objects: call them "partial x" (xxxooo) and "full x" (xxxxxx)
+#
+# two words: "some" and "all"
+
+d = dom(adjectives=2, objects=2)
+listener = conpact2.FixedSupportImportanceSampler(d, 0, PARTICLES * 10)
+
+# Contexts where "partial" and "full" meanings are a priori equiprobable
+some_for_partial_not_full = conpact2.LDataUttWithMeaning(LISTENER_DEPTH - 1, 0, 0)
+all_for_full_not_partial = conpact2.LDataUttWithMeaning(LISTENER_DEPTH - 1, 1, 1)
+# Contexts where either "partial" or "full" is a priori overwhelmingly likely
+some_for_partial_alone = conpact2.LDataUttWithMeaning(LISTENER_DEPTH - 1, 0, 0, log_object_prior=np.log([0.99, 0.01]))
+
+def show_scalar_posterior(path, listener):
+    show_lex_posterior(path, listener.weighted_lexicons(), (0, 0), (1, 0),
+                       xlabel="P(Some means partial, not full)",
+                       ylabel="P(All means partial, not full)")
+
+# Only pragmatically strengthened examples:
+
+show_scalar_posterior("some-all-only-pragmatic.pdf",
+                      listener.with_data([some_for_partial_not_full,
+                                          all_for_full_not_partial] * 5))
+
+# A mix of pragmatically strengthened examples, and examples where the meaning
+# from context is obviously 'partial', and 'some' gets used to describe it.
+
+show_scalar_posterior("some-all-pragmatic+unambiguous.pdf",
+                      listener.with_data([some_for_partial_not_full,
+                                          all_for_full_not_partial,
+                                          some_for_partial_alone] * 5))

### 2013-05-30 praglearn.tex
start hacking figures into shape

--- 2013-05-30 praglearn_d1.tex
@@
 \usepackage{graphicx}
-%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09
 
@@
 Edinburgh, Scotland, UK\\
-\texttt{nsmith23@inf.ed.ac.uk} \\
+\texttt{nathaniel.smith@ed.ac.uk} \\
 \AND
@@
 
-To simplify matters we envision a simple world that has $w$ words and $b$ objects. The \textit{lexicon} defines the literal meaning of each word, ...
+To simplify matters we envision a simple world that has $k$ words and $l$ objects. The \textit{lexicon} defines the literal meaning of each word, ...
+
+extensional as marginalization over intensional
 
@@
 
-\subsection{"Social anxiety" model}
+\subsection{``Social anxiety'' model}
 
@@
 \centering
-\includegraphics[width=5.5in]{figures/galantucci3.png}
-\caption{\label{fig:glantucci} Simulations....} 
+\includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
+\includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
+\includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
+\includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
+\caption{\label{fig:emergence} Simulations....}
 \end{figure}
@@
 \centering
-\includegraphics[width=5.5in]{figures/horn.png}
+\includegraphics[width=0.48\textwidth]{figures/horn-emergence-0.pdf}
+\includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
 \caption{\label{fig:horn} Simulations.}
@@
 \centering
-\includegraphics[width=2in]{figures/pragmatic_scalar.png}
-\includegraphics[width=2in]{figures/nonpragmatic_scalar.png}
+\includegraphics[width=2in]{figures/some-all-only-pragmatic.pdf}
+\includegraphics[width=2in]{figures/some-all-pragmatic+unambiguous.pdf}
 \caption{\label{fig:scalar} Simulations..}

### 2013-05-30 figures/ME.py

@@ -6,22 +6,25 @@
 
 from nips import *
 
+f = open("ME.log", "w")
+
 d = dom(adjectives=2, objects=2)
 lexicon_prior = [[5, 1], [1, 1]]
 listener = conpact2.FixedSupportImportanceSampler(d, 0, PARTICLES * 10,
                                                   lexicon_prior=lexicon_prior)
-def show_ME_posterior(listener):
-    show_lex_posterior(listener.weighted_lexicons(), (0, 0), (1, 0),
+def show_ME_posterior(path, listener):
+    show_lex_posterior(path, listener.weighted_lexicons(), (0, 0), (1, 0),
                        xlabel="P(Familiar word literally means familiar object)",
                        ylabel="P(Novel word literally means familiar object)")
-show_ME_posterior(listener)
+show_ME_posterior("ME-prior.pdf", listener)
 
 # Then we hear someone use the novel word. On average we think this word might
 # mean anything, so on average it might refer to either, so on average it acts
 # like "some" or "glasses" and gets pragmatically strengthened into referring
 # to the novel object.
 
-print "Novel word means: Familiar object: %0.1f%%. Novel object: %0.1f%%." % tuple(100 * np.exp(listener.marginal_dist_n(LISTENER_DEPTH)[1, :]))
+f.write("Novel word means: Familiar object: %0.1f%%. Novel object: %0.1f%%.\n"
+        % tuple(100 * np.exp(listener.marginal_dist_n(LISTENER_DEPTH)[1, :])))
 
 # *But* we don't know what was meant. And if it meant the novel object, then
 # *that could be because the literal meaning is ambiguous; and if they meant
@@ -29,7 +32,7 @@ print "Novel word means: Familiar object: %0.1f%%. Novel object: %0.1f%%." % tup
 # *on net, we actually end up slightly more convinced that the novel word
 # *refers to the familiar object!
 
-show_ME_posterior(listener.with_data([conpact2.LDataUtt(LISTENER_DEPTH - 1, 1)]))
+show_ME_posterior("ME-1dax.pdf", listener.with_data([conpact2.LDataUtt(LISTENER_DEPTH - 1, 1)]))
 
 # That's for hearing one example. If we hear both words said multiple times in
[…]

### 2013-05-30 figures/cross-sit.py

@@ -41,10 +41,10 @@ for missing_obj in xrange(CS_OBJECTS):
                                              np.log(object_prior)))
 
 d = dom(adjectives=4, objects=3)
-learner = conpact2.FixedSupportImportanceSampler(d, 23, PARTICLES * 1,
+learner = conpact2.FixedSupportImportanceSampler(d, 23, PARTICLES * 100,
                                                  data=data_points * 10)
 matshow(np.exp(learner.marginal_dist_n(LISTENER_DEPTH)), cmap="binary",
         vmin=0, vmax=1)
 xlabel("Objects")
 ylabel("Words")
-gcf.savefig("cross-sit.pdf")
+gcf().savefig("cross-sit.pdf", bbox_inches="tight")

### 2013-05-30 figures/horn.py

@@ -1,5 +1,7 @@
 from nips import *
 
+f = open("horn.log", "w")
+
 # Horn implicature alone
 
 # A trivial version of Leon's common/rare cheap/expensive domain
@@ -8,10 +10,10 @@ bgl_domain = dom(adjectives=2, objects=2,
                  default_object_prior=[0.8, 0.2])
 dialogue = conpact2.Dialogue(bgl_domain, LISTENER_DEPTH, 0, PARTICLES)
 
-print "How speaker refers to \"common\" and \"rare\" objects:"
-print np.exp(dialogue.uncertain_s_dist())
-print "How listener interprets \"cheap\" and \"expensive\" words:"
-print np.exp(dialogue.uncertain_l_dist())
+f.write("How speaker refers to \"common\" and \"rare\" objects:\n")
+f.write(repr(np.exp(dialogue.uncertain_s_dist())))
+f.write("\n\nHow listener interprets \"cheap\" and \"expensive\" words:\n")
+f.write(repr(np.exp(dialogue.uncertain_l_dist())))
 
 # Emergence + Horn implicature
 

### 2013-05-30 figures/nips.py

@@ -14,6 +14,11 @@ import matplotlib
 matplotlib.use("Cairo")
 matplotlib.rcParams["xtick.direction"] = "out"
 matplotlib.rcParams["ytick.direction"] = "out"
+matplotlib.rcParams["font.size"] = 24
+for tick in ["xtick", "ytick"]:
+    for size in ["major", "minor"]:
+        for attr in ["pad", "size"]:
+            matplotlib.rcParams["%s.%s.%s" % (tick, size, attr)] *= 2
 from matplotlib.pyplot import *
 import matplotlib.pyplot as plt
 
@@ -42,7 +47,7 @@ def show_lex_posterior(path, weights_particles, idx1, idx2,
             plt.ylabel("P(word %s means obj %s | data)" % idx2)
         else:
             plt.ylabel(ylabel)
-    savefig(path)
+    savefig(path, bbox_inches="tight")
     close(f)
 
 def simulate_dialogues(count, turns_per_dialogue, domain, lexicon_prior=None,
@@ -117,7 +122,7 @@ def show_dialogue(path, dialogue):
             sub_ax.set_yticks([])
     ax.set_zorder(10)
     ax.patch.set_alpha(0)
-    fig.savefig(path)
+    fig.savefig(path, bbox_inches="tight")
 
 def show_dialogues(base_path, dialogues, **kwargs):
     for i, dialogue in enumerate(dialogues):

### 2013-05-30 figures/scalar-implic.py

@@ -1,15 +1,17 @@
 from nips import *
 
+f = open("scalar-implic.log", "w")
+
 # Demonstrating scalar implicature
 
 some_all_domain = dom(adjectives=2, objects=2)
 some_all_prior = [[10, 10], [1, 10]]
 dialogue = conpact2.Dialogue(some_all_domain, LISTENER_DEPTH, 0, PARTICLES, lexicon_prior=some_all_prior)
 
-print "How speaker refers to \"some\" and \"all\" objects:"
-print np.exp(dialogue.uncertain_s_dist())
-print "How listener interprets \"some\" and \"all\" words:"
-print np.exp(dialogue.uncertain_l_dist())
+f.write("How speaker refers to \"some\" and \"all\" objects:\n")
+f.write(repr(np.exp(dialogue.uncertain_s_dist())))
+f.write("\n\nHow listener interprets \"some\" and \"all\" words:\n")
+f.write(np.exp(dialogue.uncertain_l_dist()))
 
 # Learning scalar implicature
 

### 2013-05-30 praglearn.tex
figure updates

@@
 
+% Simulation parameters:
+%   prior on meaning of "some": dirichlet(10, 10)
+%   prior on meaning of "all": dirichlet(1, 10)
+%
+% How speaker refers to "some" and "all" objects:
+% array([[  9.99951162e-01,   2.87541198e-03],
+%        [  4.88378335e-05,   9.97124588e-01]])
+%
+% How listener interprets "some" and "all" words:
+% array([[ 0.86214895,  0.13785105],
+%        [ 0.03151434,  0.96848566]])
+
 \subsubsection{Horn implicature}
+
+% Simulation parameters:
+%   word costs: 0.5, 1.0
+%   object prior: 0.8, 0.2
+%
+% How speaker refers to "common" and "rare" objects:
+% array([[ 0.88491473,  0.53858973],
+%        [ 0.11508527,  0.46141027]])
+%
+% How listener interprets "cheap" and "expensive" words:
+% array([[ 0.77458316,  0.22541684],
+%        [ 0.64702551,  0.35297449]])
 
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=0.20\textwidth]{figures/cross-sit.pdf}
+  \caption{Cross-situational simulation}
+  \label{fig:cross-sit}
+\end{figure}
+
 \subsubsection{Disambiguation using known words}
+
+% Novel word means: Familiar object: 25.5%. Novel object: 74.5%.
+% To anti-sparse listener, novel word means: Familiar object: 21.4%. Novel object: 78.6%.
+% To sparse listener, novel word means: Familiar object: 30.5%. Novel object: 69.5%.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.20\textwidth]{figures/ME-1dax.pdf}
+  \includegraphics[width=0.20\textwidth]{figures/ME-flat-10dog-10dax.pdf}
+  \includegraphics[width=0.20\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
+  \includegraphics[width=0.20\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
+  \caption{Mutual exclusivity simulation}
+  \label{fig:mutual-exclusivity}
+\end{figure}
 
@@
 
-\subsubsection{Lexicaliation and preservation of scalar implicature}
+% To show that they're lexicalized, at end of each of 4 dialogues, we
+% look at how speaker and listener talk about/interpret the objects/words:
+%
+% Dialogue 0, after disabling implicature:
+%   How speaker refers to "common" and "rare" objects:
+% array([[  9.99994601e-01,   1.67862018e-02],
+%        [  5.39888463e-06,   9.83213798e-01]])
+%   Listener's interpretation of "cheap" and "expensive" words:
+% array([[ 0.84765063,  0.15234937],
+%        [ 0.02451694,  0.97548306]])
+
+% Dialogue 1, after disabling implicature:
+%   How speaker refers to "common" and "rare" objects:
+% array([[  9.99997339e-01,   1.04639034e-02],
+%        [  2.66058112e-06,   9.89536097e-01]])
+%   Listener's interpretation of "cheap" and "expensive" words:
+% array([[ 0.86951534,  0.13048466],
+%        [ 0.01986478,  0.98013522]])
+
+% Dialogue 2, after disabling implicature:
+%   How speaker refers to "common" and "rare" objects:
+% array([[  9.99984637e-01,   2.19135388e-02],
+%        [  1.53629560e-05,   9.78086461e-01]])
+%   Listener's interpretation of "cheap" and "expensive" words:
+% array([[ 0.83486332,  0.16513668],
+%        [ 0.03421814,  0.96578186]])
+
+% Dialogue 3, after disabling implicature:
+%   How speaker refers to "common" and "rare" objects:
+% array([[  9.99996148e-01,   1.92821280e-02],
+%        [  3.85172462e-06,   9.80717872e-01]])
+%   Listener's interpretation of "cheap" and "expensive" words:
+% array([[ 0.83985034,  0.16014966],
+%        [ 0.02170542,  0.97829458]])
+
+\subsubsection{Lexicalization and preservation of scalar implicature}
 

### 2013-05-30 figures/conpact2.py

@@ -269,6 +269,7 @@ class Domain(object):
         return U
 
     def dist_n(self, n, lexicon, log_object_prior=None):
+        assert n >= 0
         log_object_prior = self.log_object_prior(log_object_prior)
         if n % 2 == 0:
             # Listener

### 2013-05-30 figures/horn.py

@@ -14,6 +14,7 @@ f.write("How speaker refers to \"common\" and \"rare\" objects:\n")
 f.write(repr(np.exp(dialogue.uncertain_s_dist())))
 f.write("\n\nHow listener interprets \"cheap\" and \"expensive\" words:\n")
 f.write(repr(np.exp(dialogue.uncertain_l_dist())))
+f.write("\n\n")
 
 # Emergence + Horn implicature
 
@@ -27,5 +28,13 @@ f.write(repr(np.exp(dialogue.uncertain_l_dist())))
 # XX mention: Horn implicature can't survive as implicatures. And connect to
 # scalar implicature, which *can* survive.
 
-bgl_dialogues = simulate_dialogues(4, 20, bgl_domain)
+bgl_dialogues = simulate_dialogues(4, 10, bgl_domain)
 show_dialogues("horn-emergence-%i.pdf", bgl_dialogues)
+
+for i, dialogue in enumerate(bgl_dialogues):
+    f.write("Dialogue %s, after disabling implicature:\n" % (i,))
+    f.write("  How speaker refers to \"common\" and \"rare\" objects:\n")
+    f.write(repr(np.exp(dialogue.uncertain_s_dist(log_object_prior=np.log([0.5, 0.5])))))
+    f.write("\n  Listener's interpretation of \"cheap\" and \"expensive\" words:\n")
+    f.write(repr(np.exp(dialogue.uncertain_l_dist(log_object_prior=np.log([0.5, 0.5])))))
+    f.write("\n\n")

### 2013-05-30 figures/nips.py

@@ -51,7 +51,17 @@ def show_lex_posterior(path, weights_particles, idx1, idx2,
     close(f)
 
 def simulate_dialogues(count, turns_per_dialogue, domain, lexicon_prior=None,
-                       listener_depth=LISTENER_DEPTH, particle_seed_base=0, particle_count=PARTICLES):
+                       listener_depth=LISTENER_DEPTH,
+                       listener_speaker_depth=None,
+                       particle_seed_base=0, particle_count=PARTICLES):
+    # The listener always models their evidence as coming from the speaker
+    # just below them, so this should normally be None. The only reason this
+    # argument exists is for the hack in non-emergence.py where we want to
+    # model the speaker just sampling from the lexicon and the listener
+    # likewise, and to do that we set it up so that S1 and S-1 are
+    # equivalent, and then L0 learns from S1.
+    if listener_speaker_depth is None:
+        listener_speaker_depth = listener_depth - 1
     dialogues = []
     for i in xrange(count):
         r = np.random.RandomState(i)
@@ -64,12 +74,10 @@ def simulate_dialogues(count, turns_per_dialogue, domain, lexicon_prior=None,
             obj, utt, interp, backchannel = dialogue.sample_obj_utt_interp_backchannel(r)
             # Speaker observes listener's interpretation; listener observe speaker's intended meaning
             s_datum = conpact2.SDataInterp(listener_depth, utt, interp)
-            l_datum = conpact2.LDataUttWithMeaning(listener_depth - 1, utt, obj)
+            l_datum = conpact2.LDataUttWithMeaning(listener_speaker_depth,
+                                                   utt, obj)
             dialogue.new_data([s_datum], [l_datum])
     return dialogues
-
-# XX graphs: write "speaker model of listener", "listener model of listener"
-# directly on the graphs somewhere? Where?
 
 def show_dialogue(path, dialogue):
     fig = figure(figsize=(CONVERGENCE_FIG_WIDTH, CONVERGENCE_FIG_WIDTH / 4.0))

### 2013-05-30 figures/scalar-implic.py

@@ -11,7 +11,7 @@ dialogue = conpact2.Dialogue(some_all_domain, LISTENER_DEPTH, 0, PARTICLES, lexi
 f.write("How speaker refers to \"some\" and \"all\" objects:\n")
 f.write(repr(np.exp(dialogue.uncertain_s_dist())))
 f.write("\n\nHow listener interprets \"some\" and \"all\" words:\n")
-f.write(np.exp(dialogue.uncertain_l_dist()))
+f.write(repr(np.exp(dialogue.uncertain_l_dist())))
 
 # Learning scalar implicature
 

### 2013-05-30 figures/conpact2.py
Fix a very serious bug in conpact2.py's dialogue simulation code

@@ -1,6 +1,3 @@
-# SNAPSHOT of the conpact2.py model code used for NIPS 2013 submission
-# See pragmods repository for the "official" version
-
 # Second attempt to model conceptual pacts via bayesian pragmatics.
 
 # We imagine a scenario in which a speaker describes an object using a
@@ -40,7 +37,7 @@
 # and they do a softmax from the resulting marginalized distribution.
 #
 # Both speaker and listener use a dirichlet prior on each word's meaning, the
-# vector lex[obj, :]. However, they have different conditioning data. For
+# vector lex[word, :]. However, they have different conditioning data. For
 # listeners, conditioning data might include:
 #   -- that certain utterances were produced (this is informative because it
 #      suggests that those utterances are distinctive)
@@ -245,6 +242,46 @@ class Domain(object):
         logZ = np.logaddexp.reduce(logP_L, axis=1)
         logP_L -= logZ[:, np.newaxis]
         return logP_L
+
+    def S_minus_one(self, lexicon):
+        # 'lexicon' is an L0 lexicon. Figure out the corresponding P(word|obj)
+        # for a speaker that would lead to this effective lexicon if L_0 were
+        # doing rational inference over S_{-1}.
+        # This sub-lexicon is an unknown matrix, that satisfied two
+        # constraints:
+        # - each column sums to 1
+        # - each entry, divided by its row's sum, equals the corresponding
+        #   entry in the lexicon
+        #   - which means, that lex entry times the row's sum equals the
+        #     sublex entry, so this is a linear constraint.
+        # We unravel sublex into a vector and solve for it.
+        # There are 'objects' constraints of the first kind, and 'objects x
+        # adjectives' constraints of the second kind. (Yes, this means some of
+        # them are redundant. Whatever.)
+        A = np.empty((self.objects + (self.adjectives * self.objects),
+                      self.adjectives * self.objects))
+        b = np.empty(self.objects + (self.adjectives * self.objects))
+        constraint_i = 0
+        for j in xrange(self.objects):
+            # Each column sums to 1
+            constraint = np.zeros((self.adjectives, self.objects))
+            constraint[:, j] = 1
+            A[constraint_i, :] = constraint.ravel()
+            b[constraint_i] = 1
+            constraint_i += 1
+        for i in xrange(self.adjectives):
[…]

### 2013-05-30 figures/emergence.py

@@ -19,5 +19,5 @@ dialogues = simulate_dialogues(4, 10, d)
 show_dialogues("emergence2x2-%i.pdf", dialogues)
 
 d = dom(adjectives=3, objects=3)
-dialogues = simulate_dialogues(4, 10, d, particle_count=PARTICLES * 10)
+dialogues = simulate_dialogues(4, 30, d, particle_count=PARTICLES * 10)
 show_dialogues("emergence3x3-%i.pdf", dialogues)

### 2013-05-30 figures/horn.py

@@ -28,7 +28,7 @@ f.write("\n\n")
 # XX mention: Horn implicature can't survive as implicatures. And connect to
 # scalar implicature, which *can* survive.
 
-bgl_dialogues = simulate_dialogues(4, 10, bgl_domain)
+bgl_dialogues = simulate_dialogues(4, 30, bgl_domain)
 show_dialogues("horn-emergence-%i.pdf", bgl_dialogues)
 
 for i, dialogue in enumerate(bgl_dialogues):

### 2013-05-30 figures/nips.py

@@ -52,16 +52,8 @@ def show_lex_posterior(path, weights_particles, idx1, idx2,
 
 def simulate_dialogues(count, turns_per_dialogue, domain, lexicon_prior=None,
                        listener_depth=LISTENER_DEPTH,
-                       listener_speaker_depth=None,
                        particle_seed_base=0, particle_count=PARTICLES):
-    # The listener always models their evidence as coming from the speaker
-    # just below them, so this should normally be None. The only reason this
-    # argument exists is for the hack in non-emergence.py where we want to
-    # model the speaker just sampling from the lexicon and the listener
-    # likewise, and to do that we set it up so that S1 and S-1 are
-    # equivalent, and then L0 learns from S1.
-    if listener_speaker_depth is None:
-        listener_speaker_depth = listener_depth - 1
+    listener_speaker_depth = listener_depth - 1
     dialogues = []
     for i in xrange(count):
         r = np.random.RandomState(i)

### 2013-05-30 figures/parallel_dialogues.py
start at code for simulating bulk dialogues for statistical purposes

@@ -1,0 +1,51 @@
+import multiprocessing
+import conpact2
+import cPickle
+import numpy as np
+
+SOFTMAX_WEIGHT = 3
+LISTENER_DEPTH = 1
+PARTICLE_COUNT = 1000
+
+def simulate_dialogue(r, dialogue, turns):
+    for turn in xrange(turns):
+        obj, utt, interp, backchannel = dialogue.sample_obj_utt_interp_backchannel(r)
+        # Speaker observes listener's interpretation; listener observe
+        # speaker's intended meaning
+        s_datum = conpact2.SDataInterp(listener_depth, utt, interp)
+        l_datum = conpact2.LDataUttWithMeaning(listener_speaker_depth,
+                                               utt, obj)
+        dialogue.new_data([s_datum], [l_datum])
+
+def horn_dialogue(i_turns):
+    i, turns = i_turns
+    bgl_domain = conpact2.MemoizedDomain(adjectives=2, objects=2,
+                                         max_utterance_length=1,
+                                         softmax_weight=SOFTMAX_WEIGHT,
+                                         word_cost=[0.5, 1.0],
+                                         default_object_prior=[0.8, 0.2])
+    r = np.random.RandomState(i)
+    flat = np.log([0.5, 0.5])
+    def trace_horn_S_lexicalization(turn, dialogue):
+        return np.exp(dialogue.uncertain_s_dist(log_object_prior=flat))
+    def trace_horn_L_lexicalization(turn, dialogue):
+        return np.exp(dialogue.uncertain_l_dist(log_object_prior=flat))
+
+    tracers = {"P-understood": conpact2.trace_P_understood,
+               "horn-S-lexicalization": trace_horn_S_lexicalization,
+               "horn-L-lexicalization": trace_horn_L_lexicalization,
+               }
+    dialogue = conpact2.Dialogue(domain, listener_depth,
+                                 i, PARTICLE_COUNT,
+                                 tracers=tracers)
+    simulate_dialogue(r, dialogue, turns)
+
+    return i, turns, dialogue.traces
+
+def par_horn_dialogues(count, turns, outpath):
+    p = multiprocessing.Pool()
+    jobs = [(i, turns) for i in xrange(count)]
+    for (i, turns, traces) in p.imap_unordered(horn_dialogue, jobs):
+        cPickle.dump((i, turns, traces),
+                     open("%s/horn-%s-%s.pickle" % (outpath, turns, i), "w"),
+                     protocol=-1)

### 2013-05-30 figures/conpact2.py
don't require pandas at import time

@@ -125,7 +125,6 @@ from itertools import combinations_with_replacement
 from collections import OrderedDict
 import numpy as np
 from scipy.special import gammaln
-import pandas
 
 ################################################################
 # The model
@@ -799,6 +798,7 @@ def trace_length_dist(turn, dialogue, log_object_prior=None):
     linear_utt_dist = np.sum(np.exp(s_dist + log_object_prior[np.newaxis, :]),
                              axis=1)
     # Sum the probability mass over all same-length utterances
+    import pandas
     length_dist = (pandas.Series(linear_utt_dist)
                      .groupby(dialogue.domain.utt_lengths)
                        .aggregate(np.sum))

### 2013-05-30 figures/parallel-report.py
oops, accidentally started writing this on compute server, committing to get it back to my laptop

@@ -1,0 +1,10 @@
+from nips import *
+from parallel_dialogues import *
+
+mean_P_understood = {}
+for name, pattern in [("flat2", "par/flat-50-2-*.pickle"),
+                      ("flat3", "par/flat-50-3-*.pickle"),
+                      ("horn", "par/horn-50-*.pickle"),
+                      ]:
+    mean_P_understood[name] = np.mean([extract_P_understood(pattern, i)
+                                       for i in xrange(50)])

### 2013-05-30 figures/cross-sit.py
yet more figure updates...

@@ -41,7 +41,7 @@ for missing_obj in xrange(CS_OBJECTS):
                                              np.log(object_prior)))
 
 d = dom(adjectives=4, objects=3)
-learner = conpact2.FixedSupportImportanceSampler(d, 23, PARTICLES * 100,
+learner = conpact2.FixedSupportImportanceSampler(d, 23, PARTICLES * 1000,
                                                  data=data_points * 10)
 matshow(np.exp(learner.marginal_dist_n(LISTENER_DEPTH)), cmap="binary",
         vmin=0, vmax=1)

### 2013-05-30 figures/horn.py

@@ -18,9 +18,6 @@ f.write("\n\n")
 
 # Emergence + Horn implicature
 
-# XX TODO: calculate some statistics on P(convergence to good system) by
-# running a few hundred dialogues
-#
 # XX mention: this is a very general result, can speculate about it as a
 # partial cause of things like common words being shorter, rise of useful
 # social conventions in general
@@ -28,7 +25,7 @@ f.write("\n\n")
 # XX mention: Horn implicature can't survive as implicatures. And connect to
 # scalar implicature, which *can* survive.
 
-bgl_dialogues = simulate_dialogues(4, 30, bgl_domain)
+bgl_dialogues = simulate_dialogues(4, 10, bgl_domain)
 show_dialogues("horn-emergence-%i.pdf", bgl_dialogues)
 
 for i, dialogue in enumerate(bgl_dialogues):

### 2013-05-30 figures/nips.py

@@ -10,6 +10,8 @@ PARTICLES = 1000
 LISTENER_DEPTH = 2
 CONVERGENCE_FIG_WIDTH = 20
 
+TURN_XLABEL = "Dialogue turn"
+
 import matplotlib
 matplotlib.use("Cairo")
 matplotlib.rcParams["xtick.direction"] = "out"
@@ -81,7 +83,7 @@ def show_dialogue(path, dialogue):
     xlim(x[0], x[-1] + 1)
     ylim(0, 1)
     ax.set_yticks([0, 0.5, 1.0])
-    xlabel("Dialogue turn")
+    xlabel(TURN_XLABEL)
     ylabel("P(L understands S)")
     # These are in data coordinates
     IDEAL_BOX_WIDTH = 0.8

### 2013-05-30 figures/parallel-report.py

@@ -1,10 +1,41 @@
 from nips import *
 from parallel_dialogues import *
 
+FLAT2_PATTERN = "par/flat-50-2-*.pickle"
+FLAT3_PATTERN = "par/flat-50-3-*.pickle"
+HORN_PATTERN = "par/horn-50-*.pickle"
+
+# We have 50, but nothing really interesting happens after 40
+TURNS = 40
+
 mean_P_understood = {}
-for name, pattern in [("flat2", "par/flat-50-2-*.pickle"),
-                      ("flat3", "par/flat-50-3-*.pickle"),
-                      ("horn", "par/horn-50-*.pickle"),
+for name, desc, pattern, lty in [("flat2", "2x2 uniform prior", FLAT2_PATTERN,
+                                  "k-."),
+                                 ("flat3", "3x3 uniform prior", FLAT3_PATTERN,
+                                  "r--"),
+                                 ("horn", "Horn implicature", HORN_PATTERN,
+                                  "g-"),
                       ]:
-    mean_P_understood[name] = np.mean([extract_P_understood(pattern, i)
-                                       for i in xrange(50)])
+    mean_P_understood[name] = [np.mean(extract_P_understood(pattern, i))
+                               for i in xrange(TURNS)]
+    plot(mean_P_understood[name], lty, label=desc, linewidth=2)
+ylim(0, 1)
+xlim(-0.5, TURNS + 0.5)
+legend(loc="lower right")
+xlabel(TURN_XLABEL)
+ylabel("Mean P(L understands S)")
+savefig("emergence-average.pdf", bbox_inches="tight")
+
+mean_horn_success = [np.mean(extract_horn_success(HORN_PATTERN, i))
+                     for i in xrange(TURNS)]
+mean_horn_antisuccess = [np.mean(extract_horn_antisuccess(HORN_PATTERN, i))
+                         for i in xrange(TURNS)]
+figure()
+plot(mean_horn_success, "g-", linewidth=2, label="Good lexicon")
+plot(mean_horn_antisuccess, "b--", linewidth=2, label="Bad lexicon")
+ylim(0, 1)
+xlim(-0.5, TURNS + 0.5)
+xlabel(TURN_XLABEL)
+ylabel("Horn lexicalization rate")
+legend(loc="upper left")
+savefig("emergence-horn-average.pdf", bbox_inches="tight")

### 2013-05-30 praglearn.tex
Fix a very serious bug in conpact2.py's dialogue simulation code
praglearn stuff
yet more figure updates...
text

@@
 
-\section{Experiments}
+\section{Experiments 1: Empirical coverage}
+
+In this section, we describe the coverage of our pragmatic learning model on phenomena previously investigated with the pragmatics and word learning components of the model.
+To summarize these simulations: the pragmatic learning model reproduces previous findings and provides some new insights into the ``disambiguation'' phenomenon described above.
 
 \begin{table}[t]
-\caption{Table of empirical results and references}
 \label{tab:results}
 \begin{center}
-\begin{tabular}{llllll}
+\begin{tabular}{lccccc}
 \hline \\
-Phenomenon & Ref. & Word learning \cite{frank2009} & Pragmatics \cite{frank2012,bergen2012}  & Model 1 & Model 2 \\
+Phenomenon & Ref. & WL & PI & Lifted PI  & PIL \\
 \hline
-Scalar implicature & \cite{stiller2011} & & & & \\
-Horn implicature & & & & & \\
-Cross-situational learning & \cite{yu2007} & & & & \\
-Disambiguation & \cite{markman1988} & & & & \\
-Communicative equilibria & \cite{galantucci2007} & & & & \\
+Scalar implicature & \cite{grice1975} &  & x &x & x\\
+Horn implicature & \cite{horn1984} & &  & x& x\\
+Cross-situational learning & \cite{smith2008} & x& & & x \\
+Disambiguation & \cite{markman1988} &x  & & & x \\
+Communicative equilibria & \cite{galantucci2005} & & & & x \\
+\hline
 \end{tabular}
 \end{center}
+\caption{Empirical results and references.
+WL refers to the word learning model of \cite{frank2009}; PI refers to the pragmatic inference model of \cite{frank2012}; Lifted PI refers to the pragmatic inference model of \cite{bergen2012}, and PIL refers to the pragmatic inference and learning model described here.
+}
+
 \end{table}
 
-\subsection{Pragmatic reasoning}
+\subsection{Scalar implicature}
 
-\subsubsection{Scalar implicature}
+Many sets of words in natural language form scales in which each term refers to successively stronger restrictions on the situations compatible with the term.
+``Some'' and ``all'' form a scale of this type.
+While ``I ate some of the cookies'' is compatible with the followup ``in fact, I ate {\it all} of the cookies,'' the reverse is not true.
+``Might'' and ``must'' are another example, as are ``OK,'' ``good,'' and ``excellent.'' All of these scales allow for {\it scalar implicatures} \cite{grice1975}: if a speaker uses a weaker term, a listener is licensed to infer that the stronger term was not true in this situation.
+So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs.
+
+The recursive pragmatic reasoning portions of our model capture set of findings on scalar implicature effectively. Consider a representation of scalar implicature as a matrix 
+$\left[
+    \begin{array}{cc}
+      1 & 1 \\
+      0 & 1
+    \end{array} 
+  \right]$,
+where rows show the terms ``some'' and ``all'' and columns indicate worlds where {\sc some} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case.
+Assuming uniform costs for the two words and uniform priors, each recursion of listener modeling speaker reduces the probability that ``some'' will be used to refer to {\sc all}.
+Using the fixed parameters employed in our other simulations results in speaker use probabilities of
+$\left[
+    \begin{array}{cc}
+      1.00 & 0.00 \\
+      0.00 & 1.00
+    \end{array} 
+  \right]$
+and listener interpretation probabilities of 
+$\left[
+    \begin{array}{cc}
+      .86 & .14 \\
+      .03 & .97
+    \end{array} 
+  \right]$.
 
@@
 
-\subsubsection{Horn implicature}
+\subsection{Horn implicature}
+
+Horn implicatures are inferences where the critical factor is not the extension of the words in context (as in scalar implicatures), but instead the relative cost of two expressions.
+In Horn's example, ``Lee got the car to stop'' implies that Lee used an unusual method (e.g.
+not the brakes) because, had he used the brakes, the speaker would have chosen the simpler and shorter (less costly) expression, ``Lee stopped the car.'' In other words, the costly expression is assumed to be associated with the unusual outcome.
+
+Such an implicature again arises naturally from our model.
+We create a situation in which expressions have differential costs (e.g.
+$[.5 1.0]$) and referents have differential probabilities (e.g.
+$[.8 .2]$).
+Speakers in such a situation choose simple words to refer to common referents, e.g.
+$\left[
+    \begin{array}{cc}
+      .88 & .54 \\
+      .12 & .46
+    \end{array} 
+  \right]$.\footnote{Rows indicate words and columns indicate referents.
+For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.} Similarly, listeners interpret ``cheap'' and ``expensive'' words to refer to common and rare objects respectively:
+$\left[
+    \begin{array}{cc}
+      .77 & .23 \\
+      .65 & .35
+    \end{array} 
+  \right]$.
 
@@
 
-\subsection{Word learning}
+\subsection{Statistical cross-situational word learning}
 
-\subsubsection{Statistical cross-situational word learning}
+A recent line of work in language acquisition has 
+% \begin{figure}[t]
+%   \centering
+%   \includegraphics[width=0.20\textwidth]{figures/cross-sit.pdf}
+%   \caption{Cross-situational simulation}
+%   \label{fig:cross-sit}
+% \end{figure}
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.20\textwidth]{figures/cross-sit.pdf}
-  \caption{Cross-situational simulation}
-  \label{fig:cross-sit}
-\end{figure}
-
-\subsubsection{Disambiguation using known words}
+\subsection{Disambiguation using known words}
 
@@
 
-\begin{figure}
+\begin{figure}[t]
   \centering
@@
 
-\subsection{Lexical learning in pragmatic contexts}
+\section{Experiments 2: Lexical learning in pragmatic contexts}
 
-\subsubsection{Emergence of communicative equilibria}
+\subsection{Emergence of communicative equilibria}
 
-\begin{figure}
+\begin{figure}[t]
 \centering
@@
 
-\subsubsection{Lexicalization of Horn implicatures}
+\subsection{Lexicalization of Horn implicatures}
 
-\begin{figure}
+\begin{figure}[t]
 \centering
@@
 
-\subsubsection{Lexicalization and preservation of scalar implicature}
+\subsection{Lexicalization and preservation of scalar implicature}
 

### 2013-05-30 praglearn.tex
text

--- 2013-05-30 praglearn.tex
@@
 \newcommand{\new}{\marginpar{NEW}}
+
+\newcommand{\word}{\text{word}}
+\newcommand{\obj}{\text{object}}
 
@@
 
+Consider a simple communicative game: two participants each,
+separately, view identical arrays of objects. On the
+\textit{Speaker's} screen, one object is highlighted; their goal is to
+get the \textit{Listener} to click on this item. To do this, they have
+available a fixed, finite set of words; they must pick one. The
+Listener then receives this word, and attempts to guess which object
+the Speaker meant by it.
+
+This suggest a simple probabilistic model. Given any particular
+decision procedure used by the Speaker $S$, the optimal way for the
+Listener $L$ to judge the probability of each object is to invert this
+decision procedure using Bayes' rule, taking into account the prior
+likelihood of each object:
+\begin{align*}
+P_L(\obj | \word) &\propto P_S(\word | \obj) P_{\text{prior}}(\obj |
+\text{context})
+\end{align*}
+$S$, meanwhile, being herself a rational Bayesian agent, chooses her
+word so as to soft-maximize a utility function $U(\word | \obj)$. She
+gains utility if $L$ is likely to understand her; she loses utility if
+the word is particularly costly to say (e.g., it may be long).
+\begin{align*}
+P_S(\word | \obj) &\propto \exp \lambda U(\word | \obj) \\
+U(\word | \obj) &= \log P_L(\obj | \word) - \text{cost}(\word)
+\end{align*}
+The degree to which she optimizes this utility is controlled by the
+softmax parameter $\lambda$. If $\lambda = 1$ and all words cost the
+same, then her probability of producing a word will be proportional to
+the likelihood that it will cause $L$ to reach the correct
+interpretation; as $\lambda$ increases, she becomes more likely to
+choose the word which maximizes this likelihood. (We somewhat
+arbitrarily set $\lambda = 3$ throughout.)
+
+This model as stated is optimal by various criteria, but suffers from
+a fatal flaw: $S$'s decision procedure depends on $L$'s interpretive
+procedure, and vice-versa. This means that it can only be implemented
+as stated for a very spei
+
+And indeed, in similar games, humans seem to reason only a few levels
+deep (XX cite p-beauty contests), and
+
 Basic idea of iterated pragmatic reasoning models.
@@
 
-\newcommand{\word}{\text{word}}
-\newcommand{\obj}{\text{object}}
 \begin{align*}
@@
 \end{align*}
-
-For the rest of this paper we somewhat arbitrarily fix the softmax parameter to $\lambda = 3$.
 

### 2013-05-30 figures/horn.py
add the simulation output files that just have numbers in them; and include deep-recursion horn results in horn.log

@@ -16,6 +16,17 @@ f.write("\n\nHow listener interprets \"cheap\" and \"expensive\" words:\n")
 f.write(repr(np.exp(dialogue.uncertain_l_dist())))
 f.write("\n\n")
 
+DEEP = 20
+dialogue_deep = conpact2.Dialogue(bgl_domain, DEEP, 0, PARTICLES * 10)
+
+f.write("How %s-deep speaker refers to \"common\" and \"rare\" objects:\n"
+        % (DEEP + 1,))
+f.write(repr(np.exp(dialogue_deep.uncertain_s_dist())))
+f.write("\n\nHow %s-deep listener interprets \"cheap\" and \"expensive\" words:\n"
+        % (DEEP,))
+f.write(repr(np.exp(dialogue_deep.uncertain_l_dist())))
+f.write("\n\n")
+
 # Emergence + Horn implicature
 
 # XX mention: this is a very general result, can speculate about it as a

### 2013-05-31 praglearn.tex
text

--- 2013-05-30 praglearn.tex
@@
 \newcommand{\new}{\marginpar{NEW}}
+
+\newcommand{\word}{\text{word}}
+\newcommand{\obj}{\text{object}}
+\newcommand{\lex}{\text{lexicon}}
+\newcommand{\prior}{P_{\text{prior}}(\obj | \text{context})}
 
@@
 
-\section{Models}
+\section{Model}
 
-Basic idea of iterated pragmatic reasoning models.
-$S_i$ defines a causal model for utterances described by a simple Bayes net.
-$L_{i+1}$ does standard rational inference over this network.
+Consider a simple communicative game: two participants each,
+separately, view identical arrays of objects. On the
+\textit{Speaker's} screen, one object is highlighted; their goal is to
+get the \textit{Listener} to click on this item. To do this, they have
+available a fixed, finite set of words; they must pick one. The
+Listener then receives this word, and attempts to guess which object
+the Speaker meant by it.
+
+When modelling human behaviour, it's useful to begin with optimal models; this has the advantage that when we are then forced to deviate from the purely rational approach, it's generally in a theoretically interpretable way.
+However, multi-agent interactions are notoriously difficult to model in a normative fashion without falling prey to paradox.
+For example, consider a simple model of the agents in the above game.
+First we define a \textit{literal listener} agent denoted $L_0$.
+This agent has a \textit{lexicon} which probabilistically maps words to meanings; specifically, it assigns each word $w$ a vector describing the extent to which this word provides evidence for each object; each entry in the vector is in $(0, 1)$, and together they sum to 1:
+\begin{align*}
+P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
+\end{align*}
+Confronted with such a listener, a rational (but noisy) speaker should attempt to choose a word which soft-maximizes the probability that the listener will assign to the target object -- possibly modulated by the effort or cost associated with producing this word:
+\begin{align*}
+P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
+\end{align*}
+But if this is how the speaker will behave, then it's not rational for a listener to implement the $L_0$ strategy.
+Instead, to best determine what object was intended, they should use Bayes rule to invert the speaker's decision procedure:
+\begin{align*}
+P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
+\end{align*}
+Now the paradox becomes apparent.
+Given such a listener, it's no longer rational for the speaker to implement strategy $S_1$; instead, they should implement strategy $S_3$ which replaces the $P_{L_0}$ term with $P_{L_2}$.
+And then the listener ought to implement $L_4$, and so on.
+
+One computationally expensive option is to continue iterating such strategies until one reaches a fixed point equilibrium.
+But, while this guarantees each agent will behave normatively given the other agent's strategy, there is no guarantee that such a strategy will be optimal for the system as a whole; equilibria strategies make up only a small subspace of all possible strategies.
+Nor is there much evidence the humans find such equilibria; their behaviour in language games (and in other games XX p-beauty) can be well-modeled as implementing $S_n$ or $L_n$ for some small $n$ \cite{frank2012}.
+Here we adopt $n = 3$ for speakers and $n = 2$ for listeners.
+
+Given such a model, it's conceptually straightforward for a listener to learn to understand a speaker -- the listener simply puts a prior on the speaker's lexicon, and updates it based on the speaker's utterances and whatever information the listener can glean about their intended meaning.
+
+\begin{align*}
+  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{L's data}) \times \prior \\
+    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{L's data}) \times \prior \\
+    &= P_{L_n}(\obj | \word, \text{L's data}). \\
+  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{S's data}) - \text{cost}(\word)\big)\Big).
+\end{align*}
+
+-- if they have the same data, $S$ is rational and $L$ is as close to rational as is possible given a finite recursion depth, i.e., if $S$ recursed one fewer times then $L$ would be rational.
+In particular, this describes all of the one-shot language games that have been considered in the literature to date, where neither agent has the time to acquire any data.
+-- if one actually does know the lexicon and the other is relatively ignorant, as in the child language learning case, then both are rational (with the same caveat).
+-- but in general, in multi-turn interactions, their beliefs will display some complicated
+
+% This suggest a simple probabilistic model. Given any particular
+% decision procedure used by the Speaker $S$, the optimal way for the
+% Listener $L$ to judge the probability of each object is to invert this
+% decision procedure using Bayes' rule, taking into account the prior
+% likelihood of each object:
+% \begin{align*}
+% P_L(\obj | \word) &\propto P_S(\word | \obj) P_{\text{prior}}(\obj |
+% \text{context})
+% \end{align*}
+% $S$, meanwhile, being herself a rational Bayesian agent, chooses her
+% word so as to soft-maximize a utility function $U(\word | \obj)$. She
+% gains utility if $L$ is likely to understand her; she loses utility if
+% the word is particularly costly to say (e.g., it may be long).
+% \begin{align*}
+% P_S(\word | \obj) &\propto \exp \lambda U(\word | \obj) \\
+% U(\word | \obj) &= \log P_L(\obj | \word) - \text{cost}(\word)
+% \end{align*}
+% The degree to which she optimizes this utility is controlled by the
+% softmax parameter $\lambda$. If $\lambda = 1$ and all words cost the
+% same, then her probability of producing a word will be proportional to
+% the likelihood that it will cause $L$ to reach the correct
+% interpretation; as $\lambda$ increases, she becomes more likely to
+% choose the word which maximizes this likelihood. (We somewhat
+% arbitrarily set $\lambda = 3$ throughout.)
+
+% This model as stated satisfies various normative criteria for each
+% agent individually, but suffers from a serious flaw that is intrinsic
+% to such multi-agent situations: $S$'s decision procedure depends on
+% $L$'s interpretive procedure, and vice-versa. If $S$ and $L$ want to
+% both be optimal simultaneously, then they must restrict themselves to
+% a low-dimensional subset of all available strategies. They must each choose
+% a strategy with the property that optimal counterpart of its optimal
+% counterpart is the original strategy. Finding such an equilibrium is
+% computationally non-trivial, and, even if found, there is no guarantee
+% that it will be optimal in any global sense. \cite{frank2012}
+% therefore propose an approximation to this model, in which $L$ and $S$
+% each perform a finite number of recursions, with $S_i$ reasoning about
+% $L_{i-1}$, who reasons about $S_{i-2}$ and so forth, until reaching
+% $L_0$, who interprets words according to an arbitrary rule (their
+% \textit{literal meaning}). In our case, we instantiate this as a
+% matrix (the \textit{lexicon}) with one column per object, and in which
+% each row represents the meaning of a particular word as a set of
+% values in $(0, 1)$ such that $\sum_\obj \lex(\word, \obj) = 1$.\footnote{In this model, words refer directly to objects; more realistically, words should refer to some sort of abstract semantic features, which in turn refer to objects.
+However, this simplification is without loss of generalization; we can interpret this model as marginalizing over such a representation, with our literal $P(\obj | \word) = \sum_\text{features} P(\obj | \text{features}) P(\text{features} | \word)$.}  Formally, we have:
+% \begin{align*}
+% P_{L_0}(\obj | \word) &= \lex(\word, \obj) P_{\text{prior}}(\obj | \text{context}). \\
+% P_{S_i}(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_{i - 1}}(\obj | \word) - \text{cost}(\word)\big)\Big) \\
+% P_{L_i}(\obj | \word) &\propto P_{S_{i-1}}(\word | \obj) P_{\text{prior}}(\obj | \text{context})
+% \end{align*}
+% This approach neatly avoids the infinite regress, and does well at predicting human behaviour in language games \cite{frank2012} and other multi-agent games (XX p-beauty).
+The minimal degree of recursion required to produce the effects we consider here is $S_3$ and $L_2$, and this what we use in the following.
+
+% When we consider adding learning to this model, however, we encounter a new and related problem.
+It's straightforward in principle to put a prior on $\lex(\word, \obj)$, and allow an agent to learn the
+
+At each stage, $S_i$ defines a causal model for utterances described
+by a simple Bayes net, and $L_{i+1}$ does Bayesian inference over this net.
 But the likelihood function in $S_i$'s model is very complicated, and includes equations derived from considering how a hypothetical $L_{i-1}$ would interpret utterances if they were doing rational inference about another hypothetical causal agent $S_{i-2}$.
-
-To simplify matters we envision a simple world that has $k$ words and $l$ objects. The \textit{lexicon} defines the literal meaning of each word, ...
 
@@
 
-\newcommand{\word}{\text{word}}
-\newcommand{\obj}{\text{object}}
-\begin{align*}
-P_{L_0}(\obj | \word) &= \text{lexicon}(\word, \obj) P_{\text{prior}}(\obj | \text{context}). \\
-P_{S_i}(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_{i - 1}}(\obj | \word) - \text{cost}(\word)\big)\Big) \\
-P_{L_i}(\obj | \word) &\propto P_{S_{i-1}}(\word | \obj) P_{\text{prior}}(\obj | \text{context})
-\end{align*}
-
-For the rest of this paper we somewhat arbitrarily fix the softmax parameter to $\lambda = 3$.
+we use $S_3$, $L_2$ and $\lambda = 3$.
 
@@
 To see how
-Make $L_0$ uncertain about values in $\text{lexicon}(\word, \obj)$, impose a prior, and then
+Make $L_0$ uncertain about values in $\lex(\word, \obj)$, impose a prior, and then
 
@@
 \subsection{Horn implicature}
+
+XX note the relation to BGL model; they need uncertainty in the middle of the pragmatics specifically, which is where we end up putting it for other reasons.
+And in Bayesian framework uncertainty and learning must always coincide.
+Also the intuition that this is a general property of uncertainty for soft-max agents -- if you don't know where you are, good worlds have more peaked utility distributions than bad worlds, and thus more consistent behaviour by other agents, so on average when you're guessing what other agents will do, good worlds contribute more to your average than bad worlds.
 

### 2013-05-31 praglearn.tex
modifications

--- 2013-05-30 praglearn.tex
@@
 \usepackage{graphicx}
+\usepackage[raggedright]{sidecap}
 
@@
 
-\section{Experiments 1: Empirical coverage}
+\section{Experiments: Pragmatic and word learning phenomena}
 
@@
 \hline
-Scalar implicature & \cite{grice1975} &  & x &x & x\\
+Specificity implicature & \cite{grice1975} &  & x &x & x\\
 Horn implicature & \cite{horn1984} & &  & x& x\\
 Cross-situational learning & \cite{smith2008} & x& & & x \\
-Disambiguation & \cite{markman1988} &x  & & & x \\
-Communicative equilibria & \cite{galantucci2005} & & & & x \\
+Disambiguation effect & \cite{markman1988} &x  & & & x \\
+Disambiguation without mapping & \cite{horst2008} &  & & & x \\
+Emergence of communicative equilibria & \cite{galantucci2005} & & & & x \\
+Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
+Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 \hline
@@
 
-\subsection{Scalar implicature}
+\subsection{Specificity implicature}
 
@@
 So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs.
+We use the term {\it specificity implicature} here (following \cite{bergen2012}) to denote that we are agnostic about whether such implicatures are conventionalized or whether they use a contextually-defined scale.
 
-The recursive pragmatic reasoning portions of our model capture set of findings on scalar implicature effectively. Consider a representation of scalar implicature as a matrix 
+The recursive pragmatic reasoning portions of our model capture findings on specificity implicature. Consider a representation of scalar implicature as a matrix 
 $\left[
@@
 $[.8 .2]$).
-Speakers in such a situation choose simple words to refer to common referents, e.g.
+At the settings used throughout the manuscript, speakers in such a situation choose simple words to refer to common referents at different rates than for uncommon references.
+This situation breaks the symmetry though does not result in as dramatic a diagonal as in the specificity implicature, e.g.
+for the speaker, \footnote{Rows indicate words and columns indicate referents.
+For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.}
 $\left[
@@
     \end{array} 
-  \right]$.\footnote{Rows indicate words and columns indicate referents.
-For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.} Similarly, listeners interpret ``cheap'' and ``expensive'' words to refer to common and rare objects respectively:
+  \right]$ 
+and for the listener,
 $\left[
@@
     \end{array} 
-  \right]$.
+  \right]$. See Section \ref{sec:hornlearn} for the consequences of repeating this implicature until it is lexicalized.
 
@@
 
-A recent line of work in language acquisition has 
-% \begin{figure}[t]
-%   \centering
-%   \includegraphics[width=0.20\textwidth]{figures/cross-sit.pdf}
-%   \caption{Cross-situational simulation}
-%   \label{fig:cross-sit}
-% \end{figure}
+Even very young children can learn words from individually ambiguous situations, e.g.
+where multiple words are spoken and multiple objects are present \cite{smith2008}.
+While no individual situation gives away the mapping, the pattern of co-occurrences between words and objects allows for learning.
+Our current model reproduces this pattern of results, converging to correct lexical mappings from ambiguous evidence.
+A representative simulation is shown in Fig.
+\ref{fig:cross-sit}.
+
+\begin{SCfigure}[2.5]
+  \includegraphics[width=0.2\textwidth]{figures/cross-sit-mod.pdf}
+  \caption{Results from a simulation of cross-situational word learning.
+The posterior distribution over the learner's lexicon (diagonal entries are correct for word/object pairs 1 -- 3; word D is a distractor).
+Training data were situations including two of the three objects and a single word, either the distractor or a word matching one of the two objects.}
+  \label{fig:cross-sit}
+\end{SCfigure}
 
 \subsection{Disambiguation using known words}
+
+\begin{SCfigure}[1]
+  \centering
+  % \includegraphics[width=0.24\textwidth]{figures/ME-1dax.pdf}
+  % \includegraphics[width=0.24\textwidth]{figures/ME-flat-10dog-10dax.pdf}
+  % \includegraphics[width=0.24\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
+  % \includegraphics[width=0.24\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
+\includegraphics[width=.6\textwidth]{figures/ME-composite-small.pdf}
+  \caption{Results for simulations of the ``disambiguation'' phenomenon.
+A novel word is used in a situation with a novel object and a familiar object.
+Each plot shows the joint posterior probability that the familiar word refers to the familiar object (horizontal) and that the novel word refers to the familiar object (vertical).
+Left plot shows the posterior after a single situation; right plot shows multiple situations.
+%The right middle plot shows the posterior after a single situation given either a prior that makes it likely that there are words referring to both the novel and familiar objects (perhaps because the two objects are physically similar; the right plot shows a situation where the prior makes it unlikely that a word refers to both (objects dissimilar).
+}
+  \label{fig:me}
+\end{SCfigure}
+
+Children, when presented with both a novel and a familiar object (e.g.
+an eggbeater and a ball), will treat a novel label (e.g.
+``dax'') as referring to the novel object, for example by supplying the eggbeater when asked to ``give me the dax'' \cite{markman1988}.
+This phenomenon is sometimes referred to as ``mutual exclusivity'' or even ``Fast mapping,'' and many distinct theoretical accounts have been proposed in the psychological literature.
+Simple probabilistic word learning models can produce a similar pattern of findings \cite{frank2009}, often due to the kinds of ``explaining away'' type phenomena that are observed in such models.
+But all such models assume that learners in fact retain the mapping between word and object that is demonstrated in the experimental situation; this observation is contradicted, however, by evidence that children often do not retain the mappings that are demonstrated by their inferences in the moment \cite{horst2008}.
+
+Our model provides an intriguing possible explanation of this finding: given a single disambiguation situation, the model gives a substantial probability (e.g.
+75\%) that the speaker is referring to the novel object.
+Nevertheless, integrating across lexicons reveals that this probability is not due to learning that the novel word refers to this object (as shown in Fig.
+\ref{fig:me}, left): instead, it comes in part from a specificity implicature because the listener knows that the familiar word \emph{does not} refer to the novel object--hence the novel word (even if it refers to the familiar object as well) is the best way to refer to the novel object.
+Nevertheless, on repeated exposure to the same novel word, novel object situation, the model does lexicalize this inference (Fig.
+\ref{fig:me}, right).
 
@@
 % To sparse listener, novel word means: Familiar object: 30.5%. Novel object: 69.5%.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=0.20\textwidth]{figures/ME-1dax.pdf}
-  \includegraphics[width=0.20\textwidth]{figures/ME-flat-10dog-10dax.pdf}
-  \includegraphics[width=0.20\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
-  \includegraphics[width=0.20\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
-  \caption{Mutual exclusivity simulation}
-  \label{fig:mutual-exclusivity}
-\end{figure}
 
@@
 
+\newpage
+
 \bibliographystyle{plain}

### 2013-05-31 praglearn.tex
more text

--- 2013-05-31 praglearn.tex
@@
 \usepackage{graphicx}
+\usepackage{booktabs}
+\usepackage{subcaption}
 
-\title{Learning words through recursive pragmatic reasoning about other agents}
+\title{Learning words by recursive pragmatic reasoning about other agents: The blessing of social anxiety}
 
@@
 
+% Tell LaTeX to break pages in the middle of displayed math whenever this would be the slightest bit useful
+\allowdisplaybreaks[4]
+
 \begin{document}
@@
 
-\section{Introduction}
+%\section{Introduction}
 
@@
 Our utterances are pervasively underspecified.
-Consider ``it's raining,'' ``I ate some of the cookies,'' or ``can you close the window?'' In each, a listener must go beyond the literal meaning of the words to fill in contextual details (``it's raining here and now''), infer that a stronger alternative is not true (``I ate some but not all of the cookies''), or more generally infer the speaker's communicative goal (``I want you to close the window right now because I'm cold.'').
+Consider ``it's raining,'' ``I ate some of the cookies,'' or ``can you close the window?'' In each, a listener must go beyond the literal meaning of the words to fill in contextual details (``it's raining here and now''), infer that a stronger alternative is not true (``I ate some but not all of the cookies''), or more generally infer the speaker's communicative goal (``I want you to close the window right now because I'm cold.'') -- a process known as \textit{pragmatic reasoning}.
 
@@
 First we define a \textit{literal listener} agent denoted $L_0$.
-This agent has a \textit{lexicon} which probabilistically maps words to meanings; specifically, it assigns each word $w$ a vector describing the extent to which this word provides evidence for each object; each entry in the vector is in $(0, 1)$, and together they sum to 1:
+This agent has a \textit{lexicon} which probabilistically maps words to meanings; specifically, it assigns each word $w$ a vector describing the extent to which this word provides evidence for each object\footnote{In this model, words refer directly to objects; more realistically, words should refer to some sort of abstract semantic features, which in turn refer to objects.
+However, our simplification is without loss of generalization, because we can interpret our model as marginalizing over such a representation, with our literal $P_\lex(\obj | \word) = \sum_\text{features} P(\obj | \text{features}) P_\lex(\text{features} | \word)$.}.
+Each entry in the vector is in $(0, 1)$, and together they sum to 1:
 \begin{align*}
@@
 Now the paradox becomes apparent.
-Given such a listener, it's no longer rational for the speaker to implement strategy $S_1$; instead, they should implement strategy $S_3$ which replaces the $P_{L_0}$ term with $P_{L_2}$.
+Given such a listener, it's no longer rational for the speaker to implement strategy $S_1$; instead, they should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead of $P_{L_0}$.
 And then the listener ought to implement $L_4$, and so on.
@@
 One computationally expensive option is to continue iterating such strategies until one reaches a fixed point equilibrium.
-But, while this guarantees each agent will behave normatively given the other agent's strategy, there is no guarantee that such a strategy will be optimal for the system as a whole; equilibria strategies make up only a small subspace of all possible strategies.
+But, while this guarantees each agent will behave normatively given the other agent's strategy, there is no guarantee that such strategies will be near the system's global optimum; equilibrium strategies make up only a small subspace of all possible strategies.
 Nor is there much evidence the humans find such equilibria; their behaviour in language games (and in other games XX p-beauty) can be well-modeled as implementing $S_n$ or $L_n$ for some small $n$ \cite{frank2012}.
-Here we adopt $n = 3$ for speakers and $n = 2$ for listeners.
+This paradox, then, is averted by simply recursing a small, finite number of times.
+The consequence is that one agent, implementing $S_n$, say, can be fully rational, while the other, implementing $L_{n-1}$, is nearly rational -- they're off by only a single recursion.
 
-Given such a model, it's conceptually straightforward for a listener to learn to understand a speaker -- the listener simply puts a prior on the speaker's lexicon, and updates it based on the speaker's utterances and whatever information the listener can glean about their intended meaning.
+But as soon as we attempt to add learning to this model, new paradoxes arise.
+Suppose the listener is a young child who is uncertain about the lexicon their partner is using.
+The obvious solution is for them to place a prior on the lexicon; they then update their posterior based on whatever utterances and contextual cues they observe, and in the mean time interpret each utterance by making their best guess, marginalizing out this uncertainty.
+This basic structure is the essence of previous models of Bayesian word learning (e.g., \cite{frank2009}).
+But when combined with the recursive pragmatic model, a new question arises: Given such a listener, what model should the speaker use?
+A rational speaker attempts to maximize the listener's likelihood of understanding, so if an uncertain listener interpets by marginalizing over some posterior, then a knowledgeable speaker should calculate this posterior and marginalize themselves, before applying the soft-max decision rule.
+But this means that the speaker is not using any particular lexicon at all, and a rational learning listener should therefore not be trying to infer the speaker's lexicon.
+In fact, such an uncertain listener's uncertainty dissolves; they may not know the lexicon, but they do know their posterior over the lexicon -- and because this is all that the speaker cares about, there is nothing for the listener to learn from the speaker's behaviour.
 
+Of course, in reality both parties will generally have some uncertainty, which makes things even worse.
+If we start from an uncertain listener with a prior over lexicons, then a first-level uncertain speaker needs a prior over priors on lexicons, a second-level uncertain listener needs a prior over priors over priors, etc.
+The original $L_0 \rightarrow S_1 \rightarrow \dots$ recursion was bad enough, but at least each step had a constant cost.
+This new recursion produces hyper$^n$-distributions for which inference almost immediately becomes intractable even in principle, since the dimensionality of the learning problem increases with each step.
+Yet, without this addition of new uncertainty at each level, the model would dissolve back into certainty as in the previous paragraph, making learning impossible.
+
+And one final problem is that in real life, the same agents alternate between speaking and listening, and generally use the same language to do so.
+There's nothing in these models to ensure this, and this is somehow reasonable: in principle, so long as I as a listener have a good model of you as a speaker, and I as a speaker have a good model of you as a listener, then we can communicate perfectly well, even if we use entirely different languages depending on who is talking.
+Yet most people do in fact speak languages that they have learned by listening to others, which suggests an additional constraint that should somehow be incorporated into learning: as I build a model of you as a speaker, it should also affect my model of myself as a speaker.
+
+\subsection{A ``social anxiety'' model of pragmatic word learning}
+
+Clearly we must deviate from pure normativity -- but how?
+Here, we propose to resolve the above problems with a model of speakers and listeners as rational agents afflicted by a kind of social anxiety.
+Specifically, each agent assumes: (a) There is some single, specific literal lexicon which everyone should be using, (b) and everyone else knows this lexicon, and believes that I know it as well, (c) but in fact I don't.
+Assumption (a) explains the lexicographer's illusion: naive language users are convinced and will vociferously that words have specific meanings, even though these meanings are unobservable to everyone who purportedly uses them.
+It also explains why children speak the language they hear, since observing your behaviour now gives me data about not just \textit{your} lexicon, but \textit{the} lexicon.
+Assumption (b) avoids the explosion of hyper$^n$-distributions: if agent $n$ knows the lexicon, they assume that all lower agents do as well, reducing to the original tractable model without uncertainty.
+And assumption (c) introduces a limited form of uncertainty, and thus the potential for learning.
+Everything else follows by applying standard Bayesian reasoning given these assumptions.
+
+Formally, let an unadorned $L$ and $S$ denote our socially-anxious listener and speaker.
+$L$ inverts their marginal distribution over the speaker; $S$ soft-maximizes their marginal distribution over the listener.
+With $L_i$ and $S_i$ as above, this gives:
 \begin{align*}
-  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{L's data}) \times \prior \\
-    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{L's data}) \times \prior \\
-    &= P_{L_n}(\obj | \word, \text{L's data}). \\
-  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{S's data}) - \text{cost}(\word)\big)\Big).
+  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
+    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior \\
+    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
+  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
 \end{align*}
+Here $L$'s data consists of $S$'s utterances together with whatever other information $L$ may have about their meaning (e.g.~from contextual clues) -- but by assumption (b), $L$ treats these as utterances as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus as being informative about the lexicon.
+Similarly, $S$ may observe $L$'s interpretations of various utterances, and will treat these as samples from $L_n$, not $L$.
+In several important cases, these assumptions are as close to rational as possible within the finite-recursion framework.
+If $L_n(\obj | \word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$ coincide, then  $S$ is fully normative and $L$ is off by only a single recursion.
+This in particular includes all previous experiments using this pragmatic reasoning framework, which considered only single-turn interactions where $\text{$L$'s data} = \text{$S$'s data} = \emptyset$.
+Alternatively, if one agent has enough data that their uncertainty is effectively non-existent, then it becomes normative for their partner to treat them as being fully knowledgeable, and this is a reasonable description of the child language learning case.
+In general, though, in arbitrary multi-turn interactions in which both agents have non-trivial uncertainty, this model induces complex and non-normative learning dynamics.
+In the remainder of this paper, we computationally investigate the model's behaviour in all three situations.\footnote{In our simulations throughout, we somewhat arbitrarily set $n = 2$ (the minimal value that produces all the qualitative phenomena), $\lambda = 3$, and assume that all agents have shared priors on the lexicon and full knowledge of the cost function.
+Inference is via importance sampling.}
 
--- if they have the same data, $S$ is rational and $L$ is as close to rational as is possible given a finite recursion depth, i.e., if $S$ recursed one fewer times then $L$ would be rational.
-In particular, this describes all of the one-shot language games that have been considered in the literature to date, where neither agent has the time to acquire any data.
--- if one actually does know the lexicon and the other is relatively ignorant, as in the child language learning case, then both are rational (with the same caveat).
--- but in general, in multi-turn interactions, their beliefs will display some complicated
-
-% This suggest a simple probabilistic model. Given any particular
-% decision procedure used by the Speaker $S$, the optimal way for the
-% Listener $L$ to judge the probability of each object is to invert this
-% decision procedure using Bayes' rule, taking into account the prior
-% likelihood of each object:
-% \begin{align*}
-% P_L(\obj | \word) &\propto P_S(\word | \obj) P_{\text{prior}}(\obj |
-% \text{context})
-% \end{align*}
-% $S$, meanwhile, being herself a rational Bayesian agent, chooses her
-% word so as to soft-maximize a utility function $U(\word | \obj)$. She
-% gains utility if $L$ is likely to understand her; she loses utility if
-% the word is particularly costly to say (e.g., it may be long).
-% \begin{align*}
-% P_S(\word | \obj) &\propto \exp \lambda U(\word | \obj) \\
-% U(\word | \obj) &= \log P_L(\obj | \word) - \text{cost}(\word)
-% \end{align*}
-% The degree to which she optimizes this utility is controlled by the
-% softmax parameter $\lambda$. If $\lambda = 1$ and all words cost the
-% same, then her probability of producing a word will be proportional to
-% the likelihood that it will cause $L$ to reach the correct
-% interpretation; as $\lambda$ increases, she becomes more likely to
-% choose the word which maximizes this likelihood. (We somewhat
-% arbitrarily set $\lambda = 3$ throughout.)
-
-% This model as stated satisfies various normative criteria for each
-% agent individually, but suffers from a serious flaw that is intrinsic
-% to such multi-agent situations: $S$'s decision procedure depends on
-% $L$'s interpretive procedure, and vice-versa. If $S$ and $L$ want to
-% both be optimal simultaneously, then they must restrict themselves to
-% a low-dimensional subset of all available strategies. They must each choose
-% a strategy with the property that optimal counterpart of its optimal
-% counterpart is the original strategy. Finding such an equilibrium is
-% computationally non-trivial, and, even if found, there is no guarantee
-% that it will be optimal in any global sense. \cite{frank2012}
-% therefore propose an approximation to this model, in which $L$ and $S$
-% each perform a finite number of recursions, with $S_i$ reasoning about
-% $L_{i-1}$, who reasons about $S_{i-2}$ and so forth, until reaching
-% $L_0$, who interprets words according to an arbitrary rule (their
-% \textit{literal meaning}). In our case, we instantiate this as a
-% matrix (the \textit{lexicon}) with one column per object, and in which
-% each row represents the meaning of a particular word as a set of
-% values in $(0, 1)$ such that $\sum_\obj \lex(\word, \obj) = 1$.\footnote{In this model, words refer directly to objects; more realistically, words should refer to some sort of abstract semantic features, which in turn refer to objects.
-However, this simplification is without loss of generalization; we can interpret this model as marginalizing over such a representation, with our literal $P(\obj | \word) = \sum_\text{features} P(\obj | \text{features}) P(\text{features} | \word)$.}  Formally, we have:
-% \begin{align*}
-% P_{L_0}(\obj | \word) &= \lex(\word, \obj) P_{\text{prior}}(\obj | \text{context}). \\
-% P_{S_i}(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_{i - 1}}(\obj | \word) - \text{cost}(\word)\big)\Big) \\
-% P_{L_i}(\obj | \word) &\propto P_{S_{i-1}}(\word | \obj) P_{\text{prior}}(\obj | \text{context})
-% \end{align*}
-% This approach neatly avoids the infinite regress, and does well at predicting human behaviour in language games \cite{frank2012} and other multi-agent games (XX p-beauty).
-The minimal degree of recursion required to produce the effects we consider here is $S_3$ and $L_2$, and this what we use in the following.
-
-% When we consider adding learning to this model, however, we encounter a new and related problem.
-It's straightforward in principle to put a prior on $\lex(\word, \obj)$, and allow an agent to learn the
-
-At each stage, $S_i$ defines a causal model for utterances described
-by a simple Bayes net, and $L_{i+1}$ does Bayesian inference over this net.
-But the likelihood function in $S_i$'s model is very complicated, and includes equations derived from considering how a hypothetical $L_{i-1}$ would interpret utterances if they were doing rational inference about another hypothetical causal agent $S_{i-2}$.
-
-extensional as marginalization over intensional
-
-we use $S_3$, $L_2$ and $\lambda = 3$.
-
-\subsection{Naive literal learner model}
-
-To see how
-Make $L_0$ uncertain about values in $\lex(\word, \obj)$, impose a prior, and then
-
-Needs feedback
-
-can't switch roles from speaker to listener
-
-knocks out the lexical uncertainty/horn implicature results (because you can only have uncertainty at one level)
-
-something wrong about taking evidence from the pragmatic speaker and using it to train the literal listener
-
-\subsection{``Social anxiety'' model}
+% \subsection{Naive literal learner model}
+%
+% To see how
+% Make $L_0$ uncertain about values in $\lex(\word, \obj)$, impose a prior, and then
+%
+% Needs feedback
+%
+% can't switch roles from speaker to listener
+%
+% knocks out the lexical uncertainty/horn implicature results (because you can only have uncertainty at one level)
+%
+% something wrong about taking evidence from the pragmatic speaker and using it to train the literal listener
 
@@
 \begin{tabular}{lccccc}
-\hline \\
-Phenomenon & Ref. & WL & PI & Lifted PI  & PIL \\
-\hline
-Scalar implicature & \cite{grice1975} &  & x &x & x\\
-Horn implicature & \cite{horn1984} & &  & x& x\\
+\toprule
+Phenomenon & Ref. & WL & PI & PI+U & This work \\
+\midrule
+Interpreting scalar implicature & \cite{grice1975} &  & x &x & x\\
+Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
 Cross-situational learning & \cite{smith2008} & x& & & x \\
-Disambiguation & \cite{markman1988} &x  & & & x \\
-Communicative equilibria & \cite{galantucci2005} & & & & x \\
-\hline
+Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
+Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
+Learning scalar implicature & & & & & x \\
+\bottomrule
 \end{tabular}
@@
 \caption{Empirical results and references.
-WL refers to the word learning model of \cite{frank2009}; PI refers to the pragmatic inference model of \cite{frank2012}; Lifted PI refers to the pragmatic inference model of \cite{bergen2012}, and PIL refers to the pragmatic inference and learning model described here.
-}
-
+WL refers to the word learning model of \cite{frank2009}; PI refers to the recursive pragmatic inference model of \cite{frank2012}; PI+U refers to the pragmatic inference model of \cite{bergen2012} which includes lexical uncertainty, marginalizes it out, and then recurses.}
 \end{table}
@@
 
-The recursive pragmatic reasoning portions of our model capture set of findings on scalar implicature effectively. Consider a representation of scalar implicature as a matrix 
+The recursive pragmatic reasoning portions of our model capture set of findings on scalar implicature effectively. Consider a lexicon
 $\left[
     \begin{array}{cc}
-      1 & 1 \\
+      0.5 & 0.5 \\
       0 & 1
@@
   \right]$,
-where rows show the terms ``some'' and ``all'' and columns indicate worlds where {\sc some} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case.
-Assuming uniform costs for the two words and uniform priors, each recursion of listener modeling speaker reduces the probability that ``some'' will be used to refer to {\sc all}.
-Using the fixed parameters employed in our other simulations results in speaker use probabilities of
+where rows show the words ``some'' and ``all'' and columns indicate worlds where {\sc some but not all} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case.
+Assuming uniform costs for the two words, a rational speaker confronted with such a lexicon will prefer to say ``some'' only when they mean \textsc{some but not all}, and this is therefore the interpretation that a rational listener will place on ``some'' as well.
+To simulate this situation, we place a Dirichlet(10, 10) prior on the meaning of \textit{some} and a Dirichlet(1, 10) prior on the meaning of \textit{all}, and find this gives speaker use probabilities of
 $\left[
@@
 \subsection{Horn implicature}
+\label{sec:horn-implicature}
 
-XX note the relation to BGL model; they need uncertainty in the middle of the pragmatics specifically, which is where we end up putting it for other reasons.
+% XX note the relation to BGL model; they need uncertainty in the middle of the pragmatics specifically, which is where we end up putting it for other reasons.
 And in Bayesian framework uncertainty and learning must always coincide.
@@
 
-Horn implicatures are inferences where the critical factor is not the extension of the words in context (as in scalar implicatures), but instead the relative cost of two expressions.
-In Horn's example, ``Lee got the car to stop'' implies that Lee used an unusual method (e.g.
-not the brakes) because, had he used the brakes, the speaker would have chosen the simpler and shorter (less costly) expression, ``Lee stopped the car.'' In other words, the costly expression is assumed to be associated with the unusual outcome.
+Consider a world in which contains two words and two types of objects.
+One word is expensive to use, and one is cheap (call them ``expensive'' and ``cheap'' for short).
+One object type is common and one is rare; denote these \textsc{common} and \textsc{rare}.
+Intuitively, there are two possible communicative systems here: a good system where ``cheap'' means \textsc{common} and ``expensive'' means \textsc{rare}, and a bad system where the opposite holds.
+Obviously we would prefer to use the good system, but it has historically proven very difficult to derive this conclusion in a game theoretic setting, because both systems are stable equilibria; if our partner uses the bad system, then we would rather follow and communicate at some cost than switch to the good system and fail entirely \cite{bergen2012}.
 
-Such an implicature again arises naturally from our model.
-We create a situation in which expressions have differential costs (e.g.
-$[.5 1.0]$) and referents have differential probabilities (e.g.
-$[.8 .2]$).
-Speakers in such a situation choose simple words to refer to common referents, e.g.
-$\left[
-    \begin{array}{cc}
-      .88 & .54 \\
-      .12 & .46
-    \end{array} 
-  \right]$.\footnote{Rows indicate words and columns indicate referents.
-For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.} Similarly, listeners interpret ``cheap'' and ``expensive'' words to refer to common and rare objects respectively:
-$\left[
-    \begin{array}{cc}
-      .77 & .23 \\
-      .65 & .35
-    \end{array} 
-  \right]$.
+Humans, however, unlike traditional game theoretic models, make the inference that if two utterances seem equivalent except that one is more costly to utter, then the costly utterance refers to an unusual event.
+We call this pattern \textit{Horn implicature}, after \cite{horn1984}.
+For instance, ``Lee got the car to stop'' implies that Lee used an unusual method (e.g.~not the brakes) because, had he used the brakes, the speaker would have chosen the simpler and shorter (less costly) expression, ``Lee stopped the car'' \cite{horn1984}.
+Surprisingly, Bergen et al \cite{bergen2012} show that the key to achieving this favorable result is ignorance.
+If a listener assigns equal probability to her partner using the good system or the bad system, then their best bet is to estimate $P_S(\word | \obj)$ as the average of $P_S(\word | \obj, \text{good system})$ and $P_S(\word | \obj, \text{bad system})$.
+These might seem to cancel out, but in fact they do not.
+In the good system, the speaker has higher expected utility than the bad system; therefore, a soft-max agent in the bad system has noiser behaviour than in the good system, and their behaviour in the good system dominates the average, breaking the symmetry.
+Similar reasoning applies to an uncertain speaker.
 
@@
 %        [ 0.64702551,  0.35297449]])
+
+The model of \cite{bergen2012} differs from ours in that they then apply further layers of pragmatic recursion on top of these marginal distributions, which as described above removes the possibility of learning; in their model it is only hypothetical agents who are ignorant, while the actual speaker and listener know everything there is to know about each other's behaviour.
+However, the core result carries over to our model.
+For example, with a uniform prior over lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$, $\text{cost}(\text{``cheap''}) = 0.5$, $\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the appropriate way: Despite total ignorance of the conventional system, speakers prefer to use simple words for common referents ($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88, P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) = 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) = 0.65$).
 
@@
 
-\subsection{Emergence of communicative equilibria}
+XX maybe make the sections ``Pure pragmatics'' (scalar and horn implicature), ``Pure learning'' (cross-situational), ``Learning and pragmatics'' (subsection 1: disambiguation and learning scalar implicature, subsection 2: emergence with and without horn implicature)
+
+We now turn to phenomena which have not been modeled by previous pragmatic or learning models, as they seem to intrinsically require both together.
+
+\subsection{Emergence of efficient communicative conventions}
+
+If two humans are forced to play a game which depends on communication, but are allowed to communicate only via a set of novel symbols which have no pre-existing conventional meaning, then over time they will generally derive a set of new conventions which allow them to accomplish their task \cite{galantucci2005}.
+Or if, less drastically, a task requires two people to repeatedly refer to novel objects, they will implicitly through the course of the interaction invent conventional names for them (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}).
+From a pure learning perspective this is anomalous: since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour.
+Furthermore, even if only one partner is producing the novel expressions, their behaviour still becomes more regular (conventional) over time, which would seem to rule out a role for learning -- even if there is some pattern in the expressions the speaker chooses to use, there is certainly nothing for the \textit{speaker} to learn by observing these patterns, and thus their behaviour should not change over time.
+An obvious hack might be to simply stipulate that speakers take their own utterances as training data for further learning, but this would fail to explain yet another wrinkle in this phenomenon: speakers only regularize their production over time if they have some kind of feedback indicating that the listener understands; if such feedback is removed, the phenomenon is dramatically reduced (XX krauss 1966, branigan catchpole pickering 2011).
 
@@
 \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
-\caption{\label{fig:emergence} Simulations....}
+\caption{\label{fig:emergence} Simulations of two learning pragmatic agents playing a naming game.
+At each stage, $S$ and $L$ have different, possibly contradictory posteriors over the literal lexicon.
+From these posteriors we derive the probability $P(\text{L understands S})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.}
 \end{figure}
 
-\subsection{Lexicalization of Horn implicatures}
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.3\textwidth]{figures/emergence-average.pdf}
+\hspace{0.5in}
+\includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
+\caption{\label{fig:emergence-average} Behaviour of learning agents playing naming games as in Figs.~\ref{fig:emergence} and \ref{fig:horn}, averaged across 300 dialogues.
+Left: Average communicative success by game type and dialogue turn.
+Right: The proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have not only converged on using the ``good'' or ``bad'' lexicons, but believe that these are the literal meanings of the words.}
+\end{figure}
+
+To model such phenomena, we imagine two agents playing the simple referential game introduced above, in which on each turn the speaker is assigned a target object, they utter some word referring to this object, the listener makes a guess at the object, and then, critically, the speaker observes the listener's guess and the listener receives feedback indicating the correct answer (i.e., the speaker's intended meaning), which they use to update their posterior over lexicons before proceeding to the next trial.
+As in \cite{krauss1964,clark1986}, the speaker and listener remain fixed in the same role throughout.
+Fig.~\ref{fig:emergence} shows the result of simulating several such games when both parties begin in total ignorance.
+Notice that: (a) agents' performance begins at chance, but quickly rises; (b) they tend towards structured, sparse lexicons with a one-to-one correspondence between objects and words; (c) as the speaker and listener have entirely different data (the listener's interpretations and the speaker's intended meaning, respectively), unlucky early guesses can lead them to believe in entirely contradictory lexicons -- but they generally recover and converge despite this.
+
+What's happening here is that each agent is, effectively, using their partner's behaviour as a basis for forming superstitions about the underlying lexicon that they assume must exist.
+Since they each then act on these superstitions, and their partner uses the actions to form new superstitions, soon they converge to using similar lexicons, and what started as a superstition becomes normatively correct.
+
+What happens if we remove feedback from the listener to the speaker?
+If the speaker is not allowed to observe the listener's guess, then the speaker does no learning and their posterior over lexicons remains perfectly uniform.
+They therefore choose words by simply flipping a coin, and performance remains at chance.
+
+The role of learning here is obvious, but what does the pragmatics part of the model do?
+It creates the sparse structure in the observed lexicons: at each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful, since if the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well.
+But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon.
+The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.
+
+% XXXX what can we say about the role pragmatics plays here?
+It's annoyingly non-trivial to sample from the equivalent model where the speaker just samples from a multinomial distribution and the listener inverts it.
+The listener's data is samples from a DM so they have conjugacy.
+But the speaker's data is listener interpretations, which necessarily do one tiny pragmatic inference step from using Bayes rule, which breaks conjugacy...
+
+\subsection{Lexicalization and loss of Horn implicatures}
+\label{sec:horn-emergence}
 
@@
 \end{figure}
+
+A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/''expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}.
+Recall that in this domain, pragmatic reasoning causes each agent to start out independently be biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}.
+Because this bias
 
@@
 
+allows us to preserve most of the normative Bayesian framework within the multi-agent setting, while avoiding paradoxes and allowing both pragmatic reasoning and learning.
+The resulting model not only unifies previous models of each phenomenon separately, it also accounts for a range of
+
+the deviations from rationality -- assumption that words have meanings, suspicion that everyone else knows what they are -- are highly interpretable and intuitively compelling, and while like any deviations from rationality they necessarily introduce biases, these biases appear in this case to improve the global performance of the system.
+
+if you're trying to collaborate with an agent who soft-maximizes utility, but there are multiple possible rules which might underlie their utility function, you should assign more weight to those rules which provide a sharp distinction in utility for different actions -- which will generally be the rules that allow for higher expected utility.
+this actually explains both the uniform prior and the horn implicature emergence results.
+
+provides new insight into use of pragmatic reasoning in word learning, the learning of literal meanings given pragmatically strengthened input, and makes the interesting prediction that some implicatures are intrinsically more stable, while others if consistently supported will tend to enter the lexicon and become part of the literal meaning of the relevant words.
+
+zipf, piantidosi et al
+
 %\subsubsection*{Acknowledgments}

### 2013-05-31 figures/non-emergence.py

@@ -1,9 +1,53 @@
 from nips import *
 
-# with a flat prior and softmax = 1, S1 becomes equivalent to old S0 --
-# basically just sampling the (renormalized) prior
+from scipy.special import gammaln
 
-d = conpact2.MemoizedDomain(max_utterance_length=1, adjectives=2, objects=2,
-                            softmax_weight=1)
-dialogues = simulate_dialogues(4, 20, d)
-show_dialogues("non-emergence-%i.pdf", dialogues)
+# speaker and listener each have dirichlet priors on a set of object -> word
+# multinomial distributions.
+# speaker to produce a word samples from dirichlet-multinomial
+# listener conditions on the word to guess the meaning
+# speaker increments this (word, interpretation) pair in its Dirichlet prior
+# listener increments (word, target) pair in its Dirichlet prior
+
+SIZE = 2
+
+def DM_dist(alpha):
+    # Joint density for N samples from a dirichlet-multinomial
+    # distribution with parameters alpha[i], and n_k
+    # repeats of each outcome is
+    #   Z * prod_k Gamma(n_k + alpha[k])
+    # where:
+    #   Z = Gamma(A) / Gamma(N + A) * prod_k 1/Gamma(alpha[k])
+    #   A = sum_k(alpha[k])
+    #   N = sum_k(n_k)
+    # For us N = 1
+    # So basically P(i|alpha) \propto gamma(1 + alpha[i])/gamma(alpha[i])
+    dist = gammaln(1 + alpha) - gammaln(alpha)
+    dist -= np.logaddexp.reduce(dist)
+    return dist
+
+def marginal_S_dist(posterior):
+    dist = np.empty((SIZE, SIZE))
+    for obj in xrange(SIZE):
+        dist[:, obj] = DM_dist(posterior[:, obj])
+    return dist
+
+def L_dist(posterior):
+
+
+def simulate_non_pragmatic(r, turns, prior=None):
+    speaker_posterior = np.ones((SIZE, SIZE))
+    listener_posterior = np.ones((SIZE, SIZE))
+
+    for i in xrange(turns):
+        target = r.randint(SIZE)
+        S_dist = marginal_S_dist(speaker_posterior)
+        L_dist = marginal_S_dist(listener_posterior)
+        L_dist -= np.logaddexp.reduce(L_dist, axis=1)[:, np.newaxis]
+        assert np.allclose(np.logaddexp.reduce(S_dist, axis=0), 1)
[…]

### 2013-05-31 praglearn.tex

@@
 \usepackage{booktabs}
-\usepackage{subcaption}
+\usepackage[raggedright]{sidecap}
 
@@
 \midrule
-Interpreting scalar implicature & \cite{grice1975} &  & x &x & x\\
+Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
 Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
@@
 Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
+Disambiguation without mapping & \cite{horst2008} &  & & & x \\
 Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
-Learning scalar implicature & & & & & x \\
-\bottomrule
+Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
+Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
+\hline
 \end{tabular}
@@
 So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs.
+We use the term {\it specificity implicature} here (following \cite{bergen2012}) to denote that we are agnostic about whether such implicatures are conventionalized or whether they use a contextually-defined scale.
 
-The recursive pragmatic reasoning portions of our model capture set of findings on scalar implicature effectively. Consider a lexicon
+The recursive pragmatic reasoning portions of our model capture findings on specificity implicature. Consider a representation of scalar implicature as a matrix 
 $\left[
@@
 
-Humans, however, unlike traditional game theoretic models, make the inference that if two utterances seem equivalent except that one is more costly to utter, then the costly utterance refers to an unusual event.
+Humans, however, unlike traditional game theoretic models, do make the inference that if two utterances seem equivalent except that one is more costly to utter, then the costly utterance refers to an unusual event.
 We call this pattern \textit{Horn implicature}, after \cite{horn1984}.
@@
 Similar reasoning applies to an uncertain speaker.
+
+% Mike's text:
+% Such an implicature again arises naturally from our model.
+We create a situation in which expressions have differential costs (e.g.
+$[.5 1.0]$) and referents have differential probabilities (e.g.
+$[.8 .2]$).
+At the settings used throughout the manuscript, speakers in such a situation choose simple words to refer to common referents at different rates than for uncommon references.
+This situation breaks the symmetry though does not result in as dramatic a diagonal as in the specificity implicature, e.g.
+for the speaker, \footnote{Rows indicate words and columns indicate referents.
+For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.}
+% $\left[
+%     \begin{array}{cc}
+%       .88 & .54 \\
+%       .12 & .46
+%     \end{array} 
+%   \right]$ 
+% and for the listener,
+% $\left[
+%     \begin{array}{cc}
+%       .77 & .23 \\
+%       .65 & .35
+%     \end{array} 
+%   \right]$. See Section \ref{sec:hornlearn} for the consequences of repeating this implicature until it is lexicalized.
 
@@
 However, the core result carries over to our model.
-For example, with a uniform prior over lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$, $\text{cost}(\text{``cheap''}) = 0.5$, $\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the appropriate way: Despite total ignorance of the conventional system, speakers prefer to use simple words for common referents ($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88, P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) = 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) = 0.65$).
+For example, with a uniform prior over lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$, $\text{cost}(\text{``cheap''}) = 0.5$, $\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the appropriate way: Despite total ignorance about the conventional system, speakers prefer to use simple words for common referents ($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88, P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) = 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) = 0.65$).
+This preference is somewhat weak, at least to start -- see \S\ref{sec:horn-emergence}
 
@@
 
-A recent line of work in language acquisition has 
-% \begin{figure}[t]
-%   \centering
-%   \includegraphics[width=0.20\textwidth]{figures/cross-sit.pdf}
-%   \caption{Cross-situational simulation}
-%   \label{fig:cross-sit}
-% \end{figure}
+Even very young children can learn words from individually ambiguous situations, e.g.
+where multiple words are spoken and multiple objects are present \cite{smith2008}.
+While no individual situation gives away the mapping, the pattern of co-occurrences between words and objects allows for learning.
+Our current model reproduces this pattern of results, converging to correct lexical mappings from ambiguous evidence.
+A representative simulation is shown in Fig.
+\ref{fig:cross-sit}.
+
+\begin{SCfigure}[2.5]
+  \includegraphics[width=0.2\textwidth]{figures/cross-sit-mod.pdf}
+  \caption{Results from a simulation of cross-situational word learning.
+The posterior distribution over the learner's lexicon (diagonal entries are correct for word/object pairs 1 -- 3; word D is a distractor).
+Training data were situations including two of the three objects and a single word, either the distractor or a word matching one of the two objects.
+Crucially, at no point does the learner observe the meaning of any word; they observe only that a word was produced in each situation.}
+  \label{fig:cross-sit}
+\end{SCfigure}
 
 \subsection{Disambiguation using known words}
+
+\begin{SCfigure}[1]
+  \centering
+  % \includegraphics[width=0.24\textwidth]{figures/ME-1dax.pdf}
+  % \includegraphics[width=0.24\textwidth]{figures/ME-flat-10dog-10dax.pdf}
+  % \includegraphics[width=0.24\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
+  % \includegraphics[width=0.24\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
+\includegraphics[width=.6\textwidth]{figures/ME-composite-small.pdf}
+  \caption{Results for simulations of the ``disambiguation'' phenomenon.
+A novel word is used in a situation with a novel object and a familiar object.
+Each plot shows the joint posterior probability that the familiar word refers to the familiar object (horizontal) and that the novel word refers to the familiar object (vertical).
+Left plot shows the posterior after a single situation; right plot shows multiple situations.
+%The right middle plot shows the posterior after a single situation given either a prior that makes it likely that there are words referring to both the novel and familiar objects (perhaps because the two objects are physically similar; the right plot shows a situation where the prior makes it unlikely that a word refers to both (objects dissimilar).
+}
+  \label{fig:me}
+\end{SCfigure}
+
+Children, when presented with both a novel and a familiar object (e.g.
+an eggbeater and a ball), will treat a novel label (e.g.
+``dax'') as referring to the novel object, for example by supplying the eggbeater when asked to ``give me the dax'' \cite{markman1988}.
+This phenomenon is sometimes referred to as ``mutual exclusivity'' or even ``Fast mapping,'' and many distinct theoretical accounts have been proposed in the psychological literature.
+Simple probabilistic word learning models can produce a similar pattern of findings \cite{frank2009}, often due to the kinds of ``explaining away'' type phenomena that are observed in such models.
+But all such models assume that learners in fact retain the mapping between word and object that is demonstrated in the experimental situation; this observation is contradicted, however, by evidence that children often do not retain the mappings that are demonstrated by their inferences in the moment \cite{horst2008}.
+
+Our model provides an intriguing possible explanation of this finding: given a single disambiguation situation, the model gives a substantial probability (e.g.
+75\%) that the speaker is referring to the novel object.
+Nevertheless, integrating across lexicons reveals that this probability is not due to learning that the novel word refers to this object (as shown in Fig.
+\ref{fig:me}, left): instead, it comes in part from a specificity implicature because the listener knows that the familiar word \emph{does not} refer to the novel object--hence the novel word (even if it refers to the familiar object as well) is the best way to refer to the novel object.
+Nevertheless, on repeated exposure to the same novel word, novel object situation, the model does lexicalize this inference (Fig.
+\ref{fig:me}, right).
 

### 2013-05-31 praglearn.tex
mini edit x
text edits

@@
 
-\section{Experiments: Pragmatic and word learning phenomena}
+\section{Results: Pragmatic inference}
 
@@
 \subsection{Horn implicature}
+
 \label{sec:horn-implicature}
@@
 Similar reasoning applies to an uncertain speaker.
-
-% Mike's text:
-% Such an implicature again arises naturally from our model.
-We create a situation in which expressions have differential costs (e.g.
-$[.5 1.0]$) and referents have differential probabilities (e.g.
-$[.8 .2]$).
-At the settings used throughout the manuscript, speakers in such a situation choose simple words to refer to common referents at different rates than for uncommon references.
-This situation breaks the symmetry though does not result in as dramatic a diagonal as in the specificity implicature, e.g.
-for the speaker, \footnote{Rows indicate words and columns indicate referents.
-For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.}
-% $\left[
-%     \begin{array}{cc}
-%       .88 & .54 \\
-%       .12 & .46
-%     \end{array} 
-%   \right]$ 
-% and for the listener,
-% $\left[
-%     \begin{array}{cc}
-%       .77 & .23 \\
-%       .65 & .35
-%     \end{array} 
-%   \right]$. See Section \ref{sec:hornlearn} for the consequences of repeating this implicature until it is lexicalized.
 
@@
 Our current model reproduces this pattern of results, converging to correct lexical mappings from ambiguous evidence.
-A representative simulation is shown in Fig.
-\ref{fig:cross-sit}.
 
-\begin{SCfigure}[2.5]
-  \includegraphics[width=0.2\textwidth]{figures/cross-sit-mod.pdf}
-  \caption{Results from a simulation of cross-situational word learning.
+% A representative simulation is shown in Fig. \ref{fig:cross-sit}.
+
+% \begin{SCfigure}[2.5]
+%   \includegraphics[width=0.2\textwidth]{figures/cross-sit-mod.pdf}
+%   \caption{Results from a simulation of cross-situational word learning.
 The posterior distribution over the learner's lexicon (diagonal entries are correct for word/object pairs 1 -- 3; word D is a distractor).
@@
 Crucially, at no point does the learner observe the meaning of any word; they observe only that a word was produced in each situation.}
-  \label{fig:cross-sit}
-\end{SCfigure}
+%   \label{fig:cross-sit}
+% \end{SCfigure}
+
+\section{Results: Learning in pragmatic contexts}
 
@@
 
-\section{Experiments 2: Lexical learning in pragmatic contexts}
-
-XX maybe make the sections ``Pure pragmatics'' (scalar and horn implicature), ``Pure learning'' (cross-situational), ``Learning and pragmatics'' (subsection 1: disambiguation and learning scalar implicature, subsection 2: emergence with and without horn implicature)
+% XX maybe make the sections ``Pure pragmatics'' (scalar and horn implicature), ``Pure learning'' (cross-situational), ``Learning and pragmatics'' (subsection 1: disambiguation and learning scalar implicature, subsection 2: emergence with and without horn implicature)
 

### 2013-05-31 praglearn.tex
edits

--- 2013-05-31 praglearn.tex
@@
 
+% You know, this is almost exactly like a quantitatively graded version of Pascal's wager.
+
 % Mike's text:
@@
 At each stage, $S$ and $L$ have different, possibly contradictory posteriors over the literal lexicon.
-From these posteriors we derive the probability $P(\text{L understands S})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.}
+From these posteriors we derive the probability $P(\text{$L$ understands $S$})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.}
 \end{figure}
@@
 What's happening here is that each agent is, effectively, using their partner's behaviour as a basis for forming superstitions about the underlying lexicon that they assume must exist.
-Since they each then act on these superstitions, and their partner uses the actions to form new superstitions, soon they converge to using similar lexicons, and what started as a superstition becomes normatively correct.
+Since they then each act on these superstitions, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a superstition becomes normatively correct.
 
@@
 The role of learning here is obvious, but what does the pragmatics part of the model do?
-It creates the sparse structure in the observed lexicons: at each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful, since if the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well.
+First, it makes it possible for the speaker to learn from listener interpretations, which is not possible without an explicit model of the speaker as a rational agent.
+Second, it creates the sparse structure in these observed lexicons: at each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful, since if the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well.
 But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon.
@@
 
-A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/''expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}.
-Recall that in this domain, pragmatic reasoning causes each agent to start out independently be biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}.
-Because this bias
+A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}.
+Recall that in this domain, pragmatic reasoning causes each agent to start out independently biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}, and that this bias arises exactly because their prior on literal meaning is uniform and uninformative.
+A fully rational listener, therefore, observing our uncertain speaker using this mapping, would discount it as arising from this bias, and conclude that the speaker in fact did not know which convention was in use.
+Our socially anxious listener, however, believes that the speaker does know which convention is in use, and therefore tends to take their behaviour as evidence that the good system is in fact in use.
+Similarly, the socially anxious speaker will wager that since on average they will succed more often if the listener is using the good system, they might as well try it -- and when they succeed, they take this as evidence that the listener was in fact using the good system all along.
+As a result, dyads in this game not only end up converging onto a stable lexicon, but onto specifically the good lexicon, at a rate far above chance.
 

### 2013-05-31 praglearn.tex

@@
 % You know, this is almost exactly like a quantitatively graded version of Pascal's wager.
-
-% Mike's text:
-% Such an implicature again arises naturally from our model.
-We create a situation in which expressions have differential costs (e.g.
-$[.5 1.0]$) and referents have differential probabilities (e.g.
-$[.8 .2]$).
-At the settings used throughout the manuscript, speakers in such a situation choose simple words to refer to common referents at different rates than for uncommon references.
-This situation breaks the symmetry though does not result in as dramatic a diagonal as in the specificity implicature, e.g.
-for the speaker, \footnote{Rows indicate words and columns indicate referents.
-For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.}
-% $\left[
-%     \begin{array}{cc}
-%       .88 & .54 \\
-%       .12 & .46
-%     \end{array} 
-%   \right]$ 
-% and for the listener,
-% $\left[
-%     \begin{array}{cc}
-%       .77 & .23 \\
-%       .65 & .35
-%     \end{array} 
-%   \right]$. See Section \ref{sec:hornlearn} for the consequences of repeating this implicature until it is lexicalized.
 

### 2013-05-31 praglearn.tex
revisions on simulation figures and on model section

@@
 
-\title{Learning words by recursive pragmatic reasoning about other agents: The blessing of social anxiety}
+\title{Learning words by recursive pragmatic reasoning about other agents}
+% : The blessing of social anxiety - not a huge fan of colon titles
 
@@
 \begin{abstract}
-  Adult language users are remarkably good at making inferences about speakers' intentions in context, and children learning their native language also display substantial pragmatic skill in acquiring the meanings of unknown words.
+Language users are remarkably good at making inferences about speakers' intentions in context, and children learning their native language also display substantial pragmatic skill in acquiring the meanings of unknown words.
 While pragmatic inference and word learning have both been characterized in probabilistic terms, no current models make appropriate inferences about the meanings of words in pragmatically-rich inferential contexts.
@@
 
-%\section{Introduction}
+\section{Introduction}
 
@@
 We begin by describing previous models in each domain as well as proposals for bringing the two together.
-We next simulate findings from the domains of pragmatic inference, word learning, and their intersection (summarized in Table \ref{tab:results}).
+We next simulate findings from the domains of pragmatic inference, word learning, and their intersection.
 We end by discussing the implications of our results for work on language change.
@@
 
-Consider a simple communicative game: two participants each,
-separately, view identical arrays of objects. On the
-\textit{Speaker's} screen, one object is highlighted; their goal is to
-get the \textit{Listener} to click on this item. To do this, they have
-available a fixed, finite set of words; they must pick one. The
-Listener then receives this word, and attempts to guess which object
+We model a standard communication game \cite{krauss1964,clark1986}: two participants each,
+separately, view identical arrays of objects.
+On the \textit{Speaker's} screen, one object is highlighted; their goal is to get the \textit{Listener} to click on this item.
+To do this, they have available a fixed, finite set of words; they must pick one.
+The Listener then receives this word, and attempts to guess which object
 the Speaker meant by it.
+In the psychology literature, as in real-world interactions, games are typically iterated; one view of our contribution here is as a generalization of one-shot models \cite{frank2012,bergen2012} to the iterated context.
 
-When modelling human behaviour, it's useful to begin with optimal models; this has the advantage that when we are then forced to deviate from the purely rational approach, it's generally in a theoretically interpretable way.
-However, multi-agent interactions are notoriously difficult to model in a normative fashion without falling prey to paradox.
+\subsection{Paradoxes in optimal models of pragmatic learning}
+% When modelling human behaviour, it's useful to begin with optimal models; this has the advantage that when we are then forced to deviate from the purely rational approach, it's generally in a theoretically interpretable way.
+However,
+Multi-agent interactions are difficult to model in a normative or optimal framework without falling prey to paradox.
 For example, consider a simple model of the agents in the above game.
-First we define a \textit{literal listener} agent denoted $L_0$.
-This agent has a \textit{lexicon} which probabilistically maps words to meanings; specifically, it assigns each word $w$ a vector describing the extent to which this word provides evidence for each object\footnote{In this model, words refer directly to objects; more realistically, words should refer to some sort of abstract semantic features, which in turn refer to objects.
-However, our simplification is without loss of generalization, because we can interpret our model as marginalizing over such a representation, with our literal $P_\lex(\obj | \word) = \sum_\text{features} P(\obj | \text{features}) P_\lex(\text{features} | \word)$.}.
+First we define a \textit{literal listener} $L_0$.
+This agent has a \textit{lexicon} which probabilistically maps words to meanings; specifically, it assigns each word $w$ a vector describing the extent to which this word provides evidence for each object\footnote{We assume words refer directly to objects, rather than to abstract semantic features.
+Our simplification is without loss of generalization, however, because we can interpret our model as marginalizing over such a representation, with our literal $P_\lex(\obj | \word) = \sum_\text{features} P(\obj | \text{features}) P_\lex(\text{features} | \word)$.}.
 Each entry in the vector is in $(0, 1)$, and together they sum to 1:
@@
 \end{align*}
-Confronted with such a listener, a rational (but noisy) speaker should attempt to choose a word which soft-maximizes the probability that the listener will assign to the target object -- possibly modulated by the effort or cost associated with producing this word:
+Confronted with such a listener, a probabilistically rational speaker should attempt to choose a word which soft-maximizes the probability that the listener will assign to the target object---modulated by the effort or cost associated with producing this word:
 \begin{align*}
@@
 \end{align*}
-But if this is how the speaker will behave, then it's not rational for a listener to implement the $L_0$ strategy.
-Instead, to best determine what object was intended, they should use Bayes rule to invert the speaker's decision procedure:
+But given this speaker, then the naive $L_0$ strategy is not optimal. Instead, listeners should use Bayes rule to invert the speaker's decision procedure \cite{frank2012}:
 \begin{align*}
@@
 Now the paradox becomes apparent.
-Given such a listener, it's no longer rational for the speaker to implement strategy $S_1$; instead, they should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead of $P_{L_0}$.
-And then the listener ought to implement $L_4$, and so on.
+Given such a listener, it is no longer optimal for speakers to implement strategy $S_1$; instead, they should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead of $P_{L_0}$.
+And then listeners ought to implement $L_4$, and so on.
 
-One computationally expensive option is to continue iterating such strategies until one reaches a fixed point equilibrium.
-But, while this guarantees each agent will behave normatively given the other agent's strategy, there is no guarantee that such strategies will be near the system's global optimum; equilibrium strategies make up only a small subspace of all possible strategies.
-Nor is there much evidence the humans find such equilibria; their behaviour in language games (and in other games XX p-beauty) can be well-modeled as implementing $S_n$ or $L_n$ for some small $n$ \cite{frank2012}.
-This paradox, then, is averted by simply recursing a small, finite number of times.
-The consequence is that one agent, implementing $S_n$, say, can be fully rational, while the other, implementing $L_{n-1}$, is nearly rational -- they're off by only a single recursion.
+One computationally expensive option is to continue iterating such strategies until reaching a fixed point equilibrium.
+While this strategy guarantees that each agent will behave normatively given the other agent's strategy, there is no guarantee that such strategies will be near the system's global optimum; equilibrium strategies make up only a small subspace of all possible strategies.
+Nor is there much evidence the humans find such equilibria; their behaviour in language games (and in other games \cite{camerer2004}) can be well-modeled as implementing $S_n$ or $L_n$ for some small $n$ \cite{frank2012}.
+Following this work, we  recurse a finite (small) number of times.
+The consequence is that one agent, implementing $S_n$, is fully optimal with respect to the game, while the other, implementing $L_{n-1}$, is nearly optimal---off by only a single recursion.
 
-But as soon as we attempt to add learning to this model, new paradoxes arise.
+As soon as we attempt to add uncertainty about the meanings of words to such a model, new paradoxes arise.
 Suppose the listener is a young child who is uncertain about the lexicon their partner is using.
 The obvious solution is for them to place a prior on the lexicon; they then update their posterior based on whatever utterances and contextual cues they observe, and in the mean time interpret each utterance by making their best guess, marginalizing out this uncertainty.
-This basic structure is the essence of previous models of Bayesian word learning (e.g., \cite{frank2009}).
+This basic structure is captured in previous models of Bayesian word learning \cite{frank2009}.
 But when combined with the recursive pragmatic model, a new question arises: Given such a listener, what model should the speaker use?
-A rational speaker attempts to maximize the listener's likelihood of understanding, so if an uncertain listener interpets by marginalizing over some posterior, then a knowledgeable speaker should calculate this posterior and marginalize themselves, before applying the soft-max decision rule.
-But this means that the speaker is not using any particular lexicon at all, and a rational learning listener should therefore not be trying to infer the speaker's lexicon.
-In fact, such an uncertain listener's uncertainty dissolves; they may not know the lexicon, but they do know their posterior over the lexicon -- and because this is all that the speaker cares about, there is nothing for the listener to learn from the speaker's behaviour.
-
-Of course, in reality both parties will generally have some uncertainty, which makes things even worse.
+A rational speaker attempts to maximize the listener's likelihood of understanding, so if an uncertain listener interpets by marginalizing over some posterior, then a knowledgeable speaker should calculate this posterior and marginalize themselves.
+But then the speaker is not using any particular lexicon at all, and a rational learning listener should therefore not be trying to infer the speaker's lexicon.\footnote{Of course, in reality both parties will generally have some uncertainty, making the situation even worse.
 If we start from an uncertain listener with a prior over lexicons, then a first-level uncertain speaker needs a prior over priors on lexicons, a second-level uncertain listener needs a prior over priors over priors, etc.
@@
 This new recursion produces hyper$^n$-distributions for which inference almost immediately becomes intractable even in principle, since the dimensionality of the learning problem increases with each step.
-Yet, without this addition of new uncertainty at each level, the model would dissolve back into certainty as in the previous paragraph, making learning impossible.
+Yet, without this addition of new uncertainty at each level, the model would dissolve back into certainty as in the previous paragraph, making learning impossible.}% In fact, such an uncertain listener's uncertainty dissolves; they may not know the lexicon, but they do know their posterior over the lexicon -- and because this is all that the speaker cares about, there is nothing for the listener to learn from the speaker's behaviour.
 
-And one final problem is that in real life, the same agents alternate between speaking and listening, and generally use the same language to do so.
-There's nothing in these models to ensure this, and this is somehow reasonable: in principle, so long as I as a listener have a good model of you as a speaker, and I as a speaker have a good model of you as a listener, then we can communicate perfectly well, even if we use entirely different languages depending on who is talking.
+To add insult to injury, one final problem is that there is nothing constraining speaker and listener to use the same language under this model.
+In principle, so long as the listener has a good model of the speaker, and the speaker has a good model of the listener, then communication succeeds.
+This result clearly presents a dramatic mismatch to typical human communication, which almost never proceeds with distinct languages spoken by each participant.
+
+% .
 Yet most people do in fact speak languages that they have learned by listening to others, which suggests an additional constraint that should somehow be incorporated into learning: as I build a model of you as a speaker, it should also affect my model of myself as a speaker.
 
-\subsection{A ``social anxiety'' model of pragmatic word learning}
+\subsection{A conventionality-based model of pragmatic word learning}
 
-Clearly we must deviate from pure normativity -- but how?
+We resolve the problems described above by assuming that speakers and listeners deviate from normative behavior by assuming a conventional lexicon. 
+% Clearly we must deviate from pure normativity -- but how?
 Here, we propose to resolve the above problems with a model of speakers and listeners as rational agents afflicted by a kind of social anxiety.
-Specifically, each agent assumes: (a) There is some single, specific literal lexicon which everyone should be using, (b) and everyone else knows this lexicon, and believes that I know it as well, (c) but in fact I don't.
-Assumption (a) explains the lexicographer's illusion: naive language users are convinced and will vociferously that words have specific meanings, even though these meanings are unobservable to everyone who purportedly uses them.
-It also explains why children speak the language they hear, since observing your behaviour now gives me data about not just \textit{your} lexicon, but \textit{the} lexicon.
-Assumption (b) avoids the explosion of hyper$^n$-distributions: if agent $n$ knows the lexicon, they assume that all lower agents do as well, reducing to the original tractable model without uncertainty.
+Each agent assumes: (a) There is some single, specific literal lexicon which everyone should be using, (b) and everyone else knows this lexicon, and believes that I know it as well, (c) but in fact I don't.
+These assumptions instantiate a kind of ``social anxiety'' in which agents are all trying to learn the correct lexicon that they assume everyone else knows.
+
+Assumption (a) corresponds to the lexicographer's illusion: Naive language users will argue vociferously that words have specific meanings, even though these meanings are unobservable to everyone who purportedly uses them.
+It also explains why learners speak the language they hear (rather than some private language that they assume listeners will eventually learn): Under assumption (a), observing other speakers' behaviour provides data about not just that speaker's idiosyncratic lexicon, but the consensus lexicon.
+Assumption (b) avoids the explosion of hyper$^n$-distributions described above: If agent $n$ knows the lexicon, they assume that all lower agents do as well, reducing to the original tractable model without uncertainty.
 And assumption (c) introduces a limited form of uncertainty, and thus the potential for learning.
-Everything else follows by applying standard Bayesian reasoning given these assumptions.
+% Everything else follows by applying standard Bayesian reasoning given these assumptions.
 
@@
 \end{align*}
-Here $L$'s data consists of $S$'s utterances together with whatever other information $L$ may have about their meaning (e.g.~from contextual clues) -- but by assumption (b), $L$ treats these as utterances as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus as being informative about the lexicon.
+Here $L$'s data consists of $S$'s utterances together with whatever other information $L$ may have about their meaning (e.g.~from contextual clues).
+By assumption (b), $L$ treats these as utterances as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus as being informative about the lexicon.
 Similarly, $S$ may observe $L$'s interpretations of various utterances, and will treat these as samples from $L_n$, not $L$.
+
 In several important cases, these assumptions are as close to rational as possible within the finite-recursion framework.
 If $L_n(\obj | \word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$ coincide, then  $S$ is fully normative and $L$ is off by only a single recursion.
-This in particular includes all previous experiments using this pragmatic reasoning framework, which considered only single-turn interactions where $\text{$L$'s data} = \text{$S$'s data} = \emptyset$.
+This case in particular includes all previous experiments using this pragmatic reasoning framework, which considered only single-turn interactions where $\text{$L$'s data} = \text{$S$'s data} = \emptyset$.
 Alternatively, if one agent has enough data that their uncertainty is effectively non-existent, then it becomes normative for their partner to treat them as being fully knowledgeable, and this is a reasonable description of the child language learning case.
 In general, though, in arbitrary multi-turn interactions in which both agents have non-trivial uncertainty, this model induces complex and non-normative learning dynamics.
-In the remainder of this paper, we computationally investigate the model's behaviour in all three situations.\footnote{In our simulations throughout, we somewhat arbitrarily set $n = 2$ (the minimal value that produces all the qualitative phenomena), $\lambda = 3$, and assume that all agents have shared priors on the lexicon and full knowledge of the cost function.
+In the remainder of this paper, we investigate the model's behaviour in all three situations.\footnote{In our simulations throughout, we somewhat arbitrarily set $n = 2$ (the minimal value that produces all the qualitative phenomena), $\lambda = 3$, and assume that all agents have shared priors on the lexicon and full knowledge of the cost function.
 Inference is via importance sampling.}
@@
 
-In this section, we describe the coverage of our pragmatic learning model on phenomena previously investigated with the pragmatics and word learning components of the model.
-To summarize these simulations: the pragmatic learning model reproduces previous findings and provides some new insights into the ``disambiguation'' phenomenon described above.
-
 \begin{table}[t]
@@
 \toprule
-Phenomenon & Ref. & WL & PI & PI+U & This work \\
+Phenomenon & Ref. & WL & PI & PI+U & PIL \\
 \midrule
@@
 \caption{Empirical results and references.
-WL refers to the word learning model of \cite{frank2009}; PI refers to the recursive pragmatic inference model of \cite{frank2012}; PI+U refers to the pragmatic inference model of \cite{bergen2012} which includes lexical uncertainty, marginalizes it out, and then recurses.}
+WL refers to the word learning model of \cite{frank2009}; PI refers to the recursive pragmatic inference model of \cite{frank2012}; PI+U refers to the pragmatic inference model of \cite{bergen2012} which includes lexical uncertainty, marginalizes it out, and then recurses.
+Our current model is referred to here as PIL, or pragmatic inference and learning.}
 \end{table}
+
+We begin by applying the model described above to a set of one-shot pragmatic inference games that have been well-studied in linguistics \cite{grice1975,horn1984} and are addressed by previous one-shot models of pragmatic inference \cite{frank2012,bergen2012}.
+These situations set the stage for simulations investigating how learning proceeds in iterated versions of such games, described in the following section.
+Results captured by our model and previous models are summarized in Table \ref{tab:results}.
+
+% In this section, we describe the coverage of our pragmatic learning model on phenomena previously investigated with the pragmatics and word learning components of the model.
+To summarize these simulations: the pragmatic learning model reproduces previous findings and provides some new insights into the ``disambiguation'' phenomenon described above.
 
@@
 \centering
-\includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
-\includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
-\includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
-\includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
-\caption{\label{fig:emergence} Simulations of two learning pragmatic agents playing a naming game.
-At each stage, $S$ and $L$ have different, possibly contradictory posteriors over the literal lexicon.
-From these posteriors we derive the probability $P(\text{$L$ understands $S$})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.}
+\includegraphics[width=0.98\textwidth]{figures/emergence-composite.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
+% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
+\caption{\label{fig:emergence} Simulations of two pragmatic agents playing a naming game.
+Each panel shows two representative simulation runs, with run 1 chosen to show strong convergence and run 2 chosen to show relatively weaker convergence.
+At each stage, $S$ and $L$ have different, possibly contradictory posteriors over the conventional, consensus lexicon.
+From these posteriors we derive the probability $P(\text{$L$ understands $S$})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.
+}
 \end{figure}

### 2013-05-31 praglearn.tex
text

--- 2013-05-31 praglearn.tex
@@
 Clearly we must deviate from pure normativity -- but how?
-Here, we propose to resolve the above problems with a model of speakers and listeners as rational agents afflicted by a kind of social anxiety.
+Here, we resolve the above problems by proposing a model of speakers and listeners as rational agents afflicted by a kind of social anxiety.
 Specifically, each agent assumes: (a) There is some single, specific literal lexicon which everyone should be using, (b) and everyone else knows this lexicon, and believes that I know it as well, (c) but in fact I don't.
@@
 
-\section{Results: Pragmatic inference}
+\section{Results: Pure pragmatic inference}
 
@@
 For example, with a uniform prior over lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$, $\text{cost}(\text{``cheap''}) = 0.5$, $\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the appropriate way: Despite total ignorance about the conventional system, speakers prefer to use simple words for common referents ($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88, P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) = 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) = 0.65$).
-This preference is somewhat weak, at least to start -- see \S\ref{sec:horn-emergence}
+This preference is somewhat weak; the critical observation is that it exists at all, given the unbiased priors.
+We return to this point in \S\ref{sec:horn-emergence}.
 
@@
 
-\section{Results: Learning in pragmatic contexts}
+\section{Results: The role of pragmatics in child language learning}
 
@@
 75\%) that the speaker is referring to the novel object.
-Nevertheless, integrating across lexicons reveals that this probability is not due to learning that the novel word refers to this object (as shown in Fig.
-\ref{fig:me}, left): instead, it comes in part from a specificity implicature because the listener knows that the familiar word \emph{does not} refer to the novel object--hence the novel word (even if it refers to the familiar object as well) is the best way to refer to the novel object.
+Nevertheless, this is not accompanied by an increased belief that the novel word literally refers to this object (as shown in Fig.
+\ref{fig:me}, left).
+This comes in part from a potential specificity implicature: the listener knows that the familiar word \emph{does not} refer to the novel object--hence the novel word (even if it refers to the familiar object as well) is the best way to refer to the novel object.
 Nevertheless, on repeated exposure to the same novel word, novel object situation, the model does lexicalize this inference (Fig.
@@
 
-% XX maybe make the sections ``Pure pragmatics'' (scalar and horn implicature), ``Pure learning'' (cross-situational), ``Learning and pragmatics'' (subsection 1: disambiguation and learning scalar implicature, subsection 2: emergence with and without horn implicature)
+\subsection{Lexicalization and preservation of specificity implicature}
+\label{sec:learning-specificity-implic}
 
-We now turn to phenomena which have not been modeled by previous pragmatic or learning models, as they seem to intrinsically require both together.
+\begin{SCfigure}[1]
+\centering
+\includegraphics[width=1.5in]{figures/some-all-only-pragmatic.pdf}
+\includegraphics[width=1.5in]{figures/some-all-pragmatic+unambiguous.pdf}
+\caption{\label{fig:scalar} Simulations..}
+\end{SCfigure}
 
-\subsection{Emergence of efficient communicative conventions}
+XX on figure, add x label: ``some'' means: ALL   SOME   SBNA, y label: ``all'' means: ALL  SOME  SBNA
+
+The acquisition of scalar adjectives like ``some'' provides a puzzle for most models of word learning: given that in many contexts, the word ``some'' is used to mean \textsc{some but not all}, how do children learn that this is not in fact its literal meaning?
+(XX cite snedeker?) As suggested above, though, our model is able to take specificity implicatures into account when learning, and thus potentially answer this puzzle.
+Fig.~\ref{fig:scalar} (left) shows the model's beliefs after being shown several examples of ``some'' meaning \textsc{some but not all} and ``all'' meaning \textsc{all}: it believes that at least one of these meanings must be literal, but it isn't convinced that both are -- it's possible ``some'' means \textsc{some}, or that \textsc{all} means \textsc{all or some}.
+Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' meaning \textsc{some but not all}, but in a context where the meaning \textsc{some but not all} was essentially known from context (has a high prior).
+
+\subsection{Mutual learning}
+
+\subsubsection{Emergence of efficient communicative conventions}
 
@@
 
-\subsection{Lexicalization and loss of Horn implicatures}
+\subsubsection{Lexicalization and loss of Horn implicatures}
 \label{sec:horn-emergence}
@@
 Similarly, the socially anxious speaker will wager that since on average they will succed more often if the listener is using the good system, they might as well try it -- and when they succeed, they take this as evidence that the listener was in fact using the good system all along.
-As a result, dyads in this game not only end up converging onto a stable lexicon, but onto specifically the good lexicon, at a rate far above chance.
+As a result, dyads in this game not only end up converging onto a stable system, but onto specifically the good system, at a rate far above chance.
+However, something fundamentally different happens here than in the specificity implicature case (\S\ref{sec:learning-specificity-implic}).
+There, we saw that our model could learn hear pragmatically strengthened usages of words while recovering the underlying literal meaning.
+In this case, though, the pragmatic strengthening depends essentially on ignorance about literal meaning.
+As the agents gather more data, therefore, the implicature disappears, to be replaced by a belief that ``cheap'' \textit{literally} means \textsc{common}.
+To demonstrate this, we queried each agent in each dyad about how they would refer to or interpret each object and word, \textit{if} the two objects were equally common, which cancels the Horn implicature.
+As shown in Fig.~\ref{emergence-average} (right), after 30 turns, in nearly 70\% of dyads both $S$ and $L$ believed in the ``good'' mapping even in this implicature-free case, while less than 20\% believed in the ``bad'' mapping (with the rest being inconsistent).
 
@@
 
-\subsection{Lexicalization and preservation of scalar implicature}
-
-\begin{figure}
-\centering
-\includegraphics[width=2in]{figures/some-all-only-pragmatic.pdf}
-\includegraphics[width=2in]{figures/some-all-pragmatic+unambiguous.pdf}
-\caption{\label{fig:scalar} Simulations..}
-\end{figure}
-
 \section{Conclusion}

### 2013-05-31 figures/scalar-implic.py

@@ -30,8 +30,7 @@ some_for_partial_alone = conpact2.LDataUttWithMeaning(LISTENER_DEPTH - 1, 0, 0, 
 
 def show_scalar_posterior(path, listener):
     show_lex_posterior(path, listener.weighted_lexicons(), (0, 0), (1, 0),
-                       xlabel="P(Some means partial, not full)",
-                       ylabel="P(All means partial, not full)")
+                       xlabel="", ylabel="")
 
 # Only pragmatically strengthened examples:
 

### 2013-05-31 praglearn.tex

--- 2013-05-31 praglearn.tex
@@
 % Clearly we must deviate from pure normativity -- but how?
-Here, we propose to resolve the above problems with a model of speakers and listeners as rational agents afflicted by a kind of social anxiety.
+Here, we resolve the above problems by proposing a model of speakers and listeners as rational agents afflicted by a kind of social anxiety.
 Each agent assumes: (a) There is some single, specific literal lexicon which everyone should be using, (b) and everyone else knows this lexicon, and believes that I know it as well, (c) but in fact I don't.

### 2013-05-31 praglearn.tex
more figures
fix sectioning
more text

--- 2013-05-31 praglearn.tex
@@
 
-\subsection{Statistical cross-situational word learning}
+% \subsection{Statistical cross-situational word learning}
 
-Even very young children can learn words from individually ambiguous situations, e.g.
+% Even very young children can learn words from individually ambiguous situations, e.g.
 where multiple words are spoken and multiple objects are present \cite{smith2008}.
@@
 
-\begin{figure}[t]
+\begin{SCfigure}
 \centering
-\includegraphics[width=0.48\textwidth]{figures/horn-emergence-0.pdf}
-\includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
+\includegraphics[width=0.5\textwidth]{figures/horn-composite.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
 \caption{\label{fig:horn} Simulations.}
-\end{figure}
+\end{SCfigure}
 

### 2013-05-31 praglearn.tex
more text edits

@@
 
-\section{Results: The role of pragmatics in child language learning}
+\section{Results: The role of pragmatics in language acquisition}
 
@@
 
-XX on figure, add x label: ``some'' means: ALL   SOME   SBNA, y label: ``all'' means: ALL  SOME  SBNA
+% XX on figure, add x label: ``some'' means: ALL   SOME   SBNA, y label: ``all'' means: ALL  SOME  SBNA
 
 The acquisition of scalar adjectives like ``some'' provides a puzzle for most models of word learning: given that in many contexts, the word ``some'' is used to mean \textsc{some but not all}, how do children learn that this is not in fact its literal meaning?
-(XX cite snedeker?) As suggested above, though, our model is able to take specificity implicatures into account when learning, and thus potentially answer this puzzle.
+Our model is able to take specificity implicatures into account when learning, and thus potentially answer this puzzle.
 Fig.~\ref{fig:scalar} (left) shows the model's beliefs after being shown several examples of ``some'' meaning \textsc{some but not all} and ``all'' meaning \textsc{all}: it believes that at least one of these meanings must be literal, but it isn't convinced that both are -- it's possible ``some'' means \textsc{some}, or that \textsc{all} means \textsc{all or some}.
@@
 
-\subsection{Mutual learning}
+\section{Results: Mutual learning}
 
-\subsubsection{Emergence of efficient communicative conventions}
+\subsection{Emergence of efficient communicative conventions}
 
-If two humans are forced to play a game which depends on communication, but are allowed to communicate only via a set of novel symbols which have no pre-existing conventional meaning, then over time they will generally derive a set of new conventions which allow them to accomplish their task \cite{galantucci2005}.
-Or if, less drastically, a task requires two people to repeatedly refer to novel objects, they will implicitly through the course of the interaction invent conventional names for them (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}).
+Experimental results suggest that communicators are able to establish novel, consensus-based communication systems.
+For example, adults playing a communication game using only novel symbols with no conventional meaning will typically converge on a set of new conventions which allow them to accomplish their task \cite{galantucci2005}.
+In a second experiment, communicators asked to refer to novel objects through the course of the interaction invent conventional names for them (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}).
 From a pure learning perspective this is anomalous: since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour.
@@
 
-\begin{figure}[t]
+\begin{SCfigure}
 \centering
 \includegraphics[width=0.3\textwidth]{figures/emergence-average.pdf}
-\hspace{0.5in}
+\hspace{0.2in}
 \includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
-\caption{\label{fig:emergence-average} Behaviour of learning agents playing naming games as in Figs.~\ref{fig:emergence} and \ref{fig:horn}, averaged across 300 dialogues.
-Left: Average communicative success by game type and dialogue turn.
-Right: The proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have not only converged on using the ``good'' or ``bad'' lexicons, but believe that these are the literal meanings of the words.}
-\end{figure}
+\caption{\label{fig:emergence-average} Averaged behaviour over 300 dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}.
+Left: Communicative success by game type and dialogue turn.
+Right: Proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or ``bad'' lexicons and believe that these are literal meanings.}
+\end{SCfigure}
 
@@
 
-\subsubsection{Lexicalization and loss of Horn implicatures}
+\subsection{Lexicalization and loss of Horn implicatures}
 \label{sec:horn-emergence}
@@
 % \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
-\caption{\label{fig:horn} Simulations.}
+\caption{\label{fig:horn} Dialogue.}
 \end{SCfigure}

### 2013-05-31 praglearn.tex
text

--- 2013-05-31 praglearn.tex
@@
 (XX cite snedeker?) As suggested above, though, our model is able to take specificity implicatures into account when learning, and thus potentially answer this puzzle.
-Fig.~\ref{fig:scalar} (left) shows the model's beliefs after being shown several examples of ``some'' meaning \textsc{some but not all} and ``all'' meaning \textsc{all}: it believes that at least one of these meanings must be literal, but it isn't convinced that both are -- it's possible ``some'' means \textsc{some}, or that \textsc{all} means \textsc{all or some}.
-Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' meaning \textsc{some but not all}, but in a context where the meaning \textsc{some but not all} was essentially known from context (has a high prior).
+Imagine two bowls: one containing a mixture of apples and oranges, the other containing apples alone.
+Fig.~\ref{fig:scalar} (left) shows the model's beliefs after encountering the words ``some'' and ``all'' used contrastively to refer to these two bowls (as in e.g.\ ``the bowl with some apples'').
+It believes that at least one of these words must have a restrictive literal meaning, but not both -- it leaves open the possibility that either ``some'' or ``all'' might be literally ambiguous.
+Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' referring to exactly the same object, but in contexts where this is the only object available, so there is no contrast and thus no implicature; this convinces the model that ``all'' literally refers only to the bowl that contains only apples, while ``some'' most likely is somewhat ambiguous.
 
-\subsection{Mutual learning}
+\section{Mutual learning}
 
-\subsubsection{Emergence of efficient communicative conventions}
+\subsection{Emergence of efficient communicative conventions}
 

### 2013-05-31 figures/non-emergence.py

@@ -2,12 +2,10 @@ from nips import *
 
 from scipy.special import gammaln
 
-# speaker and listener each have dirichlet priors on a set of object -> word
+# two agents each have dirichlet priors on a set of object -> word
 # multinomial distributions.
-# speaker to produce a word samples from dirichlet-multinomial
-# listener conditions on the word to guess the meaning
-# speaker increments this (word, interpretation) pair in its Dirichlet prior
-# listener increments (word, target) pair in its Dirichlet prior
+# without pragmatics we can't condition on listener interpretation, so
+# instead, they take turns speaking
 
 SIZE = 2
 

### 2013-05-31 praglearn.tex

--- 2013-05-31 praglearn.tex
@@
 Our model is able to take specificity implicatures into account when learning, and thus potentially answer this puzzle.
-Fig.~\ref{fig:scalar} (left) shows the model's beliefs after being shown several examples of ``some'' meaning \textsc{some but not all} and ``all'' meaning \textsc{all}: it believes that at least one of these meanings must be literal, but it isn't convinced that both are -- it's possible ``some'' means \textsc{some}, or that \textsc{all} means \textsc{all or some}.
-Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' meaning \textsc{some but not all}, but in a context where the meaning \textsc{some but not all} was essentially known from context (has a high prior).
+Imagine two bowls: one containing a mixture of apples and oranges, the other containing apples alone.
+Fig.~\ref{fig:scalar} (left) shows the model's beliefs after encountering the words ``some'' and ``all'' used contrastively to refer to these two bowls (as in e.g.\ ``the bowl with some apples'').
+It believes that at least one of these words must have a restrictive literal meaning, but not both -- it leaves open the possibility that either ``some'' or ``all'' might be literally ambiguous.
+Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' referring to exactly the same object, but in contexts where this is the only object available, so there is no contrast and thus no implicature; this convinces the model that ``all'' literally refers only to the bowl that contains only apples, while ``some'' most likely is somewhat ambiguous.
 

### 2013-05-31 praglearn.tex
simulation text edits

--- 2013-05-31 praglearn.tex
@@
 For example, adults playing a communication game using only novel symbols with no conventional meaning will typically converge on a set of new conventions which allow them to accomplish their task \cite{galantucci2005}.
-In a second experiment, communicators asked to refer to novel objects through the course of the interaction invent conventional names for them (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}).
-From a pure learning perspective this is anomalous: since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour.
-Furthermore, even if only one partner is producing the novel expressions, their behaviour still becomes more regular (conventional) over time, which would seem to rule out a role for learning -- even if there is some pattern in the expressions the speaker chooses to use, there is certainly nothing for the \textit{speaker} to learn by observing these patterns, and thus their behaviour should not change over time.
-An obvious hack might be to simply stipulate that speakers take their own utterances as training data for further learning, but this would fail to explain yet another wrinkle in this phenomenon: speakers only regularize their production over time if they have some kind of feedback indicating that the listener understands; if such feedback is removed, the phenomenon is dramatically reduced (XX krauss 1966, branigan catchpole pickering 2011).
+Or in a less extreme example, communicators asked to refer to novel objects invent conventional names for them over the course of repeated interactions (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}).
+From a pure learning perspective this is anomalous, however: Since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour.
+Furthermore, even if only one partner is producing the novel expressions, their behaviour still becomes more regular (conventional) over time, which would seem to rule out a role for learning---even if there is some pattern in the expressions the speaker chooses to use, there is certainly nothing for the \textit{speaker} to learn by observing these patterns, and thus their behaviour should not change over time.
+% An obvious hack might be to simply stipulate that speakers take their own utterances as training data for further learning, but this would fail to explain yet another wrinkle in this phenomenon: speakers only regularize their production over time if they have some kind of feedback indicating that the listener understands; if such feedback is removed, the phenomenon is dramatically reduced (XX krauss 1966, branigan catchpole pickering 2011).
 
@@
 
-\begin{SCfigure}
-\centering
-\includegraphics[width=0.3\textwidth]{figures/emergence-average.pdf}
-\hspace{0.2in}
-\includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
-\caption{\label{fig:emergence-average} Averaged behaviour over 300 dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}.
-Left: Communicative success by game type and dialogue turn.
-Right: Proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or ``bad'' lexicons and believe that these are literal meanings.}
-\end{SCfigure}
+To model such phenomena, we imagine two agents playing the simple referential game introduced above.
+On each turn the speaker is assigned a target object, utters some word referring to this object, the listener makes a guess at the object, and then, critically, the speaker observes the listener's guess and the listener receives feedback indicating the correct answer (i.e., the speaker's intended meaning).
+The listener then updates their posterior over lexicons before proceeding to the next trial.
+As in \cite{krauss1964,clark1986}, the speaker and listener remain fixed in the same role throughout.
 
-To model such phenomena, we imagine two agents playing the simple referential game introduced above, in which on each turn the speaker is assigned a target object, they utter some word referring to this object, the listener makes a guess at the object, and then, critically, the speaker observes the listener's guess and the listener receives feedback indicating the correct answer (i.e., the speaker's intended meaning), which they use to update their posterior over lexicons before proceeding to the next trial.
-As in \cite{krauss1964,clark1986}, the speaker and listener remain fixed in the same role throughout.
 Fig.~\ref{fig:emergence} shows the result of simulating several such games when both parties begin in total ignorance.
-Notice that: (a) agents' performance begins at chance, but quickly rises; (b) they tend towards structured, sparse lexicons with a one-to-one correspondence between objects and words; (c) as the speaker and listener have entirely different data (the listener's interpretations and the speaker's intended meaning, respectively), unlucky early guesses can lead them to believe in entirely contradictory lexicons -- but they generally recover and converge despite this.
-
-What's happening here is that each agent is, effectively, using their partner's behaviour as a basis for forming superstitions about the underlying lexicon that they assume must exist.
-Since they then each act on these superstitions, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a superstition becomes normatively correct.
-
-What happens if we remove feedback from the listener to the speaker?
-If the speaker is not allowed to observe the listener's guess, then the speaker does no learning and their posterior over lexicons remains perfectly uniform.
+Notice that: (a) agents' performance begins at chance, but quickly rises; (b) they tend towards structured, sparse lexicons with a one-to-one correspondence between objects and words; and (c) as the speaker and listener have entirely different data (the listener's interpretations and the speaker's intended meaning, respectively), unlucky early guesses can lead them to believe in entirely contradictory lexicons---but they generally recover and converge.
+Each agent effectively uses their partner's behavior as a basis for forming weak beliefs about the underlying lexicon that they assume must exist.
+Since they then each act on these beliefs, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a ``superstition'' becomes normatively correct.
+%On the other hand, if the speaker is not allowed to observe the listener's guess, then the speaker does no learning and their posterior over lexicons remains perfectly uniform.
 They therefore choose words by simply flipping a coin, and performance remains at chance.
 
-The role of learning here is obvious, but what does the pragmatics part of the model do?
-First, it makes it possible for the speaker to learn from listener interpretations, which is not possible without an explicit model of the speaker as a rational agent.
-Second, it creates the sparse structure in these observed lexicons: at each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful, since if the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well.
-But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon.
+Pragmatic inference here does two things.
+First, the pragmatic recursion makes it possible for the speaker to learn from the listener's interpretations, which would be impossible without an explicit model of the listener as a rational agent.
+Second, pragmatics creates the sparse structure in the observed lexicons: At each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful.
+If the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well.
+
+% But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon.
 The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.
@@
 % \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
-\caption{\label{fig:horn} Dialogue.}
+\caption{\label{fig:horn} Example simulations showing the lexicalization of horn implicatures. Plotting conventions are as above.}
 \end{SCfigure}
 
+\begin{SCfigure}
+\centering
+\includegraphics[width=0.3\textwidth]{figures/emergence-average.pdf}
+\hspace{0.2in}
+\includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
+\caption{\label{fig:emergence-average} Averaged behaviour over 300 dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}.
+Left: Communicative success by game type and dialogue turn.
+Right: Proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or ``bad'' lexicons and believe that these are literal meanings.}
+\end{SCfigure}
 A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}.

### 2013-05-31 praglearn.tex
text

--- 2013-05-31 praglearn.tex
@@
 It believes that at least one of these words must have a restrictive literal meaning, but not both -- it leaves open the possibility that either ``some'' or ``all'' might be literally ambiguous.
-Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' referring to exactly the same object, but in contexts where this is the only object available, so there is no contrast and thus no implicature; this convinces the model that ``all'' literally refers only to the bowl that contains only apples, while ``some'' most likely is somewhat ambiguous.
+Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' referring to exactly the same object, but in contexts where this is the only object available, so there is no contrast and thus no implicature; this convinces the model that ``all apple'' literally refers only to the bowl that contains no oranges, while ``some'' most likely is somewhat ambiguous.
+Crucially, the model is able to learn this without once ever observing ``some'' used
 

### 2013-05-31 praglearn.tex
reflow all the paragraphs!

@@
 
-% Tell LaTeX to break pages in the middle of displayed math whenever this would be the slightest bit useful
+% Tell LaTeX to break pages in the middle of displayed math whenever
+% this would be the slightest bit useful
 \allowdisplaybreaks[4]
@@
 \begin{abstract}
-Language users are remarkably good at making inferences about speakers' intentions in context, and children learning their native language also display substantial pragmatic skill in acquiring the meanings of unknown words.
-While pragmatic inference and word learning have both been characterized in probabilistic terms, no current models make appropriate inferences about the meanings of words in pragmatically-rich inferential contexts.
-We describe a model in which language learners assume that they jointly approximate a shared, external lexicon and reason recursively about the goals of others in using this lexicon.
-This model captures phenomena in word learning and pragmatic inference; it additionally leads to insights about the emergence of communicative systems in conversation and the mechanisms by which pragmatic inferences become incorporated into word meanings.
+  Language users are remarkably good at making inferences about
+  speakers' intentions in context, and children learning their native
+  language also display substantial pragmatic skill in acquiring the
+  meanings of unknown words. While pragmatic inference and word
+  learning have both been characterized in probabilistic terms, no
+  current models make appropriate inferences about the meanings of
+  words in pragmatically-rich inferential contexts. We describe a
+  model in which language learners assume that they jointly
+  approximate a shared, external lexicon and reason recursively about
+  the goals of others in using this lexicon. This model captures
+  phenomena in word learning and pragmatic inference; it additionally
+  leads to insights about the emergence of communicative systems in
+  conversation and the mechanisms by which pragmatic inferences become
+  incorporated into word meanings.
 \end{abstract}
@@
 
-Natural language is a powerful tool for conveying meaning in context.
-Our utterances are pervasively underspecified.
-Consider ``it's raining,'' ``I ate some of the cookies,'' or ``can you close the window?'' In each, a listener must go beyond the literal meaning of the words to fill in contextual details (``it's raining here and now''), infer that a stronger alternative is not true (``I ate some but not all of the cookies''), or more generally infer the speaker's communicative goal (``I want you to close the window right now because I'm cold.'') -- a process known as \textit{pragmatic reasoning}.
+Natural language is a powerful tool for conveying meaning in
+context. Our utterances are pervasively underspecified. Consider
+``it's raining,'' ``I ate some of the cookies,'' or ``can you close
+the window?'' In each, a listener must go beyond the literal meaning
+of the words to fill in contextual details (``it's raining here and
+now''), infer that a stronger alternative is not true (``I ate some
+but not all of the cookies''), or more generally infer the speaker's
+communicative goal (``I want you to close the window right now because
+I'm cold.'') -- a process known as \textit{pragmatic reasoning}.
 
-Following this intuition, theories of pragmatic communication frame the process of language comprehension as inference about the generating goal of an utterance given a rational speaker \cite{grice1975,dale1995,frank2012}.
-For example, a listener might reason, ``if she had wanted me to think `all' of the cookies, she would have said `all'---but she didn't.
-Hence `all' must not be true and she must have eaten some {\it but not all} of the cookies.'' This kind of counterfactual reasoning has a recursive character to it: listener models speaker (who may in turn be modeling a listener).
+Following this intuition, theories of pragmatic communication frame
+the process of language comprehension as inference about the
+generating goal of an utterance given a rational speaker
+\cite{grice1975,dale1995,frank2012}. For example, a listener might
+reason, ``if she had wanted me to think `all' of the cookies, she
+would have said `all'---but she didn't. Hence `all' must not be true
+and she must have eaten some {\it but not all} of the cookies.'' This
+kind of counterfactual reasoning has a recursive character to it:
+listener models speaker (who may in turn be modeling a listener).
 
-Even before they are proficient language users, children learn the meanings of words from their pattern of use in context.
-For example, in demonstrations of the ``disambiguation effect,'' a novel word (e.g.
-``dax'') is used in a context containing both a novel and a familiar object.
-From quite young, children tend to make the inference that ``dax'' refers to the novel object \cite{markman1988}.
-In another demonstration, even young toddlers can extract mappings between words and objects from a pattern of individually-ambiguous co-occurrences between pairs of words and pairs of objects \cite{smith2008}.
-Taking inspiration from children's learning prowess, recent work in the field of grounded language learning has explored the possibility that even relatively complex word meanings can be inferred from corpus data (e.g.
-\cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).
+Even before they are proficient language users, children learn the
+meanings of words from their pattern of use in context. For example,
+in demonstrations of the ``disambiguation effect,'' a novel word
+(e.g. ``dax'') is used in a context containing both a novel and a
+familiar object. From quite young, children tend to make the inference
+that ``dax'' refers to the novel object \cite{markman1988}. In another
+demonstration, even young toddlers can extract mappings between words
+and objects from a pattern of individually-ambiguous co-occurrences
+between pairs of words and pairs of objects \cite{smith2008}. Taking
+inspiration from children's learning prowess, recent work in the field
+of grounded language learning has explored the possibility that even
+relatively complex word meanings can be inferred from corpus data
+(e.g. \cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).
 
-Recursive pragmatic inferences such as those described above could be an important part of language learning.\footnote{Very young children make inferences that are often labeled as ``pragmatic'' in that they involve reasoning about context \cite{clark1988,baldwin1993}, though they also show weaknesses in specific inferences (e.g.
-strengthening ``some'' to ``some but not all'' \cite{papafragou2003}).
-Here we remain agnostic about the age at which children are able to make such inferences robustly, as it may vary depending on the linguistic materials being used in the inference \cite{barner2011}.} For example, the disambiguation inference described above can emerge from such reasoning (``why would the speaker have said ``dax'' if there was a word for the known object that she knew I would know'') \cite{clark1988}.
-In addition, there are a variety of intriguing cases in which listeners seem to create situation- and task-specific ways of referring to particular objects.
-For example, when asked to refer to idiosyncratic geometric shapes, over the course of an experimental session, participants create conventionalized descriptions that allow them to perform accurately even though they do not begin with a shared label \cite{krauss1964,clark1986}.
-Yet models of grounded language learning have failed to take pragmatic factors into account.
+Recursive pragmatic inferences such as those described above could be
+an important part of language learning.\footnote{Very young children
+  make inferences that are often labeled as ``pragmatic'' in that they
+  involve reasoning about context \cite{clark1988,baldwin1993}, though
+  they also show weaknesses in specific inferences (e.g. strengthening
+  ``some'' to ``some but not all'' \cite{papafragou2003}). Here we
+  remain agnostic about the age at which children are able to make
+  such inferences robustly, as it may vary depending on the linguistic
+  materials being used in the inference \cite{barner2011}.} For
+example, the disambiguation inference described above can emerge from
+such reasoning (``why would the speaker have said ``dax'' if there was
+a word for the known object that she knew I would know'')
+\cite{clark1988}. In addition, there are a variety of intriguing cases
+in which listeners seem to create situation- and task-specific ways of
+referring to particular objects. For example, when asked to refer to
+idiosyncratic geometric shapes, over the course of an experimental
+session, participants create conventionalized descriptions that allow
+them to perform accurately even though they do not begin with a shared
+label \cite{krauss1964,clark1986}. Yet models of grounded language
+learning have failed to take pragmatic factors into account.
 
-The goal of our current work is to investigate the possibilities for integrating models of recursive pragmatic reasoning with models of language learning, with the hope of capturing phenomena in both domains.
-We begin by describing previous models in each domain as well as proposals for bringing the two together.
-We next simulate findings from the domains of pragmatic inference, word learning, and their intersection.
-We end by discussing the implications of our results for work on language change.
+The goal of our current work is to investigate the possibilities for
+integrating models of recursive pragmatic reasoning with models of
+language learning, with the hope of capturing phenomena in both
+domains. We begin by describing previous models in each domain as well
+as proposals for bringing the two together. We next simulate findings
+from the domains of pragmatic inference, word learning, and their
+intersection. We end by discussing the implications of our results for
+work on language change.
 
@@
 
-We model a standard communication game \cite{krauss1964,clark1986}: two participants each,
-separately, view identical arrays of objects.
-On the \textit{Speaker's} screen, one object is highlighted; their goal is to get the \textit{Listener} to click on this item.
-To do this, they have available a fixed, finite set of words; they must pick one.
-The Listener then receives this word, and attempts to guess which object
-the Speaker meant by it.
-In the psychology literature, as in real-world interactions, games are typically iterated; one view of our contribution here is as a generalization of one-shot models \cite{frank2012,bergen2012} to the iterated context.
+We model a standard communication game \cite{krauss1964,clark1986}:
+two participants each, separately, view identical arrays of
+objects. On the \textit{Speaker's} screen, one object is highlighted;
+their goal is to get the \textit{Listener} to click on this item. To
+do this, they have available a fixed, finite set of words; they must
+pick one. The Listener then receives this word, and attempts to guess
+which object the Speaker meant by it. In the psychology literature, as
+in real-world interactions, games are typically iterated; one view of
+our contribution here is as a generalization of one-shot models
+\cite{frank2012,bergen2012} to the iterated context.
 
 \subsection{Paradoxes in optimal models of pragmatic learning}
-% When modelling human behaviour, it's useful to begin with optimal models; this has the advantage that when we are then forced to deviate from the purely rational approach, it's generally in a theoretically interpretable way.
-However,
-Multi-agent interactions are difficult to model in a normative or optimal framework without falling prey to paradox.
-For example, consider a simple model of the agents in the above game.
-First we define a \textit{literal listener} $L_0$.
-This agent has a \textit{lexicon} which probabilistically maps words to meanings; specifically, it assigns each word $w$ a vector describing the extent to which this word provides evidence for each object\footnote{We assume words refer directly to objects, rather than to abstract semantic features.
-Our simplification is without loss of generalization, however, because we can interpret our model as marginalizing over such a representation, with our literal $P_\lex(\obj | \word) = \sum_\text{features} P(\obj | \text{features}) P_\lex(\text{features} | \word)$.}.
-Each entry in the vector is in $(0, 1)$, and together they sum to 1:
+% When modelling human behaviour, it's useful to begin with optimal
+% models; this has the advantage that when we are then forced to
+% deviate from the purely rational approach, it's generally in a
+% theoretically interpretable way. However,
+Multi-agent interactions are difficult to model in a normative or
+optimal framework without falling prey to paradox. For example,
+consider a simple model of the agents in the above game. First we
+define a \textit{literal listener} $L_0$. This agent has a
+\textit{lexicon} which probabilistically maps words to meanings;
+specifically, it assigns each word $w$ a vector describing the extent
+to which this word provides evidence for each object\footnote{We
+  assume words refer directly to objects, rather than to abstract
+  semantic features. Our simplification is without loss of
+  generalization, however, because we can interpret our model as
+  marginalizing over such a representation, with our literal
+  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
+  \text{features}) P_\lex(\text{features} | \word)$.}. Each entry in
+the vector is in $(0, 1)$, and together they sum to 1:
 \begin{align*}
@@
 \end{align*}
-Confronted with such a listener, a probabilistically rational speaker should attempt to choose a word which soft-maximizes the probability that the listener will assign to the target object---modulated by the effort or cost associated with producing this word:
+Confronted with such a listener, a probabilistically rational speaker
+should attempt to choose a word which soft-maximizes the probability
+that the listener will assign to the target object---modulated by the
+effort or cost associated with producing this word:
 \begin{align*}
@@
 \end{align*}
-But given this speaker, then the naive $L_0$ strategy is not optimal. Instead, listeners should use Bayes rule to invert the speaker's decision procedure \cite{frank2012}:
+But given this speaker, then the naive $L_0$ strategy is not
+optimal. Instead, listeners should use Bayes rule to invert the
+speaker's decision procedure \cite{frank2012}:
 \begin{align*}
@@
 \end{align*}
-Now the paradox becomes apparent.
-Given such a listener, it is no longer optimal for speakers to implement strategy $S_1$; instead, they should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead of $P_{L_0}$.
-And then listeners ought to implement $L_4$, and so on.
+Now the paradox becomes apparent. Given such a listener, it is no
+longer optimal for speakers to implement strategy $S_1$; instead, they
+should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
+of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.
 
-One computationally expensive option is to continue iterating such strategies until reaching a fixed point equilibrium.
-While this strategy guarantees that each agent will behave normatively given the other agent's strategy, there is no guarantee that such strategies will be near the system's global optimum; equilibrium strategies make up only a small subspace of all possible strategies.
-Nor is there much evidence the humans find such equilibria; their behaviour in language games (and in other games \cite{camerer2004}) can be well-modeled as implementing $S_n$ or $L_n$ for some small $n$ \cite{frank2012}.
-Following this work, we  recurse a finite (small) number of times.
-The consequence is that one agent, implementing $S_n$, is fully optimal with respect to the game, while the other, implementing $L_{n-1}$, is nearly optimal---off by only a single recursion.
+One computationally expensive option is to continue iterating such
+strategies until reaching a fixed point equilibrium. While this
+strategy guarantees that each agent will behave normatively given the
+other agent's strategy, there is no guarantee that such strategies
+will be near the system's global optimum; equilibrium strategies make
+up only a small subspace of all possible strategies. Nor is there much
+evidence the humans find such equilibria; their behaviour in language
+games (and in other games \cite{camerer2004}) can be well-modeled as
+implementing $S_n$ or $L_n$ for some small $n$
+\cite{frank2012}. Following this work, we recurse a finite (small)
+number of times. The consequence is that one agent, implementing
+$S_n$, is fully optimal with respect to the game, while the other,
+implementing $L_{n-1}$, is nearly optimal---off by only a single
+recursion.
 
-As soon as we attempt to add uncertainty about the meanings of words to such a model, new paradoxes arise.
-Suppose the listener is a young child who is uncertain about the lexicon their partner is using.
-The obvious solution is for them to place a prior on the lexicon; they then update their posterior based on whatever utterances and contextual cues they observe, and in the mean time interpret each utterance by making their best guess, marginalizing out this uncertainty.
-This basic structure is captured in previous models of Bayesian word learning \cite{frank2009}.
-But when combined with the recursive pragmatic model, a new question arises: Given such a listener, what model should the speaker use?
-A rational speaker attempts to maximize the listener's likelihood of understanding, so if an uncertain listener interpets by marginalizing over some posterior, then a knowledgeable speaker should calculate this posterior and marginalize themselves.
-But then the speaker is not using any particular lexicon at all, and a rational learning listener should therefore not be trying to infer the speaker's lexicon.\footnote{Of course, in reality both parties will generally have some uncertainty, making the situation even worse.
-If we start from an uncertain listener with a prior over lexicons, then a first-level uncertain speaker needs a prior over priors on lexicons, a second-level uncertain listener needs a prior over priors over priors, etc.
-The original $L_0 \rightarrow S_1 \rightarrow \dots$ recursion was bad enough, but at least each step had a constant cost.
-This new recursion produces hyper$^n$-distributions for which inference almost immediately becomes intractable even in principle, since the dimensionality of the learning problem increases with each step.
-Yet, without this addition of new uncertainty at each level, the model would dissolve back into certainty as in the previous paragraph, making learning impossible.}% In fact, such an uncertain listener's uncertainty dissolves; they may not know the lexicon, but they do know their posterior over the lexicon -- and because this is all that the speaker cares about, there is nothing for the listener to learn from the speaker's behaviour.
+As soon as we attempt to add uncertainty about the meanings of words
+to such a model, new paradoxes arise. Suppose the listener is a young
+child who is uncertain about the lexicon their partner is using. The
+obvious solution is for them to place a prior on the lexicon; they
+then update their posterior based on whatever utterances and
+contextual cues they observe, and in the mean time interpret each
+utterance by making their best guess, marginalizing out this
+uncertainty. This basic structure is captured in previous models of
+Bayesian word learning \cite{frank2009}. But when combined with the
+recursive pragmatic model, a new question arises: Given such a
+listener, what model should the speaker use? A rational speaker
+attempts to maximize the listener's likelihood of understanding, so if
+an uncertain listener interpets by marginalizing over some posterior,
+then a knowledgeable speaker should calculate this posterior and
+marginalize themselves. But then the speaker is not using any
+particular lexicon at all, and a rational learning listener should
+therefore not be trying to infer the speaker's lexicon.\footnote{Of
+  course, in reality both parties will generally have some
+  uncertainty, making the situation even worse. If we start from an
+  uncertain listener with a prior over lexicons, then a first-level
+  uncertain speaker needs a prior over priors on lexicons, a
+  second-level uncertain listener needs a prior over priors over
+  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
+  recursion was bad enough, but at least each step had a constant
+  cost. This new recursion produces hyper$^n$-distributions for which
+  inference almost immediately becomes intractable even in principle,
+  since the dimensionality of the learning problem increases with each
+  step. Yet, without this addition of new uncertainty at each level,
+  the model would dissolve back into certainty as in the previous
+  paragraph, making learning
+  impossible.}
+% In fact, such an uncertain listener's uncertainty dissolves; they
+% may not know the lexicon, but they do
+% know their posterior over the lexicon -- and because
+% this is all that the speaker cares about, there is
+% nothing for the listener to learn from the speaker's
+% behaviour.
 
-To add insult to injury, one final problem is that there is nothing constraining speaker and listener to use the same language under this model.
-In principle, so long as the listener has a good model of the speaker, and the speaker has a good model of the listener, then communication succeeds.
-This result clearly presents a dramatic mismatch to typical human communication, which almost never proceeds with distinct languages spoken by each participant.
+To add insult to injury, one final problem is that there is nothing
+constraining speaker and listener to use the same language under this
+model. In principle, so long as the listener has a good model of the
+speaker, and the speaker has a good model of the listener, then
+communication succeeds. This result clearly presents a dramatic
+mismatch to typical human communication, which almost never proceeds
+with distinct languages spoken by each participant.
 
-% .
-Yet most people do in fact speak languages that they have learned by listening to others, which suggests an additional constraint that should somehow be incorporated into learning: as I build a model of you as a speaker, it should also affect my model of myself as a speaker.
+% . Yet most people do in fact speak languages that they have learned
+% by listening to others, which suggests an additional constraint that
+% should somehow be incorporated into learning: as I build a model of
+% you as a speaker, it should also affect my model of myself as a
+% speaker.
 
@@
 
-We resolve the problems described above by assuming that speakers and listeners deviate from normative behavior by assuming a conventional lexicon. 
-% Clearly we must deviate from pure normativity -- but how?
-Here, we resolve the above problems by proposing a model of speakers and listeners as rational agents afflicted by a kind of social anxiety.
-Each agent assumes: (a) There is some single, specific literal lexicon which everyone should be using, (b) and everyone else knows this lexicon, and believes that I know it as well, (c) but in fact I don't.
-These assumptions instantiate a kind of ``social anxiety'' in which agents are all trying to learn the correct lexicon that they assume everyone else knows.
+We resolve the problems described above by assuming that speakers and
+listeners deviate from normative behavior by assuming a conventional
+lexicon.
+% Clearly we must deviate from pure normativity -- but how? Here, we
+% resolve the above problems by proposing a model of speakers and
+% listeners as rational agents afflicted by a kind of social anxiety.
+Each agent assumes: (a) There is some single, specific literal lexicon
+which everyone should be using, (b) and everyone else knows this
+lexicon, and believes that I know it as well, (c) but in fact I
+don't. These assumptions instantiate a kind of ``social anxiety'' in
+which agents are all trying to learn the correct lexicon that they
+assume everyone else knows.
 
-Assumption (a) corresponds to the lexicographer's illusion: Naive language users will argue vociferously that words have specific meanings, even though these meanings are unobservable to everyone who purportedly uses them.
-It also explains why learners speak the language they hear (rather than some private language that they assume listeners will eventually learn): Under assumption (a), observing other speakers' behaviour provides data about not just that speaker's idiosyncratic lexicon, but the consensus lexicon.
-Assumption (b) avoids the explosion of hyper$^n$-distributions described above: If agent $n$ knows the lexicon, they assume that all lower agents do as well, reducing to the original tractable model without uncertainty.
-And assumption (c) introduces a limited form of uncertainty, and thus the potential for learning.
-% Everything else follows by applying standard Bayesian reasoning given these assumptions.
+Assumption (a) corresponds to the lexicographer's illusion: Naive
+language users will argue vociferously that words have specific
+meanings, even though these meanings are unobservable to everyone who
+purportedly uses them. It also explains why learners speak the
+language they hear (rather than some private language that they assume
+listeners will eventually learn): Under assumption (a), observing
+other speakers' behaviour provides data about not just that speaker's
+idiosyncratic lexicon, but the consensus lexicon. Assumption (b)
+avoids the explosion of hyper$^n$-distributions described above: If
+agent $n$ knows the lexicon, they assume that all lower agents do as
+well, reducing to the original tractable model without
+uncertainty. And assumption (c) introduces a limited form of
+uncertainty, and thus the potential for learning.
+% Everything else follows by applying standard Bayesian reasoning
+% given these assumptions.
 
-Formally, let an unadorned $L$ and $S$ denote our socially-anxious listener and speaker.
-$L$ inverts their marginal distribution over the speaker; $S$ soft-maximizes their marginal distribution over the listener.
-With $L_i$ and $S_i$ as above, this gives:
+Formally, let an unadorned $L$ and $S$ denote our socially-anxious
+listener and speaker. $L$ inverts their marginal distribution over the
+speaker; $S$ soft-maximizes their marginal distribution over the
+listener. With $L_i$ and $S_i$ as above, this gives:
 \begin{align*}
@@
 \end{align*}
-Here $L$'s data consists of $S$'s utterances together with whatever other information $L$ may have about their meaning (e.g.~from contextual clues).
-By assumption (b), $L$ treats these as utterances as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus as being informative about the lexicon.
-Similarly, $S$ may observe $L$'s interpretations of various utterances, and will treat these as samples from $L_n$, not $L$.
+Here $L$'s data consists of $S$'s utterances together with whatever
+other information $L$ may have about their meaning (e.g.~from
+contextual clues). By assumption (b), $L$ treats these as utterances
+as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
+as being informative about the lexicon. Similarly, $S$ may observe
+$L$'s interpretations of various utterances, and will treat these as
+samples from $L_n$, not $L$.
 
-In several important cases, these assumptions are as close to rational as possible within the finite-recursion framework.
-If $L_n(\obj | \word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$ coincide, then  $S$ is fully normative and $L$ is off by only a single recursion.
-This case in particular includes all previous experiments using this pragmatic reasoning framework, which considered only single-turn interactions where $\text{$L$'s data} = \text{$S$'s data} = \emptyset$.
-Alternatively, if one agent has enough data that their uncertainty is effectively non-existent, then it becomes normative for their partner to treat them as being fully knowledgeable, and this is a reasonable description of the child language learning case.
-In general, though, in arbitrary multi-turn interactions in which both agents have non-trivial uncertainty, this model induces complex and non-normative learning dynamics.
-In the remainder of this paper, we investigate the model's behaviour in all three situations.\footnote{In our simulations throughout, we somewhat arbitrarily set $n = 2$ (the minimal value that produces all the qualitative phenomena), $\lambda = 3$, and assume that all agents have shared priors on the lexicon and full knowledge of the cost function.
-Inference is via importance sampling.}
+In several important cases, these assumptions are as close to rational
+as possible within the finite-recursion framework. If $L_n(\obj |
+\word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$
+coincide, then $S$ is fully normative and $L$ is off by only a single
+recursion. This case in particular includes all previous experiments
+using this pragmatic reasoning framework, which considered only
+single-turn interactions where $\text{$L$'s data} = \text{$S$'s data}
+= \emptyset$. Alternatively, if one agent has enough data that their
+uncertainty is effectively non-existent, then it becomes normative for
+their partner to treat them as being fully knowledgeable, and this is
+a reasonable description of the child language learning case. In
+general, though, in arbitrary multi-turn interactions in which both
+agents have non-trivial uncertainty, this model induces complex and
+non-normative learning dynamics. In the remainder of this paper, we
+investigate the model's behaviour in all three situations.\footnote{In
+  our simulations throughout, we somewhat arbitrarily set $n = 2$ (the
+  minimal value that produces all the qualitative phenomena), $\lambda
+  = 3$, and assume that all agents have shared priors on the lexicon
+  and full knowledge of the cost function. Inference is via importance
+  sampling.}
 
@@
 \end{center}
-\caption{Empirical results and references.
-WL refers to the word learning model of \cite{frank2009}; PI refers to the recursive pragmatic inference model of \cite{frank2012}; PI+U refers to the pragmatic inference model of \cite{bergen2012} which includes lexical uncertainty, marginalizes it out, and then recurses.
-Our current model is referred to here as PIL, or pragmatic inference and learning.}
+\caption{Empirical results and references. WL refers to the word
+  learning model of \cite{frank2009}; PI refers to the recursive
+  pragmatic inference model of \cite{frank2012}; PI+U refers to the
+  pragmatic inference model of \cite{bergen2012} which includes
+  lexical uncertainty, marginalizes it out, and then recurses. Our
+  current model is referred to here as PIL, or pragmatic inference and
+  learning.}
 \end{table}
 
-We begin by applying the model described above to a set of one-shot pragmatic inference games that have been well-studied in linguistics \cite{grice1975,horn1984} and are addressed by previous one-shot models of pragmatic inference \cite{frank2012,bergen2012}.
-These situations set the stage for simulations investigating how learning proceeds in iterated versions of such games, described in the following section.
-Results captured by our model and previous models are summarized in Table \ref{tab:results}.
+We begin by applying the model described above to a set of one-shot
+pragmatic inference games that have been well-studied in linguistics
+\cite{grice1975,horn1984} and are addressed by previous one-shot
+models of pragmatic inference \cite{frank2012,bergen2012}. These
+situations set the stage for simulations investigating how learning
+proceeds in iterated versions of such games, described in the
+following section. Results captured by our model and previous models
+are summarized in Table \ref{tab:results}.
 
-% In this section, we describe the coverage of our pragmatic learning model on phenomena previously investigated with the pragmatics and word learning components of the model.
-To summarize these simulations: the pragmatic learning model reproduces previous findings and provides some new insights into the ``disambiguation'' phenomenon described above.
+% In this section, we describe the coverage of our pragmatic learning
+% model on phenomena previously investigated with the pragmatics and
+% word learning components of the model. To summarize these
+% simulations: the pragmatic learning model reproduces previous
+% findings and provides some new insights into the ``disambiguation''
+% phenomenon described above.
 
@@
 
-Many sets of words in natural language form scales in which each term refers to successively stronger restrictions on the situations compatible with the term.
-``Some'' and ``all'' form a scale of this type.
-While ``I ate some of the cookies'' is compatible with the followup ``in fact, I ate {\it all} of the cookies,'' the reverse is not true.
-``Might'' and ``must'' are another example, as are ``OK,'' ``good,'' and ``excellent.'' All of these scales allow for {\it scalar implicatures} \cite{grice1975}: if a speaker uses a weaker term, a listener is licensed to infer that the stronger term was not true in this situation.
-So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs.
-We use the term {\it specificity implicature} here (following \cite{bergen2012}) to denote that we are agnostic about whether such implicatures are conventionalized or whether they use a contextually-defined scale.
+Many sets of words in natural language form scales in which each term
+refers to successively stronger restrictions on the situations
+compatible with the term. ``Some'' and ``all'' form a scale of this
+type. While ``I ate some of the cookies'' is compatible with the
+followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
+not true. ``Might'' and ``must'' are another example, as are ``OK,''
+``good,'' and ``excellent.'' All of these scales allow for {\it scalar
+  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
+listener is licensed to infer that the stronger term was not true in
+this situation. So although ``I ate some of the cookies'' could in
+principle be compatible with eating all of them, the listener is lead
+to believe that ``some but not all'' is the likely state of
+affairs. We use the term {\it specificity implicature} here (following
+\cite{bergen2012}) to denote that we are agnostic about whether such
+implicatures are conventionalized or whether they use a
+contextually-defined scale.
 
-The recursive pragmatic reasoning portions of our model capture findings on specificity implicature. Consider a representation of scalar implicature as a matrix 
-$\left[
+The recursive pragmatic reasoning portions of our model capture
+findings on specificity implicature. Consider a representation of
+scalar implicature as a matrix $\left[
     \begin{array}{cc}
@@
     \end{array}
-  \right]$,
-where rows show the words ``some'' and ``all'' and columns indicate worlds where {\sc some but not all} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case.
-Assuming uniform costs for the two words, a rational speaker confronted with such a lexicon will prefer to say ``some'' only when they mean \textsc{some but not all}, and this is therefore the interpretation that a rational listener will place on ``some'' as well.
-To simulate this situation, we place a Dirichlet(10, 10) prior on the meaning of \textit{some} and a Dirichlet(1, 10) prior on the meaning of \textit{all}, and find this gives speaker use probabilities of
+  \right]$, where rows show the words ``some'' and ``all'' and columns
+  indicate worlds where {\sc some but not all} and {\sc all} are true,
+  respectively, capturing the regularity that ``some'' can be said of
+  situations where {\sc all} is actually the case. Assuming uniform
+  costs for the two words, a rational speaker confronted with such a
+  lexicon will prefer to say ``some'' only when they mean \textsc{some
+    but not all}, and this is therefore the interpretation that a
+  rational listener will place on ``some'' as well. To simulate this
+  situation, we place a Dirichlet(10, 10) prior on the meaning of
+  \textit{some} and a Dirichlet(1, 10) prior on the meaning of
+  \textit{all}, and find this gives speaker use probabilities of
   $\left[
@@
 
-% XX note the relation to BGL model; they need uncertainty in the middle of the pragmatics specifically, which is where we end up putting it for other reasons.
-And in Bayesian framework uncertainty and learning must always coincide.
-Also the intuition that this is a general property of uncertainty for soft-max agents -- if you don't know where you are, good worlds have more peaked utility distributions than bad worlds, and thus more consistent behaviour by other agents, so on average when you're guessing what other agents will do, good worlds contribute more to your average than bad worlds.
+% XX note the relation to BGL model; they need uncertainty in the
+% middle of the pragmatics specifically, which is where we end up
+% putting it for other reasons. And in Bayesian framework uncertainty
+% and learning must always coincide. Also the intuition that this is a
+% general property of uncertainty for soft-max agents -- if you don't
+% know where you are, good worlds have more peaked utility
+% distributions than bad worlds, and thus more consistent behaviour by
+% other agents, so on average when you're guessing what other agents
+% will do, good worlds contribute more to your average than bad
+% worlds.
 
-Consider a world in which contains two words and two types of objects.
-One word is expensive to use, and one is cheap (call them ``expensive'' and ``cheap'' for short).
-One object type is common and one is rare; denote these \textsc{common} and \textsc{rare}.
-Intuitively, there are two possible communicative systems here: a good system where ``cheap'' means \textsc{common} and ``expensive'' means \textsc{rare}, and a bad system where the opposite holds.
-Obviously we would prefer to use the good system, but it has historically proven very difficult to derive this conclusion in a game theoretic setting, because both systems are stable equilibria; if our partner uses the bad system, then we would rather follow and communicate at some cost than switch to the good system and fail entirely \cite{bergen2012}.
+Consider a world in which contains two words and two types of
+objects. One word is expensive to use, and one is cheap (call them
+``expensive'' and ``cheap'' for short). One object type is common and
+one is rare; denote these \textsc{common} and
+\textsc{rare}. Intuitively, there are two possible communicative
+systems here: a good system where ``cheap'' means \textsc{common} and
+``expensive'' means \textsc{rare}, and a bad system where the opposite
+holds. Obviously we would prefer to use the good system, but it has
+historically proven very difficult to derive this conclusion in a game
+theoretic setting, because both systems are stable equilibria; if our
+partner uses the bad system, then we would rather follow and
+communicate at some cost than switch to the good system and fail
+entirely \cite{bergen2012}.
 
-Humans, however, unlike traditional game theoretic models, do make the inference that if two utterances seem equivalent except that one is more costly to utter, then the costly utterance refers to an unusual event.
-We call this pattern \textit{Horn implicature}, after \cite{horn1984}.
-For instance, ``Lee got the car to stop'' implies that Lee used an unusual method (e.g.~not the brakes) because, had he used the brakes, the speaker would have chosen the simpler and shorter (less costly) expression, ``Lee stopped the car'' \cite{horn1984}.
-Surprisingly, Bergen et al \cite{bergen2012} show that the key to achieving this favorable result is ignorance.
-If a listener assigns equal probability to her partner using the good system or the bad system, then their best bet is to estimate $P_S(\word | \obj)$ as the average of $P_S(\word | \obj, \text{good system})$ and $P_S(\word | \obj, \text{bad system})$.
-These might seem to cancel out, but in fact they do not.
-In the good system, the speaker has higher expected utility than the bad system; therefore, a soft-max agent in the bad system has noiser behaviour than in the good system, and their behaviour in the good system dominates the average, breaking the symmetry.
-Similar reasoning applies to an uncertain speaker.
+Humans, however, unlike traditional game theoretic models, do make the
+inference that if two utterances seem equivalent except that one is
+more costly to utter, then the costly utterance refers to an unusual
+event. We call this pattern \textit{Horn implicature}, after
+\cite{horn1984}. For instance, ``Lee got the car to stop'' implies
+that Lee used an unusual method (e.g.~not the brakes) because, had he
+used the brakes, the speaker would have chosen the simpler and shorter
+(less costly) expression, ``Lee stopped the car''
+\cite{horn1984}. Surprisingly, Bergen et al \cite{bergen2012} show
+that the key to achieving this favorable result is ignorance. If a
+listener assigns equal probability to her partner using the good
+system or the bad system, then their best bet is to estimate
+$P_S(\word | \obj)$ as the average of $P_S(\word | \obj, \text{good
+  system})$ and $P_S(\word | \obj, \text{bad system})$. These might
+seem to cancel out, but in fact they do not. In the good system, the
+speaker has higher expected utility than the bad system; therefore, a
+soft-max agent in the bad system has noiser behaviour than in the good
+system, and their behaviour in the good system dominates the average,
+breaking the symmetry. Similar reasoning applies to an uncertain
+speaker.
 
-% You know, this is almost exactly like a quantitatively graded version of Pascal's wager.
+% You know, this is almost exactly like a quantitatively graded
+% version of Pascal's wager.
 
@@
 
-The model of \cite{bergen2012} differs from ours in that they then apply further layers of pragmatic recursion on top of these marginal distributions, which as described above removes the possibility of learning; in their model it is only hypothetical agents who are ignorant, while the actual speaker and listener know everything there is to know about each other's behaviour.
-However, the core result carries over to our model.
-For example, with a uniform prior over lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$, $\text{cost}(\text{``cheap''}) = 0.5$, $\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the appropriate way: Despite total ignorance about the conventional system, speakers prefer to use simple words for common referents ($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88, P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) = 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) = 0.65$).
-This preference is somewhat weak; the critical observation is that it exists at all, given the unbiased priors.
-We return to this point in \S\ref{sec:horn-emergence}.
+The model of \cite{bergen2012} differs from ours in that they then
+apply further layers of pragmatic recursion on top of these marginal
+distributions, which as described above removes the possibility of
+learning; in their model it is only hypothetical agents who are
+ignorant, while the actual speaker and listener know everything there
+is to know about each other's behaviour. However, the core result
+carries over to our model. For example, with a uniform prior over
+lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$,
+$\text{cost}(\text{``cheap''}) = 0.5$,
+$\text{cost}(\text{``expensive''}) = 1.0$, the symmetry breaks in the
+appropriate way: Despite total ignorance about the conventional
+system, speakers prefer to use simple words for common referents
+($P_S(\text{``cheap''} | \text{\textsc{common}}) = 0.88,
+P_S(\text{``cheap''} | \text{\textsc{rare}}) = 0.46$), and listeners
+show a similar bias ($P_L(\text{\textsc{common}} | \text{``cheap''}) =
+0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) =
+0.65$). This preference is somewhat weak; the critical observation is
+that it exists at all, given the unbiased priors. We return to this
+point in \S\ref{sec:horn-emergence}.
 
@@
 
-% Even very young children can learn words from individually ambiguous situations, e.g.
-where multiple words are spoken and multiple objects are present \cite{smith2008}.
-While no individual situation gives away the mapping, the pattern of co-occurrences between words and objects allows for learning.
-Our current model reproduces this pattern of results, converging to correct lexical mappings from ambiguous evidence.
+% Even very young children can learn words from individually ambiguous
+% situations, e.g. where multiple words are spoken and multiple
+% objects are present \cite{smith2008}. While no individual situation
+% gives away the mapping, the pattern of co-occurrences between words
+% and objects allows for learning. Our current model reproduces this
+% pattern of results, converging to correct lexical mappings from
+% ambiguous evidence.
 
@@
 
-Children, when presented with both a novel and a familiar object (e.g.
-an eggbeater and a ball), will treat a novel label (e.g.
-``dax'') as referring to the novel object, for example by supplying the eggbeater when asked to ``give me the dax'' \cite{markman1988}.
-This phenomenon is sometimes referred to as ``mutual exclusivity'' or even ``Fast mapping,'' and many distinct theoretical accounts have been proposed in the psychological literature.
-Simple probabilistic word learning models can produce a similar pattern of findings \cite{frank2009}, often due to the kinds of ``explaining away'' type phenomena that are observed in such models.
-But all such models assume that learners in fact retain the mapping between word and object that is demonstrated in the experimental situation; this observation is contradicted, however, by evidence that children often do not retain the mappings that are demonstrated by their inferences in the moment \cite{horst2008}.
+Children, when presented with both a novel and a familiar object
+(e.g. an eggbeater and a ball), will treat a novel label
+(e.g. ``dax'') as referring to the novel object, for example by
+supplying the eggbeater when asked to ``give me the dax''
+\cite{markman1988}. This phenomenon is sometimes referred to as
+``mutual exclusivity'' or even ``Fast mapping,'' and many distinct
+theoretical accounts have been proposed in the psychological
+literature. Simple probabilistic word learning models can produce a
+similar pattern of findings \cite{frank2009}, often due to the kinds
+of ``explaining away'' type phenomena that are observed in such
+models. But all such models assume that learners in fact retain the
+mapping between word and object that is demonstrated in the
+experimental situation; this observation is contradicted, however, by
+evidence that children often do not retain the mappings that are
+demonstrated by their inferences in the moment \cite{horst2008}.
 
-Our model provides an intriguing possible explanation of this finding: given a single disambiguation situation, the model gives a substantial probability (e.g.
-75\%) that the speaker is referring to the novel object.
-Nevertheless, this is not accompanied by an increased belief that the novel word literally refers to this object (as shown in Fig.
-\ref{fig:me}, left).
-This comes in part from a potential specificity implicature: the listener knows that the familiar word \emph{does not} refer to the novel object--hence the novel word (even if it refers to the familiar object as well) is the best way to refer to the novel object.
-Nevertheless, on repeated exposure to the same novel word, novel object situation, the model does lexicalize this inference (Fig.
-\ref{fig:me}, right).
+Our model provides an intriguing possible explanation of this finding:
+given a single disambiguation situation, the model gives a substantial
+probability (e.g. 75\%) that the speaker is referring to the novel
+object. Nevertheless, this is not accompanied by an increased belief
+that the novel word literally refers to this object (as shown in
+Fig. \ref{fig:me}, left). This comes in part from a potential
+specificity implicature: the listener knows that the familiar word
+\emph{does not} refer to the novel object--hence the novel word (even
+if it refers to the familiar object as well) is the best way to refer
+to the novel object. Nevertheless, on repeated exposure to the same
+novel word, novel object situation, the model does lexicalize this
+inference (Fig. \ref{fig:me}, right).
 
@@
 
-The acquisition of scalar adjectives like ``some'' provides a puzzle for most models of word learning: given that in many contexts, the word ``some'' is used to mean \textsc{some but not all}, how do children learn that this is not in fact its literal meaning?
-Our model is able to take specificity implicatures into account when learning, and thus potentially answer this puzzle.
-Imagine two bowls: one containing a mixture of apples and oranges, the other containing apples alone.
-Fig.~\ref{fig:scalar} (left) shows the model's beliefs after encountering the words ``some'' and ``all'' used contrastively to refer to these two bowls (as in e.g.\ ``the bowl with some apples'').
-It believes that at least one of these words must have a restrictive literal meaning, but not both -- it leaves open the possibility that either ``some'' or ``all'' might be literally ambiguous.
-Fig.~\ref{fig:scalar} (right) then shows the model's beliefs if we also provide examples of ``some'' referring to exactly the same object, but in contexts where this is the only object available, so there is no contrast and thus no implicature; this convinces the model that ``all apple'' literally refers only to the bowl that contains no oranges, while ``some'' most likely is somewhat ambiguous.
-Crucially, the model is able to learn this without once ever observing ``some'' used
+The acquisition of scalar adjectives like ``some'' provides a puzzle
+for most models of word learning: given that in many contexts, the
+word ``some'' is used to mean \textsc{some but not all}, how do
+children learn that this is not in fact its literal meaning? Our model
+is able to take specificity implicatures into account when learning,
+and thus potentially answer this puzzle. Imagine two bowls: one
+containing a mixture of apples and oranges, the other containing
+apples alone. Fig.~\ref{fig:scalar} (left) shows the model's beliefs
+after encountering the words ``some'' and ``all'' used contrastively
+to refer to these two bowls (as in e.g.\ ``the bowl with some
+apples''). It believes that at least one of these words must have a
+restrictive literal meaning, but not both -- it leaves open the
+possibility that either ``some'' or ``all'' might be literally
+ambiguous. Fig.~\ref{fig:scalar} (right) then shows the model's
+beliefs if we also provide examples of ``some'' referring to exactly
+the same object, but in contexts where this is the only object
+available, so there is no contrast and thus no implicature; this
+convinces the model that ``all apple'' literally refers only to the
+bowl that contains no oranges, while ``some'' most likely is somewhat
+ambiguous. Crucially, the model is able to learn this without once
+ever observing ``some'' used
 
@@
 
-Experimental results suggest that communicators are able to establish novel, consensus-based communication systems.
-For example, adults playing a communication game using only novel symbols with no conventional meaning will typically converge on a set of new conventions which allow them to accomplish their task \cite{galantucci2005}.
-Or in a less extreme example, communicators asked to refer to novel objects invent conventional names for them over the course of repeated interactions (e.g., ``the ice skater'' for an abstract figure vaguely resembling an ice skater, \cite{clark1986}).
-From a pure learning perspective this is anomalous, however: Since both agents know perfectly well that there is no underlying convention to discover, there is nothing to learn from the other's behaviour.
-Furthermore, even if only one partner is producing the novel expressions, their behaviour still becomes more regular (conventional) over time, which would seem to rule out a role for learning---even if there is some pattern in the expressions the speaker chooses to use, there is certainly nothing for the \textit{speaker} to learn by observing these patterns, and thus their behaviour should not change over time.
-% An obvious hack might be to simply stipulate that speakers take their own utterances as training data for further learning, but this would fail to explain yet another wrinkle in this phenomenon: speakers only regularize their production over time if they have some kind of feedback indicating that the listener understands; if such feedback is removed, the phenomenon is dramatically reduced (XX krauss 1966, branigan catchpole pickering 2011).
+Experimental results suggest that communicators are able to establish
+novel, consensus-based communication systems. For example, adults
+playing a communication game using only novel symbols with no
+conventional meaning will typically converge on a set of new
+conventions which allow them to accomplish their task
+\cite{galantucci2005}. Or in a less extreme example, communicators
+asked to refer to novel objects invent conventional names for them
+over the course of repeated interactions (e.g., ``the ice skater'' for
+an abstract figure vaguely resembling an ice skater,
+\cite{clark1986}). From a pure learning perspective this is anomalous,
+however: Since both agents know perfectly well that there is no
+underlying convention to discover, there is nothing to learn from the
+other's behaviour. Furthermore, even if only one partner is producing
+the novel expressions, their behaviour still becomes more regular
+(conventional) over time, which would seem to rule out a role for
+learning---even if there is some pattern in the expressions the
+speaker chooses to use, there is certainly nothing for the
+\textit{speaker} to learn by observing these patterns, and thus their
+behaviour should not change over time.
+% An obvious hack might be to simply stipulate that speakers take
+% their own utterances as training data for further learning, but this
+% would fail to explain yet another wrinkle in this phenomenon:
+% speakers only regularize their production over time if they have
+% some kind of feedback indicating that the listener understands; if
+% such feedback is removed, the phenomenon is dramatically reduced (XX
+% krauss 1966, branigan catchpole pickering 2011).
 
@@
 % \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
-\caption{\label{fig:emergence} Simulations of two pragmatic agents playing a naming game.
-Each panel shows two representative simulation runs, with run 1 chosen to show strong convergence and run 2 chosen to show relatively weaker convergence.
-At each stage, $S$ and $L$ have different, possibly contradictory posteriors over the conventional, consensus lexicon.
-From these posteriors we derive the probability $P(\text{$L$ understands $S$})$ (marginalizing over target objects and word choices), and also depict graphically $S$'s model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$, and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s data})$.
-}
+\caption{\label{fig:emergence} Simulations of two pragmatic agents
+  playing a naming game. Each panel shows two representative
+  simulation runs, with run 1 chosen to show strong convergence and
+  run 2 chosen to show relatively weaker convergence. At each stage,
+  $S$ and $L$ have different, possibly contradictory posteriors over
+  the conventional, consensus lexicon. From these posteriors we derive
+  the probability $P(\text{$L$ understands $S$})$ (marginalizing over
+  target objects and word choices), and also depict graphically $S$'s
+  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$,
+  and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s
+    data})$. }
 \end{figure}
 
-To model such phenomena, we imagine two agents playing the simple referential game introduced above.
-On each turn the speaker is assigned a target object, utters some word referring to this object, the listener makes a guess at the object, and then, critically, the speaker observes the listener's guess and the listener receives feedback indicating the correct answer (i.e., the speaker's intended meaning).
-The listener then updates their posterior over lexicons before proceeding to the next trial.
-As in \cite{krauss1964,clark1986}, the speaker and listener remain fixed in the same role throughout.
+To model such phenomena, we imagine two agents playing the simple
+referential game introduced above. On each turn the speaker is
+assigned a target object, utters some word referring to this object,
+the listener makes a guess at the object, and then, critically, the
+speaker observes the listener's guess and the listener receives
+feedback indicating the correct answer (i.e., the speaker's intended
+meaning). The listener then updates their posterior over lexicons
+before proceeding to the next trial. As in
+\cite{krauss1964,clark1986}, the speaker and listener remain fixed in
+the same role throughout.
 
-Fig.~\ref{fig:emergence} shows the result of simulating several such games when both parties begin in total ignorance.
-Notice that: (a) agents' performance begins at chance, but quickly rises; (b) they tend towards structured, sparse lexicons with a one-to-one correspondence between objects and words; and (c) as the speaker and listener have entirely different data (the listener's interpretations and the speaker's intended meaning, respectively), unlucky early guesses can lead them to believe in entirely contradictory lexicons---but they generally recover and converge.
-Each agent effectively uses their partner's behavior as a basis for forming weak beliefs about the underlying lexicon that they assume must exist.
-Since they then each act on these beliefs, and their partner uses the resulting actions to form new superstitions, they soon converge on using similar lexicons, and what started as a ``superstition'' becomes normatively correct.
-%On the other hand, if the speaker is not allowed to observe the listener's guess, then the speaker does no learning and their posterior over lexicons remains perfectly uniform.
-They therefore choose words by simply flipping a coin, and performance remains at chance.
+Fig.~\ref{fig:emergence} shows the result of simulating several such
+games when both parties begin in total ignorance. Notice that: (a)
+agents' performance begins at chance, but quickly rises; (b) they tend
+towards structured, sparse lexicons with a one-to-one correspondence
+between objects and words; and (c) as the speaker and listener have
+entirely different data (the listener's interpretations and the
+speaker's intended meaning, respectively), unlucky early guesses can
+lead them to believe in entirely contradictory lexicons---but they
+generally recover and converge. Each agent effectively uses their
+partner's behavior as a basis for forming weak beliefs about the
+underlying lexicon that they assume must exist. Since they then each
+act on these beliefs, and their partner uses the resulting actions to
+form new superstitions, they soon converge on using similar lexicons,
+and what started as a ``superstition'' becomes normatively correct.
+% On the other hand, if the speaker is not allowed to observe the
+% listener's guess, then the speaker does no learning and their
+% posterior over lexicons remains perfectly uniform. They therefore
+% choose words by simply flipping a coin, and performance remains at
+% chance.
 
-Pragmatic inference here does two things.
-First, the pragmatic recursion makes it possible for the speaker to learn from the listener's interpretations, which would be impossible without an explicit model of the listener as a rational agent.
-Second, pragmatics creates the sparse structure in the observed lexicons: At each step, the goal-directed speaker attempts to exploit whatever communicatively useful asymmetries exist in its current model of the listener, and these attempts tend to be successful.
-If the listener has even a small asymmetry in their current posterior then their pragmatic reasoning will magnify it as well.
+Pragmatic inference here does two things. First, the pragmatic
+recursion makes it possible for the speaker to learn from the
+listener's interpretations, which would be impossible without an
+explicit model of the listener as a rational agent. Second, pragmatics
+creates the sparse structure in the observed lexicons: At each step,
+the goal-directed speaker attempts to exploit whatever communicatively
+useful asymmetries exist in its current model of the listener, and
+these attempts tend to be successful. If the listener has even a small
+asymmetry in their current posterior then their pragmatic reasoning
+will magnify it as well.
 
-% But over time, the speaker's use of such patterns, and the listener's ability to understand them, will each be taken by the other as evidence that these patterns actually exist in the literal lexicon.
-The result is that while both agents start out with uniform priors over the lexicon, they are biased to end up believing in lexicons that are communicatively useful for the task at hand.
+% But over time, the speaker's use of such patterns, and the
+% listener's ability to understand them, will each be taken by the
+% other as evidence that these patterns actually exist in the literal
+% lexicon. The result is that while both agents start out with uniform
+% priors over the lexicon, they are biased to end up believing in
+% lexicons that are communicatively useful for the task at hand.
 
-% XXXX what can we say about the role pragmatics plays here?
-It's annoyingly non-trivial to sample from the equivalent model where the speaker just samples from a multinomial distribution and the listener inverts it.
-The listener's data is samples from a DM so they have conjugacy.
-But the speaker's data is listener interpretations, which necessarily do one tiny pragmatic inference step from using Bayes rule, which breaks conjugacy...
+% XXXX what can we say about the role pragmatics plays here? It's
+% annoyingly non-trivial to sample from the equivalent model where the
+% speaker just samples from a multinomial distribution and the
+% listener inverts it. The listener's data is samples from a DM so
+% they have conjugacy. But the speaker's data is listener
+% interpretations, which necessarily do one tiny pragmatic inference
+% step from using Bayes rule, which breaks conjugacy...
 
@@
 \includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
-\caption{\label{fig:emergence-average} Averaged behaviour over 300 dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}.
-Left: Communicative success by game type and dialogue turn.
-Right: Proportion of dyads in the Horn implicature game (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or ``bad'' lexicons and believe that these are literal meanings.}
+\caption{\label{fig:emergence-average} Averaged behaviour over 300
+  dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}. Left:
+  Communicative success by game type and dialogue turn. Right:
+  Proportion of dyads in the Horn implicature game
+  (\S\ref{sec:horn-emergence}) who have converged on the ``good'' or
+  ``bad'' lexicons and believe that these are literal meanings.}
 \end{SCfigure}
-A stronger example of how pragmatics can create such biases can be observed if we simulate agents playing this naming game in the ``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of \S\ref{sec:horn-implicature}.
-Recall that in this domain, pragmatic reasoning causes each agent to start out independently biased -- but only weakly -- towards the associations ``cheap'' $\leftrightarrow$ \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}, and that this bias arises exactly because their prior on literal meaning is uniform and uninformative.
-A fully rational listener, therefore, observing our uncertain speaker using this mapping, would discount it as arising from this bias, and conclude that the speaker in fact did not know which convention was in use.
-Our socially anxious listener, however, believes that the speaker does know which convention is in use, and therefore tends to take their behaviour as evidence that the good system is in fact in use.
-Similarly, the socially anxious speaker will wager that since on average they will succed more often if the listener is using the good system, they might as well try it -- and when they succeed, they take this as evidence that the listener was in fact using the good system all along.
-As a result, dyads in this game not only end up converging onto a stable system, but onto specifically the good system, at a rate far above chance.
-However, something fundamentally different happens here than in the specificity implicature case (\S\ref{sec:learning-specificity-implic}).
-There, we saw that our model could learn hear pragmatically strengthened usages of words while recovering the underlying literal meaning.
-In this case, though, the pragmatic strengthening depends essentially on ignorance about literal meaning.
-As the agents gather more data, therefore, the implicature disappears, to be replaced by a belief that ``cheap'' \textit{literally} means \textsc{common}.
-To demonstrate this, we queried each agent in each dyad about how they would refer to or interpret each object and word, \textit{if} the two objects were equally common, which cancels the Horn implicature.
-As shown in Fig.~\ref{emergence-average} (right), after 30 turns, in nearly 70\% of dyads both $S$ and $L$ believed in the ``good'' mapping even in this implicature-free case, while less than 20\% believed in the ``bad'' mapping (with the rest being inconsistent).
+A stronger example of how pragmatics can create such biases can be
+observed if we simulate agents playing this naming game in the
+``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of
+\S\ref{sec:horn-implicature}. Recall that in this domain, pragmatic
+reasoning causes each agent to start out independently biased -- but
+only weakly -- towards the associations ``cheap'' $\leftrightarrow$
+\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}, and
+that this bias arises exactly because their prior on literal meaning
+is uniform and uninformative. A fully rational listener, therefore,
+observing our uncertain speaker using this mapping, would discount it
+as arising from this bias, and conclude that the speaker in fact did
+not know which convention was in use. Our socially anxious listener,
+however, believes that the speaker does know which convention is in
+use, and therefore tends to take their behaviour as evidence that the
+good system is in fact in use. Similarly, the socially anxious speaker
+will wager that since on average they will succed more often if the
+listener is using the good system, they might as well try it -- and
+when they succeed, they take this as evidence that the listener was in
+fact using the good system all along. As a result, dyads in this game
+not only end up converging onto a stable system, but onto specifically
+the good system, at a rate far above chance. However, something
+fundamentally different happens here than in the specificity
+implicature case (\S\ref{sec:learning-specificity-implic}). There, we
+saw that our model could learn hear pragmatically strengthened usages
+of words while recovering the underlying literal meaning. In this
+case, though, the pragmatic strengthening depends essentially on
+ignorance about literal meaning. As the agents gather more data,
+therefore, the implicature disappears, to be replaced by a belief that
+``cheap'' \textit{literally} means \textsc{common}. To demonstrate
+this, we queried each agent in each dyad about how they would refer to
+or interpret each object and word, \textit{if} the two objects were
+equally common, which cancels the Horn implicature. As shown in
+Fig.~\ref{emergence-average} (right), after 30 turns, in nearly 70\%
+of dyads both $S$ and $L$ believed in the ``good'' mapping even in
+this implicature-free case, while less than 20\% believed in the
+``bad'' mapping (with the rest being inconsistent).
 
@@
 
-allows us to preserve most of the normative Bayesian framework within the multi-agent setting, while avoiding paradoxes and allowing both pragmatic reasoning and learning.
-The resulting model not only unifies previous models of each phenomenon separately, it also accounts for a range of
+allows us to preserve most of the normative Bayesian framework within
+the multi-agent setting, while avoiding paradoxes and allowing both
+pragmatic reasoning and learning. The resulting model not only unifies
+previous models of each phenomenon separately, it also accounts for a
+range of
 
-the deviations from rationality -- assumption that words have meanings, suspicion that everyone else knows what they are -- are highly interpretable and intuitively compelling, and while like any deviations from rationality they necessarily introduce biases, these biases appear in this case to improve the global performance of the system.
+the deviations from rationality -- assumption that words have
+meanings, suspicion that everyone else knows what they are -- are
+highly interpretable and intuitively compelling, and while like any
+deviations from rationality they necessarily introduce biases, these
+biases appear in this case to improve the global performance of the
+system.
 
-if you're trying to collaborate with an agent who soft-maximizes utility, but there are multiple possible rules which might underlie their utility function, you should assign more weight to those rules which provide a sharp distinction in utility for different actions -- which will generally be the rules that allow for higher expected utility.
-this actually explains both the uniform prior and the horn implicature emergence results.
+if you're trying to collaborate with an agent who soft-maximizes
+utility, but there are multiple possible rules which might underlie
+their utility function, you should assign more weight to those rules
+which provide a sharp distinction in utility for different actions --
+which will generally be the rules that allow for higher expected
+utility. this actually explains both the uniform prior and the horn
+implicature emergence results.
 
-provides new insight into use of pragmatic reasoning in word learning, the learning of literal meanings given pragmatically strengthened input, and makes the interesting prediction that some implicatures are intrinsically more stable, while others if consistently supported will tend to enter the lexicon and become part of the literal meaning of the relevant words.
+provides new insight into use of pragmatic reasoning in word learning,
+the learning of literal meanings given pragmatically strengthened
+input, and makes the interesting prediction that some implicatures are
+intrinsically more stable, while others if consistently supported will
+tend to enter the lexicon and become part of the literal meaning of
+the relevant words.
 

### 2013-05-31 praglearn.tex
more text tinkering, especially results

@@
 \begin{document}
-
 \maketitle
@@
 as proposals for bringing the two together. We next simulate findings
-from the domains of pragmatic inference, word learning, and their
-intersection. We end by discussing the implications of our results for
+on pragmatic inference in one-shot games (replicating previous
+work). We then build on these results to simulate the results of pragmatic
+learning in the language acquisition setting where one communicator is
+uncertain about the lexicon and in iterated communication games where both
+communicators are uncertain about the lexicon. We end by discussing the implications of our results for
 work on language change.
@@
 \subsection{Paradoxes in optimal models of pragmatic learning}
-% When modelling human behaviour, it's useful to begin with optimal
+% When modelling human behavior, it's useful to begin with optimal
 % models; this has the advantage that when we are then forced to
@@
 up only a small subspace of all possible strategies. Nor is there much
-evidence the humans find such equilibria; their behaviour in language
+evidence the humans find such equilibria; their behavior in language
 games (and in other games \cite{camerer2004}) can be well-modeled as
@@
 % nothing for the listener to learn from the speaker's
-% behaviour.
+% behavior.
 
@@
 listeners will eventually learn): Under assumption (a), observing
-other speakers' behaviour provides data about not just that speaker's
+other speakers' behavior provides data about not just that speaker's
 idiosyncratic lexicon, but the consensus lexicon. Assumption (b)
@@
 agents have non-trivial uncertainty, this model induces complex and
-non-normative learning dynamics. In the remainder of this paper, we
-investigate the model's behaviour in all three situations.\footnote{In
+non-normative learning dynamics.
+
+In the remainder of the paper, we apply the model described above to a set of one-shot
+pragmatic inference games that have been well-studied in linguistics
+\cite{grice1975,horn1984} and are addressed by previous one-shot
+models of pragmatic inference \cite{frank2012,bergen2012}. These
+situations set the stage for simulations investigating how learning
+proceeds in iterated versions of such games, described in the
+following section. Results captured by our model and previous models
+are summarized in Table \ref{tab:results}. In
   our simulations throughout, we somewhat arbitrarily set $n = 2$ (the
@@
   and full knowledge of the cost function. Inference is via importance
-  sampling.}
+  sampling.
 
@@
 
-\section{Results: Pure pragmatic inference}
+\section{Pragmatic inference in one-shot games}
 
@@
 \end{table}
-
-We begin by applying the model described above to a set of one-shot
-pragmatic inference games that have been well-studied in linguistics
-\cite{grice1975,horn1984} and are addressed by previous one-shot
-models of pragmatic inference \cite{frank2012,bergen2012}. These
-situations set the stage for simulations investigating how learning
-proceeds in iterated versions of such games, described in the
-following section. Results captured by our model and previous models
-are summarized in Table \ref{tab:results}.
 
@@
 % know where you are, good worlds have more peaked utility
-% distributions than bad worlds, and thus more consistent behaviour by
+% distributions than bad worlds, and thus more consistent behavior by
 % other agents, so on average when you're guessing what other agents
@@
 speaker has higher expected utility than the bad system; therefore, a
-soft-max agent in the bad system has noiser behaviour than in the good
-system, and their behaviour in the good system dominates the average,
+soft-max agent in the bad system has noiser behavior than in the good
+system, and their behavior in the good system dominates the average,
 breaking the symmetry. Similar reasoning applies to an uncertain
@@
 ignorant, while the actual speaker and listener know everything there
-is to know about each other's behaviour. However, the core result
+is to know about each other's behavior. However, the core result
 carries over to our model. For example, with a uniform prior over
@@
 
-\section{Results: The role of pragmatics in language acquisition}
+\section{The role of pragmatics in language acquisition}
 
@@
 
-\begin{SCfigure}[1]
-  \centering
-  % \includegraphics[width=0.24\textwidth]{figures/ME-1dax.pdf}
-  % \includegraphics[width=0.24\textwidth]{figures/ME-flat-10dog-10dax.pdf}
-  % \includegraphics[width=0.24\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
-  % \includegraphics[width=0.24\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
-\includegraphics[width=.6\textwidth]{figures/ME-composite-small.pdf}
-  \caption{Results for simulations of the ``disambiguation'' phenomenon.
+% \begin{SCfigure}[1]
+%   \centering
+%   % \includegraphics[width=0.24\textwidth]{figures/ME-1dax.pdf}
+%   % \includegraphics[width=0.24\textwidth]{figures/ME-flat-10dog-10dax.pdf}
+%   % \includegraphics[width=0.24\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
+%   % \includegraphics[width=0.24\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
+% \includegraphics[width=.6\textwidth]{figures/ME-composite-small.pdf}
+%   \caption{Results for simulations of the ``disambiguation'' phenomenon.
 A novel word is used in a situation with a novel object and a familiar object.
@@
 Left plot shows the posterior after a single situation; right plot shows multiple situations.
-%The right middle plot shows the posterior after a single situation given either a prior that makes it likely that there are words referring to both the novel and familiar objects (perhaps because the two objects are physically similar; the right plot shows a situation where the prior makes it unlikely that a word refers to both (objects dissimilar).
-  }
-  \label{fig:me}
-\end{SCfigure}
+% %The right middle plot shows the posterior after a single situation given either a prior that makes it likely that there are words referring to both the novel and familiar objects (perhaps because the two objects are physically similar; the right plot shows a situation where the prior makes it unlikely that a word refers to both (objects dissimilar).
+%   }
+%   \label{fig:me}
+% \end{SCfigure}
 
@@
 \cite{markman1988}. This phenomenon is sometimes referred to as
-``mutual exclusivity'' or even ``Fast mapping,'' and many distinct
-theoretical accounts have been proposed in the psychological
-literature. Simple probabilistic word learning models can produce a
-similar pattern of findings \cite{frank2009}, often due to the kinds
-of ``explaining away'' type phenomena that are observed in such
-models. But all such models assume that learners in fact retain the
-mapping between word and object that is demonstrated in the
-experimental situation; this observation is contradicted, however, by
+``mutual exclusivity'' or even ``fast mapping.'' Simple probabilistic word learning models can produce a
+similar pattern of findings \cite{frank2009}, but all such models assume that learners retain the
+mapping between novel word and novel object demonstrated in the
+experimental situation. This observation is contradicted, however, by
 evidence that children often do not retain the mappings that are
@@
 probability (e.g. 75\%) that the speaker is referring to the novel
-object. Nevertheless, this is not accompanied by an increased belief
-that the novel word literally refers to this object (as shown in
-Fig. \ref{fig:me}, left). This comes in part from a potential
+object. Nevertheless, this inference is not accompanied by an increased belief
+that the novel word literally refers to this object. The learner's
+interpretation arises not from lexical mapping but instead from a kind
+of 
 specificity implicature: the listener knows that the familiar word
-\emph{does not} refer to the novel object--hence the novel word (even
+\emph{does not} refer to the novel object---hence the novel word (even
 if it refers to the familiar object as well) is the best way to refer
 to the novel object. Nevertheless, on repeated exposure to the same
-novel word, novel object situation, the model does lexicalize this
-inference (Fig. \ref{fig:me}, right).
+novel word, novel object situation, the learner does learn the mapping
+as part of the lexicon (congruent with other data on repeated training
+on disambiguation situations \cite{bion2012}).
 
@@
 
-\begin{SCfigure}[1]
-\centering
-\includegraphics[width=1.5in]{figures/some-all-only-pragmatic.pdf}
-\includegraphics[width=1.5in]{figures/some-all-pragmatic+unambiguous.pdf}
-\caption{\label{fig:scalar} Simulations..}
-\end{SCfigure}
+% \begin{SCfigure}[1]
+% \centering
+% \includegraphics[width=1.5in]{figures/some-all-only-pragmatic.pdf}
+% \includegraphics[width=1.5in]{figures/some-all-pragmatic+unambiguous.pdf}
+% \caption{\label{fig:scalar} Simulations..}
+% \end{SCfigure}
 
@@
 word ``some'' is used to mean \textsc{some but not all}, how do
-children learn that this is not in fact its literal meaning? Our model
+children learn that {\sc some but not all} is not in fact its literal meaning? Our model
 is able to take specificity implicatures into account when learning,
-and thus potentially answer this puzzle. Imagine two bowls: one
+and thus provide a potential solution.
+
+Imagine two bowls: one
 containing a mixture of apples and oranges, the other containing
-apples alone. Fig.~\ref{fig:scalar} (left) shows the model's beliefs
-after encountering the words ``some'' and ``all'' used contrastively
+apples alone. After encountering the words ``some'' and ``all'' used contrastively
 to refer to these two bowls (as in e.g.\ ``the bowl with some
-apples''). It believes that at least one of these words must have a
-restrictive literal meaning, but not both -- it leaves open the
+apples''), a learner under our model believes that at least one of these words must have a
+restrictive literal meaning, but not both---it leaves open the
 possibility that either ``some'' or ``all'' might be literally
-ambiguous. Fig.~\ref{fig:scalar} (right) then shows the model's
-beliefs if we also provide examples of ``some'' referring to exactly
-the same object, but in contexts where this is the only object
+ambiguous. But if examples of ``some'' are given that refer to exactly
+the same object in contexts where this is the only object
 available, so there is no contrast and thus no implicature; this
@@
 
-\section{Results: Mutual learning}
+\section{Mutual learning situations}
 
@@
 an abstract figure vaguely resembling an ice skater,
-\cite{clark1986}). From a pure learning perspective this is anomalous,
+\cite{clark1986}). From a pure learning perspective this behavior is anomalous,
 however: Since both agents know perfectly well that there is no
 underlying convention to discover, there is nothing to learn from the
-other's behaviour. Furthermore, even if only one partner is producing
-the novel expressions, their behaviour still becomes more regular
+other's behavior. Furthermore, even if only one partner is producing
+the novel expressions, their behavior still becomes more regular
 (conventional) over time, which would seem to rule out a role for
@@
 \textit{speaker} to learn by observing these patterns, and thus their
-behaviour should not change over time.
+behavior should not change over time.
 % An obvious hack might be to simply stipulate that speakers take
@@
 
-Pragmatic inference here does two things. First, the pragmatic
-recursion makes it possible for the speaker to learn from the
-listener's interpretations, which would be impossible without an
-explicit model of the listener as a rational agent. Second, pragmatics
-creates the sparse structure in the observed lexicons: At each step,
-the goal-directed speaker attempts to exploit whatever communicatively
-useful asymmetries exist in its current model of the listener, and
-these attempts tend to be successful. If the listener has even a small
-asymmetry in their current posterior then their pragmatic reasoning
-will magnify it as well.
+% Pragmatic inference here does two things. First, the pragmatic
+% recursion makes it possible for the speaker to learn from the
+% listener's interpretations, which would be impossible without an
+% explicit model of the listener as a rational agent. Second, pragmatics
+% creates the sparse structure in the observed lexicons: At each step,
+% the goal-directed speaker attempts to exploit whatever communicatively
+% useful asymmetries exist in its current model of the listener, and
+% these attempts tend to be successful. If the listener has even a small
+% asymmetry in their current posterior then their pragmatic reasoning
+% will magnify it as well.
 
@@
 % \includegraphics[width=0.48\textwidth]{figures/horn-emergence-1.pdf}
-\caption{\label{fig:horn} Example simulations showing the lexicalization of horn implicatures. Plotting conventions are as above.}
+\caption{\label{fig:horn} Example simulations showing the
+  lexicalization of Horn implicatures. Plotting conventions are as
+  above. In the first run, speaker and listener converge on a sparse
+  and  efficient communicative equilibrium, while in the second they reach
+  a sub-optimal equilibrium.}
 \end{SCfigure}
@@
 \includegraphics[width=0.3\textwidth]{figures/emergence-horn-average.pdf}
-\caption{\label{fig:emergence-average} Averaged behaviour over 300
+\caption{\label{fig:emergence-average} Averaged behavior over 300
   dialogues as in Figs.~\ref{fig:emergence} and \ref{fig:horn}. Left:
@@
 \end{SCfigure}
-A stronger example of how pragmatics can create such biases can be
-observed if we simulate agents playing this naming game in the
-``cheap''/``expensive'' \textsc{common}/\textsc{rare} domain of
-\S\ref{sec:horn-implicature}. Recall that in this domain, pragmatic
-reasoning causes each agent to start out independently biased -- but
-only weakly -- towards the associations ``cheap'' $\leftrightarrow$
-\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}, and
-that this bias arises exactly because their prior on literal meaning
+
+A stronger example of how pragmatics can create communicative biases can be
+observed by considering the Horn implicatures introduced in \S\ref{sec:horn-implicature}. 
+In these games, pragmatic reasoning causes each agent to start out weakly biasedtowards the associations ``cheap'' $\leftrightarrow$
+\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}; this bias arises because their prior on literal meaning
 is uniform and uninformative. A fully rational listener, therefore,
-observing our uncertain speaker using this mapping, would discount it
+observing an uncertain speaker using this mapping, would discount it
 as arising from this bias, and conclude that the speaker in fact did
-not know which convention was in use. Our socially anxious listener,
-however, believes that the speaker does know which convention is in
-use, and therefore tends to take their behaviour as evidence that the
-good system is in fact in use. Similarly, the socially anxious speaker
-will wager that since on average they will succed more often if the
-listener is using the good system, they might as well try it -- and
-when they succeed, they take this as evidence that the listener was in
+not know which convention was in use. Conventional listeners,
+however, believe that speakers do know which convention is in
+use, and therefore tend to take their behavior as positive evidence that the
+``good'' system is in use. Similarly, conventional speakers
+will wager that since on average they will succed more often if
+listeners are using the ``good'' system, they might as well try
+it. When they succeed, they take their success as evidence that the listener was in
 fact using the good system all along. As a result, dyads in this game
-not only end up converging onto a stable system, but onto specifically
-the good system, at a rate far above chance. However, something
-fundamentally different happens here than in the specificity
-implicature case (\S\ref{sec:learning-specificity-implic}). There, we
-saw that our model could learn hear pragmatically strengthened usages
-of words while recovering the underlying literal meaning. In this
+not only end up converging onto a stable system at a rate far above
+chance (Fig.~\ref{fig:horn}.
+
+Something
+fundamentally different happens in this case than in the specificity
+implicature case (\S\ref{sec:learning-specificity-implic}), however. There, we
+saw that our model could hear pragmatically strengthened usages
+of words while recovering the underlying literal meaning. In the Horn
 case, though, the pragmatic strengthening depends essentially on
 ignorance about literal meaning. As the agents gather more data,
-therefore, the implicature disappears, to be replaced by a belief that
+the implicature disappears, to be replaced by a belief that
 ``cheap'' \textit{literally} means \textsc{common}. To demonstrate
-this, we queried each agent in each dyad about how they would refer to
+this phenomenon, we queried each agent in each dyad about how they would refer to
 or interpret each object and word, \textit{if} the two objects were
 equally common, which cancels the Horn implicature. As shown in
-Fig.~\ref{emergence-average} (right), after 30 turns, in nearly 70\%
+Fig.~\ref{fig:emergence-average} (right), after 30 turns, in nearly 70\%
 of dyads both $S$ and $L$ believed in the ``good'' mapping even in
@@
 \newpage
-
+\small
 \bibliographystyle{plain}

### 2013-05-31 praglearn.tex
tiny bit of conclusion

--- 2013-05-31 praglearn.tex
@@
 
-allows us to preserve most of the normative Bayesian framework within
-the multi-agent setting, while avoiding paradoxes and allowing both
-pragmatic reasoning and learning. The resulting model not only unifies
-previous models of each phenomenon separately, it also accounts for a
-range of
+If each agent in a multi-agent setting assumes the existence of a
+shared, conventional norm underlying everyone's behaviour, and further
+assumes that other agents are knowledgeable about this norm while they
+themselves are not, then we are able to preserve most of the normative
+Bayesian framework, explicitly model both learning and recursive
+social reasoning simultaneously within the same framework, and avoid
+paradox. The resulting theory unifies two previously separate
+strands of modelling work on language usage and acquisition,
+accounting for a superset of previous phenomena, while producing a
+variety of new qualitative patterns. In particular, we make new
 

### 2013-05-31 praglearn.tex
little edits

--- 2013-05-31 praglearn.tex
@@
 
-\title{Learning words by recursive pragmatic reasoning about other agents}
+\title{Learning communication systems through recursive pragmatic
+  reasoning about other agents}
+% \title{Learning words via recursive pragmatic reasoning about other agents}
 % : The blessing of social anxiety - not a huge fan of colon titles
@@
 communicative goal (``I want you to close the window right now because
-I'm cold.'') -- a process known as \textit{pragmatic reasoning}.
+I'm cold''), a process known as \textit{pragmatic reasoning}.
 
@@
 specifically, it assigns each word $w$ a vector describing the extent
-to which this word provides evidence for each object\footnote{We
+to which this word provides evidence for each object.\footnote{We
   assume words refer directly to objects, rather than to abstract
@@
   $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
-  \text{features}) P_\lex(\text{features} | \word)$.}. Each entry in
+  \text{features}) P_\lex(\text{features} | \word)$.} Each entry in
 the vector is in $(0, 1)$, and together they sum to 1:

### 2013-05-31 praglearn.tex
edits to model section

--- 2013-05-31 praglearn.tex
@@
 specifically, it assigns each word $w$ a vector describing the extent
-to which this word provides evidence for each object\footnote{We
+to which this word provides evidence for each possible object\footnote{We
   assume words refer directly to objects, rather than to abstract
@@
   \text{features}) P_\lex(\text{features} | \word)$.}. Each entry in
-the vector is in $(0, 1)$, and together they sum to 1:
+the vector is in $(0, 1)$, and together they sum to 1. By Bayes rule:
 \begin{align*}
@@
 
-If each agent in a multi-agent setting assumes the existence of a
-shared, conventional norm underlying everyone's behaviour, and further
-assumes that other agents are knowledgeable about this norm while they
-themselves are not, then we are able to preserve most of the normative
-Bayesian framework, explicitly model both learning and recursive
-social reasoning simultaneously within the same framework, and avoid
-paradox. The resulting theory unifies two previously separate
-strands of modelling work on language usage and acquisition,
-accounting for a superset of previous phenomena, while producing a
-variety of new qualitative patterns. In particular, we make new
+allows us to preserve most of the normative Bayesian framework within
+the multi-agent setting, while avoiding paradoxes and allowing both
+pragmatic reasoning and learning. The resulting model not only unifies
+previous models of each phenomenon separately, it also accounts for a
+range of
 

### 2013-05-31 praglearn.tex

--- 2013-05-31 praglearn.tex
@@
 specifically, it assigns each word $w$ a vector describing the extent
-to which this word provides evidence for each object.\footnote{We
+to which this word provides evidence for each possible object\footnote{We
   assume words refer directly to objects, rather than to abstract

### 2013-05-31 praglearn.tex
first pass at conclusion

--- 2013-05-31 praglearn.tex
@@
 in demonstrations of the ``disambiguation effect,'' a novel word
-(e.g. ``dax'') is used in a context containing both a novel and a
+(e.g.~``dax'') is used in a context containing both a novel and a
 familiar object. From quite young, children tend to make the inference
@@
 relatively complex word meanings can be inferred from corpus data
-(e.g. \cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).
+(e.g.~\cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).
 
@@
   involve reasoning about context \cite{clark1988,baldwin1993}, though
-  they also show weaknesses in specific inferences (e.g. strengthening
+  they also show weaknesses in specific inferences (e.g.~strengthening
   ``some'' to ``some but not all'' \cite{papafragou2003}). Here we
@@
 \toprule
-Phenomenon & Ref. & WL & PI & PI+U & PIL \\
+Phenomenon & Ref. & WL & PI & PI+U & PI+WL \\
 \midrule
@@
 Disambiguation without mapping & \cite{horst2008} &  & & & x \\
+Learning/lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
 Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
-Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 \hline
@@
   lexical uncertainty, marginalizes it out, and then recurses. Our
-  current model is referred to here as PIL, or pragmatic inference and
-  learning.}
+  current model is referred to here as PI+WL, and combines pragmatic
+  inference with word learning.}
 \end{table}
@@
 Children, when presented with both a novel and a familiar object
-(e.g. an eggbeater and a ball), will treat a novel label
-(e.g. ``dax'') as referring to the novel object, for example by
+(e.g.~an eggbeater and a ball), will treat a novel label
+(e.g.~``dax'') as referring to the novel object, for example by
 supplying the eggbeater when asked to ``give me the dax''
@@
 given a single disambiguation situation, the model gives a substantial
-probability (e.g. 75\%) that the speaker is referring to the novel
+probability (e.g.~75\%) that the speaker is referring to the novel
 object. Nevertheless, this inference is not accompanied by an increased belief
@@
 
+agents not sure whether some literally means some-not-all or not,
+could lexicalize given the right kind of input, but regular pragmatic
+strengthening alone is not enough
+
 \section{Mutual learning situations}
@@
 \subsection{Emergence of efficient communicative conventions}
+\label{sec:emergence}
 
@@
 this implicature-free case, while less than 20\% believed in the
-``bad'' mapping (with the rest being inconsistent).
+``bad'' mapping (with the rest being inconsistent). This makes the
+interesting prediction that while specificity implicatures can survive
+indefinitely in a language system, any Horn implicatures that arise
+consistently should transition to becoming part of the literal meaning
+of the words involved.
 
@@
 assumes that other agents are knowledgeable about this norm while they
-themselves are not, then we are able to preserve most of the normative
-Bayesian framework, explicitly model both learning and recursive
-social reasoning simultaneously within the same framework, and avoid
-paradox. The resulting theory unifies two previously separate
-strands of modelling work on language usage and acquisition,
-accounting for a superset of previous phenomena, while producing a
-variety of new qualitative patterns. In particular, we make new
+themselves are not, then we are able to simultaneously preserve most
+of the normative Bayesian framework, explicitly model learning and
+recursive social reasoning within the same framework, and avoid
+paradox. The resulting model unifies two previously separate strands
+of modelling work on language usage and acquisition, and -- having
+done so -- accounts for not just the phenomena these previous models
+accounted for, but a variety of new ones as well. We propose new
+explanations of the child language acquisition phenomena of
+disambiguation without mapping \cite{horst2008} and the acquisition of
+scalar adjectives (XX ask Mike what the levinson cite is really, see
+also table), and demonstrate that our model is capable of developing
+novel and efficient communicative systems through iterated learning
+within the context of a single simulated conversation.
 
-the deviations from rationality -- assumption that words have
-meanings, suspicion that everyone else knows what they are -- are
-highly interpretable and intuitively compelling, and while like any
-deviations from rationality they necessarily introduce biases, these
-biases appear in this case to improve the global performance of the
-system.
+Accomplishing this requires limited deviations from rationality, and
+as such necessarily introduces biases in agent behaviour. We've
+identified two such biases here: the tendency for pragmatic speakers
+and listeners to accentuate useful, sparse patterns in their
+communicative systems (\S\ref{sec:emergence}), and for cheap words to
+be assigned to common objects (\S\ref{sec:horn-emergence}).
+Remarkably, while both are locally sub-optimal for individual agents,
+they both systematically drive the overall communicative system
+towards greater global efficiency. In the long term, these processes
+should leave their mark on the structure of the language itself, which
+may provide a partial explanation for how languages become so tightly
+optimized for effective communication \cite{zipf1936,piantadosi2011}.
 
-if you're trying to collaborate with an agent who soft-maximizes
-utility, but there are multiple possible rules which might underlie
-their utility function, you should assign more weight to those rules
-which provide a sharp distinction in utility for different actions --
-which will generally be the rules that allow for higher expected
-utility. this actually explains both the uniform prior and the horn
-implicature emergence results.
+% the deviations from rationality -- assumption that words have
+% meanings, suspicion that everyone else knows what they are -- are
+% highly interpretable and intuitively compelling, and while like any
+% deviations from rationality they necessarily introduce biases, these
+% biases appear in this case to improve the global performance of the
+% system.
 
-provides new insight into use of pragmatic reasoning in word learning,
-the learning of literal meanings given pragmatically strengthened
-input, and makes the interesting prediction that some implicatures are
-intrinsically more stable, while others if consistently supported will
-tend to enter the lexicon and become part of the literal meaning of
-the relevant words.
+% if you're trying to collaborate with an agent who soft-maximizes
+% utility, but there are multiple possible rules which might underlie
+% their utility function, you should assign more weight to those rules
+% which provide a sharp distinction in utility for different actions --
+% which will generally be the rules that allow for higher expected
+% utility. this actually explains both the uniform prior and the horn
+% implicature emergence results.
 
-zipf, piantidosi et al
+% provides new insight into use of pragmatic reasoning in word learning,
+% the learning of literal meanings given pragmatically strengthened
+% input, and makes the interesting prediction that some implicatures are
+% intrinsically more stable, while others if consistently supported will
+% tend to enter the lexicon and become part of the literal meaning of
+% the relevant words.
+
+% zipf, piantidosi et al
 

### 2013-05-31 praglearn.tex
clean up some todo notes
section 4.2

--- 2013-05-31 praglearn.tex
@@
 
-\title{Learning communication systems through recursive pragmatic
+\title{Learning to communicate using recursive pragmatic
   reasoning about other agents}
@@
   speakers' intentions in context, and children learning their native
-  language also display substantial pragmatic skill in acquiring the
-  meanings of unknown words. While pragmatic inference and word
-  learning have both been characterized in probabilistic terms, no
-  current models make appropriate inferences about the meanings of
-  words in pragmatically-rich inferential contexts. We describe a
+  language also display substantial skill in acquiring the
+  meanings of unknown words. These two cases are deeply related:
+  Language users invent new terms in conversation, and language
+  learners learn the literal meanings of words based on their
+  pragmatic inferences about how those words are used. 
+ While pragmatic inference and word
+  learning have both been independently characterized in probabilistic
+  terms, no current work unifies these two. We describe a
   model in which language learners assume that they jointly
@@
 Natural language is a powerful tool for conveying meaning in
-context. Our utterances are pervasively underspecified. Consider
+context. Consider
 ``it's raining,'' ``I ate some of the cookies,'' or ``can you close
@@
 I'm cold''), a process known as \textit{pragmatic reasoning}. 
+This kind of pragmatic reasoning is core to language use,
 
@@
 specifically, it assigns each word $w$ a vector describing the extent
-to which this word provides evidence for each possible object\footnote{We
+to which this word provides evidence for each possible object.\footnote{We
   assume words refer directly to objects, rather than to abstract
@@
 Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
-Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
+Non-lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 \hline
@@
 \subsection{Specificity implicature}
+
+\label{sec:spec}
 
@@
   \textit{some} and a Dirichlet(1, 10) prior on the meaning of
-  \textit{all}, and find this gives speaker use probabilities of
+  \textit{all}, giving speaker probabilities of
   $\left[
@@
   \right]$
-and listener interpretation probabilities of
+and listener probabilities of
 $\left[
@@
     \end{array}
-  \right]$.
+  \right]$. These revised interpretation probabilities minimize
+  ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.
 
@@
 
-\subsection{Lexicalization and preservation of specificity implicature}
+\subsection{Preservation of specificity implicatures}
 \label{sec:learning-specificity-implic}
@@
 for most models of word learning: given that in many contexts, the
-word ``some'' is used to mean \textsc{some but not all}, how do
-children learn that {\sc some but not all} is not in fact its literal meaning? Our model
+word ``some'' is used to mean \textsc{some-but-not-all}, how do
+children learn that {\sc some-but-not-all} is not in fact its literal meaning? Our model
 is able to take specificity implicatures into account when learning,
-and thus provide a potential solution.
+and thus provide a potential solution, congruent with the observation
+that no known language in fact lexicalizes {\sc some-but-not-all}.
 
-Imagine two bowls: one
-containing a mixture of apples and oranges, the other containing
-apples alone. After encountering the words ``some'' and ``all'' used contrastively
-to refer to these two bowls (as in e.g.\ ``the bowl with some
-apples''), a learner under our model believes that at least one of these words must have a
-restrictive literal meaning, but not both---it leaves open the
-possibility that either ``some'' or ``all'' might be literally
-ambiguous. But if examples of ``some'' are given that refer to exactly
-the same object in contexts where this is the only object
-available, so there is no contrast and thus no implicature; this
-convinces the model that ``all apple'' literally refers only to the
-bowl that contains no oranges, while ``some'' most likely is somewhat
-ambiguous. Crucially, the model is able to learn this without once
-ever observing ``some'' used
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.98\textwidth]{figures/emergence-composite.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
+% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
+\caption{\label{fig:emergence} Simulations of two pragmatic agents
+  playing a naming game. Each panel shows two representative
+  simulation runs, with run 1 chosen to show strong convergence and
+  run 2 chosen to show relatively weaker convergence. At each stage,
+  $S$ and $L$ have different, possibly contradictory posteriors over
+  the conventional, consensus lexicon. From these posteriors we derive
+  the probability $P(\text{$L$ understands $S$})$ (marginalizing over
+  target objects and word choices), and also depict graphically $S$'s
+  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$,
+  and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s
+    data})$. }
+\end{figure}
+
+Following the details of \S\ref{sec:spec}, we created a simulation in
+which the meaning of ``all'' was fixed to be a particular set {\sc
+  all}, and where the learner must infer the meaning of ``some,''
+distinguishing two possible meanings: {\sc some-but-not-all}
+(unambiguous, 
+pragmatically strengthened) or
+{\sc some-but-not-all or all} (ambiguous). Training exposure in the
+simulation consisted of situations in which ``some'' was pragmatically
+strengthened to refer to {\sc some-but-not-all}. Despite this
+unambiguous training, the model maintained substantial
+posterior probability on the correct (ambiguous) hypothesis about the
+meaning of ``some.'' In each
+situation, the model essentially reasoned that although the evidence
+was unambiguous, it could have been produced by an implicature and
+hence should be discounted. Thus, a pragmatically-informed learner
+might both be able to maintain the true meaning of {\sc some} and
+interpret it correctly as denoting {\sc some-but-not-all} in
+appropriate contexts.
+
+% Exposure to these
+% situations does not result in mapping of ``some'' to {\sc
+%   some-but-not-all}, despite the evidence that 
+% Imagine two bowls: one
+% containing a mixture of apples and oranges, the other containing
+% apples alone. After encountering the words ``some'' and ``all'' used contrastively
+% to refer to these two bowls (as in e.g.\ ``the bowl with some
+% apples''), a learner under our model believes that at least one of these words must have a
+% restrictive literal meaning, but not both---it leaves open the
+% possibility that either ``some'' or ``all'' might be literally
+% ambiguous. But if examples of ``some'' are given that refer to exactly
+% the same object in contexts where this is the only object
+% available, so there is no contrast and thus no implicature; this
+% convinces the model that ``all apple'' literally refers only to the
+% bowl that contains no oranges, while ``some'' most likely is somewhat
+% ambiguous. Crucially, the model is able to learn this without once
+% ever observing ``some'' used
+
+% s1: reasonable chance that ``some'' literally refers to only one, but
+% maybe ``some'' refers to both. so you can learn ambiguity even with
+% unambiguous evidence.
+
+% s2:
 
@@
 % krauss 1966, branigan catchpole pickering 2011).
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.98\textwidth]{figures/emergence-composite.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
-% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
-\caption{\label{fig:emergence} Simulations of two pragmatic agents
-  playing a naming game. Each panel shows two representative
-  simulation runs, with run 1 chosen to show strong convergence and
-  run 2 chosen to show relatively weaker convergence. At each stage,
-  $S$ and $L$ have different, possibly contradictory posteriors over
-  the conventional, consensus lexicon. From these posteriors we derive
-  the probability $P(\text{$L$ understands $S$})$ (marginalizing over
-  target objects and word choices), and also depict graphically $S$'s
-  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$,
-  and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s
-    data})$. }
-\end{figure}
 

### 2013-05-31 praglearn.tex

--- 2013-05-31 praglearn.tex
@@
 specifically, it assigns each word $w$ a vector describing the extent
-to which this word provides evidence for each object.\footnote{We
+to which this word provides evidence for each possible object.\footnote{We
   assume words refer directly to objects, rather than to abstract
@@
 Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
+Non-lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 \hline
@@
 \label{sec:horn-implicature}
-
-% XX note the relation to BGL model; they need uncertainty in the
-% middle of the pragmatics specifically, which is where we end up
-% putting it for other reasons. And in Bayesian framework uncertainty
-% and learning must always coincide. Also the intuition that this is a
-% general property of uncertainty for soft-max agents -- if you don't
-% know where you are, good worlds have more peaked utility
-% distributions than bad worlds, and thus more consistent behavior by
-% other agents, so on average when you're guessing what other agents
-% will do, good worlds contribute more to your average than bad
-% worlds.
 
@@
 % some kind of feedback indicating that the listener understands; if
-% such feedback is removed, the phenomenon is dramatically reduced (XX
+% such feedback is removed, the phenomenon is dramatically reduced (X X
 % krauss 1966, branigan catchpole pickering 2011).
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.98\textwidth]{figures/emergence-composite.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
-% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
-\caption{\label{fig:emergence} Simulations of two pragmatic agents
-  playing a naming game. Each panel shows two representative
-  simulation runs, with run 1 chosen to show strong convergence and
-  run 2 chosen to show relatively weaker convergence. At each stage,
-  $S$ and $L$ have different, possibly contradictory posteriors over
-  the conventional, consensus lexicon. From these posteriors we derive
-  the probability $P(\text{$L$ understands $S$})$ (marginalizing over
-  target objects and word choices), and also depict graphically $S$'s
-  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$,
-  and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s
-    data})$. }
-\end{figure}
 
@@
 
-% XXXX what can we say about the role pragmatics plays here? It's
+% X X X X what can we say about the role pragmatics plays here? It's
 % annoyingly non-trivial to sample from the equivalent model where the

### 2013-05-31 praglearn.tex
models ection

--- 2013-05-31 praglearn.tex
@@
 
-\title{Learning to communicate using recursive pragmatic
+\title{Learning communication systems through recursive pragmatic
   reasoning about other agents}
@@
 \newcommand{\lex}{\text{lexicon}}
-\newcommand{\prior}{P_{\text{prior}}(\obj | \text{context})}
+%\newcommand{\prior}{P_{\text{prior}}(\obj | \text{context})}
+\newcommand{\prior}{P_{\text{prior}}(\obj)}
 
@@
   speakers' intentions in context, and children learning their native
-  language also display substantial skill in acquiring the
-  meanings of unknown words. These two cases are deeply related:
-  Language users invent new terms in conversation, and language
-  learners learn the literal meanings of words based on their
-  pragmatic inferences about how those words are used. 
- While pragmatic inference and word
-  learning have both been independently characterized in probabilistic
-  terms, no current work unifies these two. We describe a
+  language also display substantial pragmatic skill in acquiring the
+  meanings of unknown words. While pragmatic inference and word
+  learning have both been characterized in probabilistic terms, no
+  current models make appropriate inferences about the meanings of
+  words in pragmatically-rich inferential contexts. We describe a
   model in which language learners assume that they jointly
@@
 Natural language is a powerful tool for conveying meaning in
-context. Consider
+context. Our utterances are pervasively underspecified. Consider
 ``it's raining,'' ``I ate some of the cookies,'' or ``can you close
@@
 I'm cold''), a process known as \textit{pragmatic reasoning}.
-This kind of pragmatic reasoning is core to language use,
 
@@
 define a \textit{literal listener} $L_0$. This agent has a
-\textit{lexicon} which probabilistically maps words to meanings;
-specifically, it assigns each word $w$ a vector describing the extent
-to which this word provides evidence for each possible object.\footnote{We
+\textit{lexicon} of associations between words and meanings;
+specifically, it assigns each word $w$ a vector of numbers in $(0, 1)$ describing the extent
+to which this word provides evidence for each possible object\footnote{We
   assume words refer directly to objects, rather than to abstract
@@
   $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
-  \text{features}) P_\lex(\text{features} | \word)$.} Each entry in
-the vector is in $(0, 1)$, and together they sum to 1:
+  \text{features}) P_\lex(\text{features} | \word)$.}.%
+If we assume the listener simply re-weights their prior expectation   
+about what is referred to using their lexicon:
 \begin{align*}
@@
 \end{align*}
+Because of the normalization in this equation, there is a systematic but unimportant symmetry among lexicons; we 
+remove this by assuming the lexicon sums to $1$ over objects for each word.
 Confronted with such a listener, a probabilistically rational speaker
@@
 
-One computationally expensive option is to continue iterating such
+One option is to continue iterating such
 strategies until reaching a fixed point equilibrium. While this
@@
 other agent's strategy, there is no guarantee that such strategies
-will be near the system's global optimum; equilibrium strategies make
-up only a small subspace of all possible strategies. Nor is there much
-evidence the humans find such equilibria; their behavior in language
+will be near the system's global optimum. More importantly, is there much
+evidence that humans do not use such equilibrium strategies; their behavior in language
 games (and in other games \cite{camerer2004}) can be well-modeled as
@@
 $S_n$, is fully optimal with respect to the game, while the other,
-implementing $L_{n-1}$, is nearly optimal---off by only a single
+implementing $L_{n-1}$, is only nearly optimal---off by a single
 recursion.
@@
 
-To add insult to injury, one final problem is that there is nothing
+One final problem is that there is nothing
 constraining speaker and listener to use the same language under this
@@
 uncertainty. And assumption (c) introduces a limited form of
-uncertainty, and thus the potential for learning.
+uncertainty at the top level, and thus the potential for learning.
 % Everything else follows by applying standard Bayesian reasoning
@@
 
-Formally, let an unadorned $L$ and $S$ denote our socially-anxious
-listener and speaker. $L$ inverts their marginal distribution over the
-speaker; $S$ soft-maximizes their marginal distribution over the
-listener. With $L_i$ and $S_i$ as above, this gives:
+Formally, let an unadorned $L$ and $S$ denote our convention-based
+listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
+%\begin{align*}
+%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
+%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
+%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
+%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
+%\end{align*}
+%
 \begin{align*}
-  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
-    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior \\
-    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
-  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
+  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
 \end{align*}
@@
 other information $L$ may have about their meaning (e.g.~from
-contextual clues). By assumption (b), $L$ treats these as utterances
+contextual clues). 
+By assumption (b), $L$ treats these  utterances
 as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
-as being informative about the lexicon. Similarly, $S$ may observe
-$L$'s interpretations of various utterances, and will treat these as
+as being informative about the lexicon. For instance, when the data is a set of fully observed 
+word-referent pairs $\{w_i, o_i\}$:
+\begin{align*}
+  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
+\end{align*}
+The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
+\begin{align*}
+  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
+\end{align*}
+Here $P(\lex|\text{$S$'s data}$ is defined similarly, when $S$ observes
+$L$'s interpretations of various utterances, and treats them as
 samples from $L_n$, not $L$.
 
-In several important cases, these assumptions are as close to rational
-as possible within the finite-recursion framework. If $L_n(\obj |
-\word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$
-coincide, then $S$ is fully normative and $L$ is off by only a single
-recursion. This case in particular includes all previous experiments
-using this pragmatic reasoning framework, which considered only
-single-turn interactions where $\text{$L$'s data} = \text{$S$'s data}
-= \emptyset$. Alternatively, if one agent has enough data that their
+%In several important cases, these assumptions are as close to rational
+%as possible within the finite-recursion framework. If $L_n(\obj |
+%\word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$
+%coincide, then $S$ is fully normative and $L$ is off by only a single
+%recursion. This case in particular includes all previous experiments
+%using this pragmatic reasoning framework, which considered only
+%single-turn interactions where $\text{$L$'s data} = \text{$S$'s data}
+%= \emptyset$.
+
+If the speaker has enough data that their
 uncertainty is effectively non-existent, then it becomes normative for
@@
 Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
-Non-lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
+Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 \hline
@@
 \subsection{Specificity implicature}
-
-\label{sec:spec}
 
@@
   \textit{some} and a Dirichlet(1, 10) prior on the meaning of
-  \textit{all}, giving speaker probabilities of
+  \textit{all}, and find this gives speaker use probabilities of
   $\left[
@@
   \right]$
-and listener probabilities of
+and listener interpretation probabilities of
 $\left[
@@
     \end{array}
-  \right]$. These revised interpretation probabilities minimize
-  ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.
+  \right]$.
 
@@
 
-\subsection{Preservation of specificity implicatures}
+\subsection{Lexicalization and preservation of specificity implicature}
 \label{sec:learning-specificity-implic}
@@
 for most models of word learning: given that in many contexts, the
-word ``some'' is used to mean \textsc{some-but-not-all}, how do
-children learn that {\sc some-but-not-all} is not in fact its literal meaning? Our model
+word ``some'' is used to mean \textsc{some but not all}, how do
+children learn that {\sc some but not all} is not in fact its literal meaning? Our model
 is able to take specificity implicatures into account when learning,
-and thus provide a potential solution, congruent with the observation
-that no known language in fact lexicalizes {\sc some-but-not-all}.
+and thus provide a potential solution.
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.98\textwidth]{figures/emergence-composite.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
-% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
-% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
-\caption{\label{fig:emergence} Simulations of two pragmatic agents
-  playing a naming game. Each panel shows two representative
-  simulation runs, with run 1 chosen to show strong convergence and
-  run 2 chosen to show relatively weaker convergence. At each stage,
-  $S$ and $L$ have different, possibly contradictory posteriors over
-  the conventional, consensus lexicon. From these posteriors we derive
-  the probability $P(\text{$L$ understands $S$})$ (marginalizing over
-  target objects and word choices), and also depict graphically $S$'s
-  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$,
-  and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s
-    data})$. }
-\end{figure}
-
-Following the details of \S\ref{sec:spec}, we created a simulation in
-which the meaning of ``all'' was fixed to be a particular set {\sc
-  all}, and where the learner must infer the meaning of ``some,''
-distinguishing two possible meanings: {\sc some-but-not-all}
-(unambiguous, 
-pragmatically strengthened) or
-{\sc some-but-not-all or all} (ambiguous). Training exposure in the
-simulation consisted of situations in which ``some'' was pragmatically
-strengthened to refer to {\sc some-but-not-all}. Despite this
-unambiguous training, the model maintained substantial
-posterior probability on the correct (ambiguous) hypothesis about the
-meaning of ``some.'' In each
-situation, the model essentially reasoned that although the evidence
-was unambiguous, it could have been produced by an implicature and
-hence should be discounted. Thus, a pragmatically-informed learner
-might both be able to maintain the true meaning of {\sc some} and
-interpret it correctly as denoting {\sc some-but-not-all} in
-appropriate contexts.
-
-% Exposure to these
-% situations does not result in mapping of ``some'' to {\sc
-%   some-but-not-all}, despite the evidence that 
-% Imagine two bowls: one
-% containing a mixture of apples and oranges, the other containing
-% apples alone. After encountering the words ``some'' and ``all'' used contrastively
-% to refer to these two bowls (as in e.g.\ ``the bowl with some
-% apples''), a learner under our model believes that at least one of these words must have a
-% restrictive literal meaning, but not both---it leaves open the
-% possibility that either ``some'' or ``all'' might be literally
-% ambiguous. But if examples of ``some'' are given that refer to exactly
-% the same object in contexts where this is the only object
-% available, so there is no contrast and thus no implicature; this
-% convinces the model that ``all apple'' literally refers only to the
-% bowl that contains no oranges, while ``some'' most likely is somewhat
-% ambiguous. Crucially, the model is able to learn this without once
-% ever observing ``some'' used
-
-% s1: reasonable chance that ``some'' literally refers to only one, but
-% maybe ``some'' refers to both. so you can learn ambiguity even with
-% unambiguous evidence.
-
-% s2:
+Imagine two bowls: one
+containing a mixture of apples and oranges, the other containing
+apples alone. After encountering the words ``some'' and ``all'' used contrastively
+to refer to these two bowls (as in e.g.\ ``the bowl with some
+apples''), a learner under our model believes that at least one of these words must have a
+restrictive literal meaning, but not both---it leaves open the
+possibility that either ``some'' or ``all'' might be literally
+ambiguous. But if examples of ``some'' are given that refer to exactly
+the same object in contexts where this is the only object
+available, so there is no contrast and thus no implicature; this
+convinces the model that ``all apple'' literally refers only to the
+bowl that contains no oranges, while ``some'' most likely is somewhat
+ambiguous. Crucially, the model is able to learn this without once
+ever observing ``some'' used
 
@@
 % krauss 1966, branigan catchpole pickering 2011).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.98\textwidth]{figures/emergence-composite.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-1.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence2x2-3.pdf} \\
+% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-0.pdf}
+% \includegraphics[width=0.48\textwidth]{figures/emergence3x3-1.pdf} \\
+\caption{\label{fig:emergence} Simulations of two pragmatic agents
+  playing a naming game. Each panel shows two representative
+  simulation runs, with run 1 chosen to show strong convergence and
+  run 2 chosen to show relatively weaker convergence. At each stage,
+  $S$ and $L$ have different, possibly contradictory posteriors over
+  the conventional, consensus lexicon. From these posteriors we derive
+  the probability $P(\text{$L$ understands $S$})$ (marginalizing over
+  target objects and word choices), and also depict graphically $S$'s
+  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$,
+  and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s
+    data})$. }
+\end{figure}
 

### 2013-05-31 praglearn.tex
fix wonky phrasing

@@
 other agent's strategy, there is no guarantee that such strategies
-will be near the system's global optimum. More importantly, is there much
+will be near the system's global optimum. More importantly, there is a
+great deal of
 evidence that humans do not use such equilibrium strategies; their behavior in language

### 2013-05-31 praglearn.tex
author initials

@@
 \author{
-Nathaniel Smith\thanks{...} \\
+Nathaniel J.~Smith\thanks{...} \\
 School of Informatics\\
@@
 \AND
-Noah D. Goodman \\
+Noah D.~Goodman \\
 Department of Psychology\\
@@
 \And
-Michael C. Frank \\
+Michael C.~Frank \\
 Department of Psychology \\

### 2013-05-31 praglearn.tex
more text

--- 2013-05-31 praglearn.tex
@@
 
-\title{Learning to communicate using recursive pragmatic
-  reasoning about other agents}
+\title{Language learning via recursive pragmatic reasoning}
 % \title{Learning words via recursive pragmatic reasoning about other agents}
@@
 \newcommand{\lex}{\text{lexicon}}
-%\newcommand{\prior}{P_{\text{prior}}(\obj | \text{context})}
-\newcommand{\prior}{P_{\text{prior}}(\obj)}
+\newcommand{\prior}{P_{\text{prior}}(\obj | \text{context})}
 
@@
 Natural language is a powerful tool for conveying meaning in
-context. Consider
+context. Consider the utterances
 ``it's raining,'' ``I ate some of the cookies,'' or ``can you close
@@
 communicative goal (``I want you to close the window right now because
-I'm cold''), a process known as \textit{pragmatic reasoning}. 
-This kind of pragmatic reasoning is core to language use,
-
-Following this intuition, theories of pragmatic communication frame
+I'm cold''), a process known as \textit{pragmatic reasoning}. Theories of pragmatics frame
 the process of language comprehension as inference about the
@@
 would have said `all'---but she didn't. Hence `all' must not be true
-and she must have eaten some {\it but not all} of the cookies.'' This
-kind of counterfactual reasoning has a recursive character to it:
-listener models speaker (who may in turn be modeling a listener).
+and she must have eaten some {\it but not all} of the cookies.'' This kind of reasoning is core to language use.
 
-Even before they are proficient language users, children learn the
-meanings of words from their pattern of use in context. For example,
-in demonstrations of the ``disambiguation effect,'' a novel word
-(e.g.~``dax'') is used in a context containing both a novel and a
-familiar object. From quite young, children tend to make the inference
-that ``dax'' refers to the novel object \cite{markman1988}. In another
-demonstration, even young toddlers can extract mappings between words
-and objects from a pattern of individually-ambiguous co-occurrences
-between pairs of words and pairs of objects \cite{smith2008}. Taking
-inspiration from children's learning prowess, recent work in the field
-of grounded language learning has explored the possibility that even
-relatively complex word meanings can be inferred from corpus data
-(e.g.~\cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).
-
-Recursive pragmatic inferences such as those described above could be
-an important part of language learning.\footnote{Very young children
+But pragmatic reasoning often intersects with language learning as
+well. In both adults and children, uncertainty about word meanings is
+common, and often considering speakers' pragmatic goals can help to resolve this uncertainty. For example,
+if a novel word is used in a context containing both a novel and a
+familiar object, young children can make the inference
+that the novel word refers to the novel object \cite{markman1988}.\footnote{Very young children
   make inferences that are often labeled as ``pragmatic'' in that they
@@
   they also show weaknesses in specific inferences (e.g.~strengthening
-  ``some'' to ``some but not all'' \cite{papafragou2003}). Here we
+  {\sc some} to {\sc some but not all} \cite{papafragou2003}). Here we
   remain agnostic about the age at which children are able to make
   such inferences robustly, as it may vary depending on the linguistic
-  materials being used in the inference \cite{barner2011}.} For
-example, the disambiguation inference described above can emerge from
-such reasoning (``why would the speaker have said ``dax'' if there was
-a word for the known object that she knew I would know'')
-\cite{clark1988}. In addition, there are a variety of intriguing cases
+  materials being used in the inference \cite{barner2011}.} For adults
+who are proficient language users, there are also a variety of intriguing cases
 in which listeners seem to create situation- and task-specific ways of
@@
 them to perform accurately even though they do not begin with a shared
-label \cite{krauss1964,clark1986}. Yet models of grounded language
-learning have failed to take pragmatic factors into account.
+label \cite{krauss1964,clark1986}. In both of these examples, reasoning about another person's goals informs language learners' estimates of what words are likely to mean.
+
+% Recursive pragmatic inferences such as those described above could be
+% an important part of language learning. For
+% example, the disambiguation inference described above can emerge from
+% such reasoning (``why would the speaker have said ``dax'' if there was
+% a word for the known object that she knew I would know'')
+% \cite{clark1988}.Yet models of grounded language
+% learning have failed to take pragmatic factors into account.
+
+Despite this intersection, there is relatively little work that takes pragmatic reasoning into account when considering language learning in context.
+Recent work on grounded language learning has attempted to learn large sets of (sometimes relatively complex) word meanings directly from corpus data
+(e.g.~\cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}). And
+a number of models have begun to formalize the consequences of
+pragmatic reasoning in situations where limited learning takes place
+\cite{golland2010,frank2012,bergen2012}. But as yet these two strands
+of research have not been brought together so that the implications of
+pragmatics for learning can be investigated directly.
+
+% \footnote{Very young children
+%   make inferences that are often labeled as ``pragmatic'' in that they
+%   involve reasoning about context \cite{clark1988,baldwin1993}, though
+%   they also show weaknesses in specific inferences (e.g.~strengthening
+%   {\sc some} to {\sc some but not all} \cite{papafragou2003}). Here we
+%   remain agnostic about the age at which children are able to make
+%   such inferences robustly, as it may vary depending on the linguistic
+%   materials being used in the inference \cite{barner2011}.}
 
@@
 work on language change.
+
+% This
+% kind of counterfactual reasoning has a recursive character to it:
+% listener models speaker (who may in turn be modeling a listener).
+
+% Even before they are proficient language users, children learn the
+% meanings of words from their pattern of use in context.
+
+% In another
+% demonstration, even young toddlers can extract mappings between words
+% and objects from a pattern of individually-ambiguous co-occurrences
+% between pairs of words and pairs of objects \cite{smith2008}.
+
+%
 
@@
 define a \textit{literal listener} $L_0$. This agent has a
-\textit{lexicon} of associations between words and meanings;
-specifically, it assigns each word $w$ a vector of numbers in $(0, 1)$ describing the extent
-to which this word provides evidence for each possible object\footnote{We
+\textit{lexicon} which probabilistically maps words to meanings;
+specifically, it assigns each word $w$ a vector describing the extent
+to which this word provides evidence for each possible object.\footnote{We
   assume words refer directly to objects, rather than to abstract
@@
   $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
-  \text{features}) P_\lex(\text{features} | \word)$.}.%
-If we assume the listener simply re-weights their prior expectation   
-about what is referred to using their lexicon:
+  \text{features}) P_\lex(\text{features} | \word)$.} Each entry in
+the vector is in $(0, 1)$, and together they sum to 1:
 \begin{align*}
@@
 \end{align*}
-Because of the normalization in this equation, there is a systematic but unimportant symmetry among lexicons; we 
-remove this by assuming the lexicon sums to $1$ over objects for each word.
 Confronted with such a listener, a probabilistically rational speaker
@@
 
-One option is to continue iterating such
+One computationally expensive option is to continue iterating such
 strategies until reaching a fixed point equilibrium. While this
@@
 other agent's strategy, there is no guarantee that such strategies
-will be near the system's global optimum. More importantly, is there much
-evidence that humans do not use such equilibrium strategies; their behavior in language
+will be near the system's global optimum; equilibrium strategies make
+up only a small subspace of all possible strategies. Nor is there much
+evidence the humans find such equilibria; their behavior in language
 games (and in other games \cite{camerer2004}) can be well-modeled as
@@
 $S_n$, is fully optimal with respect to the game, while the other,
-implementing $L_{n-1}$, is only nearly optimal---off by a single
+implementing $L_{n-1}$, is nearly optimal---off by only a single
 recursion.
@@
 
-One final problem is that there is nothing
+To add insult to injury, one final problem is that there is nothing
 constraining speaker and listener to use the same language under this
@@
 uncertainty. And assumption (c) introduces a limited form of
-uncertainty at the top level, and thus the potential for learning.
+uncertainty, and thus the potential for learning.
 % Everything else follows by applying standard Bayesian reasoning
@@
 
-Formally, let an unadorned $L$ and $S$ denote our convention-based
-listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
-%\begin{align*}
-%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
-%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
-%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
-%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
-%\end{align*}
-%
+Formally, let an unadorned $L$ and $S$ denote our socially-anxious
+listener and speaker. $L$ inverts their marginal distribution over the
+speaker; $S$ soft-maximizes their marginal distribution over the
+listener. With $L_i$ and $S_i$ as above, this gives:
 \begin{align*}
-  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
+  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
+    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior \\
+    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
+  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
 \end{align*}
@@
 other information $L$ may have about their meaning (e.g.~from
-contextual clues). 
-By assumption (b), $L$ treats these  utterances
+contextual clues). By assumption (b), $L$ treats these as utterances
 as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
-as being informative about the lexicon. For instance, when the data is a set of fully observed 
-word-referent pairs $\{w_i, o_i\}$:
-\begin{align*}
-  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
-\end{align*}
-The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
-\begin{align*}
-  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
-\end{align*}
-Here $P(\lex|\text{$S$'s data}$ is defined similarly, when $S$ observes
-$L$'s interpretations of various utterances, and treats them as
+as being informative about the lexicon. Similarly, $S$ may observe
+$L$'s interpretations of various utterances, and will treat these as
 samples from $L_n$, not $L$.
 
-%In several important cases, these assumptions are as close to rational
-%as possible within the finite-recursion framework. If $L_n(\obj |
-%\word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$
-%coincide, then $S$ is fully normative and $L$ is off by only a single
-%recursion. This case in particular includes all previous experiments
-%using this pragmatic reasoning framework, which considered only
-%single-turn interactions where $\text{$L$'s data} = \text{$S$'s data}
-%= \emptyset$.
-
-If the speaker has enough data that their
+In several important cases, these assumptions are as close to rational
+as possible within the finite-recursion framework. If $L_n(\obj |
+\word, \text{$S$'s data})$ and $L_n(\obj | \word, \text{$L$'s data})$
+coincide, then $S$ is fully normative and $L$ is off by only a single
+recursion. This case in particular includes all previous experiments
+using this pragmatic reasoning framework, which considered only
+single-turn interactions where $\text{$L$'s data} = \text{$S$'s data}
+= \emptyset$. Alternatively, if one agent has enough data that their
 uncertainty is effectively non-existent, then it becomes normative for
@@
 this situation. So although ``I ate some of the cookies'' could in
-principle be compatible with eating all of them, the listener is lead
-to believe that ``some but not all'' is the likely state of
+principle be compatible with eating {\sc all} of them, the listener is lead
+to believe that {\sc some-but-not-all}' is the likely state of
 affairs. We use the term {\it specificity implicature} here (following
@@
   situation, we place a Dirichlet(10, 10) prior on the meaning of
-  \textit{some} and a Dirichlet(1, 10) prior on the meaning of
-  \textit{all}, giving speaker probabilities of
+``some'' and a Dirichlet(1, 10) prior on the meaning of
+``some'', giving speaker probabilities of
   $\left[

### 2013-05-31 praglearn.tex
merge

--- 2013-05-31 praglearn.tex
@@
 \author{
-Nathaniel J.~Smith\thanks{...} \\
+Nathaniel Smith\thanks{...} \\
 School of Informatics\\
@@
 \AND
-Noah D.~Goodman \\
+Noah D. Goodman \\
 Department of Psychology\\
@@
 \And
-Michael C.~Frank \\
+Michael C. Frank \\
 Department of Psychology \\
@@
 other agent's strategy, there is no guarantee that such strategies
-will be near the system's global optimum. More importantly, there is a
-great deal of
+will be near the system's global optimum. More importantly, is there much
 evidence that humans do not use such equilibrium strategies; their behavior in language
@@
 \end{align*}
-Here $P(\lex|\text{$S$'s data}$ is defined similarly, when $S$ observes
+Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
 $L$'s interpretations of various utterances, and treats them as
@@
 Many sets of words in natural language form scales in which each term
-refers to successively stronger restrictions on the situations
-compatible with the term. ``Some'' and ``all'' form a scale of this
+is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
 type. While ``I ate some of the cookies'' is compatible with the

### 2013-05-31 praglearn.tex
remove stray bit of detritus
remove repeated line in table

--- 2013-05-31 praglearn.tex
@@
 Disambiguation without mapping & \cite{horst2008} &  & & & x \\
-Learning/lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
@@
 
-agents not sure whether some literally means some-not-all or not,
-could lexicalize given the right kind of input, but regular pragmatic
-strengthening alone is not enough
+% agents not sure whether some literally means some-not-all or not,
+% could lexicalize given the right kind of input, but regular pragmatic
+% strengthening alone is not enough
 

### 2013-05-31 praglearn.tex
center side captions, and clarify fig. 1 caption

@@
 \usepackage[raggedright]{sidecap}
+
+% http://www.latex-community.org/forum/viewtopic.php?f=45&t=12364
+\sidecaptionvpos{figure}{c}
 
@@
   target objects and word choices), and also depict graphically $S$'s
-  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$,
+  model of the listener, $P_{L_n}(\obj | \word, \text{$S$'s data})$
+  (top row within each run),
   and $L$'s actual model, $P_{L_n}(\obj | \word, \text{$L$'s
-    data})$. }
+    data})$ (bottom row within each run). }
 \end{figure}

### 2013-05-31 praglearn.tex
merge

--- 2013-05-31 praglearn.tex
@@
 
-Natural language is a powerful tool for conveying meaning in
-context. Consider the utterances
+Two puzzles present themselves to language users: What do words mean in general and what do they mean in context?
+%Natural language is a powerful tool for conveying meaning in context. 
+Consider the utterances
 ``it's raining,'' ``I ate some of the cookies,'' or ``can you close
@@
 
-But pragmatic reasoning often intersects with language learning as
-well. In both adults and children, uncertainty about word meanings is
+But pragmatic reasoning about meaning-in-context relies on stable literal meanings that must  themselves be learned.
+%and often intersects with learning the meaning of language in general. 
+In both adults and children, uncertainty about word meanings is
 common, and often considering speakers' pragmatic goals can help to resolve this uncertainty. For example,
@@
 Disambiguation without mapping & \cite{horst2008} &  & & & x \\
+Learning/lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
 Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
@@
 
-% agents not sure whether some literally means some-not-all or not,
-% could lexicalize given the right kind of input, but regular pragmatic
-% strengthening alone is not enough
+agents not sure whether some literally means some-not-all or not,
+could lexicalize given the right kind of input, but regular pragmatic
+strengthening alone is not enough
 

### 2013-05-31 praglearn.tex
cnclusion

--- 2013-05-31 praglearn.tex
@@
 
-Two puzzles present themselves to language users: What do words mean in general and what do they mean in context?
-%Natural language is a powerful tool for conveying meaning in context. 
-Consider the utterances
+Natural language is a powerful tool for conveying meaning in
+context. Consider the utterances
 ``it's raining,'' ``I ate some of the cookies,'' or ``can you close
@@
 
-But pragmatic reasoning about meaning-in-context relies on stable literal meanings that must  themselves be learned.
-%and often intersects with learning the meaning of language in general. 
-In both adults and children, uncertainty about word meanings is
+But pragmatic reasoning often intersects with language learning as
+well. In both adults and children, uncertainty about word meanings is
 common, and often considering speakers' pragmatic goals can help to resolve this uncertainty. For example,
@@
 \end{align*}
-Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
+Here $P(\lex|\text{$S$'s data}$ is defined similarly, when $S$ observes
 $L$'s interpretations of various utterances, and treats them as
@@
 Many sets of words in natural language form scales in which each term
-is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
+refers to successively stronger restrictions on the situations
+compatible with the term. ``Some'' and ``all'' form a scale of this
 type. While ``I ate some of the cookies'' is compatible with the
@@
     but not all}, and this is therefore the interpretation that a
-  rational listener will place on ``some'' as well. To simulate this
-  situation, we place a Dirichlet(10, 10) prior on the meaning of
-``some'' and a Dirichlet(1, 10) prior on the meaning of
-``some'', giving speaker probabilities of
-  $\left[
-    \begin{array}{cc}
-      1.00 & 0.00 \\
-      0.00 & 1.00
-    \end{array}
-  \right]$
-and listener probabilities of
-$\left[
-    \begin{array}{cc}
-      .86 & .14 \\
-      .03 & .97
-    \end{array}
-  \right]$. These revised interpretation probabilities minimize
-  ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.
+  rational listener will place on ``some'' as well.
+
+% To simulate this
+%   situation, we place a Dirichlet(10, 10) prior on the meaning of
+% ``some'' and a Dirichlet(1, 10) prior on the meaning of
+% ``some'', giving speaker probabilities of
+%   $\left[
+%     \begin{array}{cc}
+%       1.00 & 0.00 \\
+%       0.00 & 1.00
+%     \end{array}
+%   \right]$
+% and listener probabilities of
+% $\left[
+%     \begin{array}{cc}
+%       .86 & .14 \\
+%       .03 & .97
+%     \end{array}
+%   \right]$. These revised interpretation probabilities minimize
+%   ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.
 
@@
 
-If each agent in a multi-agent setting assumes the existence of a
-shared, conventional norm underlying everyone's behaviour, and further
-assumes that other agents are knowledgeable about this norm while they
-themselves are not, then we are able to simultaneously preserve most
-of the normative Bayesian framework, explicitly model learning and
-recursive social reasoning within the same framework, and avoid
-paradox. The resulting model unifies two previously separate strands
-of modelling work on language usage and acquisition, and -- having
-done so -- accounts for not just the phenomena these previous models
-accounted for, but a variety of new ones as well. We propose new
-explanations of the child language acquisition phenomena of
-disambiguation without mapping \cite{horst2008} and the acquisition of
-scalar adjectives (XX ask Mike what the levinson cite is really, see
-also table), and demonstrate that our model is capable of developing
+Language learners and language users must consider word meanings both
+within and across contexts. A critical part of this process is
+reasoning pragmatically about agents' goals in individual
+situations. In the current work we assume that agents communicating
+with one another assume that there is a shared conventional lexicon
+which they both rely on. They then
+reason recursively about how this lexicon should be used to convey
+particular meanings in context. This pair of assumptions allows us to
+create a model that unifies two previously separate strands
+of modeling work on language usage and acquisition, and account for a
+variety of new phenomena. In particular, we consider new
+explanations of disambiguation in early word learning and the acquisition of
+scalar adjectives, and demonstrate that our model is capable of developing
 novel and efficient communicative systems through iterated learning
@@
 
-Accomplishing this requires limited deviations from rationality, and
-as such necessarily introduces biases in agent behaviour. We've
-identified two such biases here: the tendency for pragmatic speakers
+Our work identifies two biases in communicators' behavior: a tendency for pragmatic speakers
 and listeners to accentuate useful, sparse patterns in their
-communicative systems (\S\ref{sec:emergence}), and for cheap words to
+communicative systems (\S\ref{sec:emergence}), and for short,
+high-frequency (``low cost'') expressions to
 be assigned to common objects (\S\ref{sec:horn-emergence}).
-Remarkably, while both are locally sub-optimal for individual agents,
+Remarkably, while both of these biases are locally sub-optimal for individual agents,
 they both systematically drive the overall communicative system
@@
 should leave their mark on the structure of the language itself, which
-may provide a partial explanation for how languages become so tightly
+may contribute to explaining how languages become 
 optimized for effective communication \cite{zipf1936,piantadosi2011}.

### 2013-05-31 praglearn.tex
conclusion

@@
 
-\title{Language learning via recursive pragmatic reasoning}
+\title{Language learning in context via recursive pragmatic reasoning}
 % \title{Learning words via recursive pragmatic reasoning about other agents}
@@
 be assigned to common objects (\S\ref{sec:horn-emergence}).
-Remarkably, while both of these biases are locally sub-optimal for individual agents,
-they both systematically drive the overall communicative system
+Both of these biases systematically drive the overall communicative system
 towards greater global efficiency. In the long term, these processes
@@
 optimized for effective communication \cite{zipf1936,piantadosi2011}.
+
+More generally, considering the interaction between pragmatics and learning may be a critical part of understanding human language.
+Our work here takes a first step towards unifying disparate strands of research that have treated language acquisition and language use as distinct.
 

### 2013-05-31 praglearn.tex
trivial changes

--- 2013-05-31 praglearn.tex
@@
 
-As soon as we attempt to add uncertainty about the meanings of words
+This resolves one paradox, but as soon as we attempt to add uncertainty about the meanings of words
 to such a model, new paradoxes arise. Suppose the listener is a young
@@
 \end{align*}
-The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
+The top-level speaker attempts to select the word which soft-maximizes
+their utility, with utility now being defined in terms of the
+informativity of the expectation (over lexicons) that the listener
+will have for the right referent\footnote{An alternative model would
+  have the speaker take the expectation over informativity, instead of
+  the informativity of the expectation.}:
 \begin{align*}
@@
 principle be compatible with eating {\sc all} of them, the listener is lead
-to believe that {\sc some-but-not-all}' is the likely state of
+to believe that {\sc some-but-not-all} is the likely state of
 affairs. We use the term {\it specificity implicature} here (following
@@
     but not all}, and this is therefore the interpretation that a
-  rational listener will place on ``some'' as well.
-
-% To simulate this
-%   situation, we place a Dirichlet(10, 10) prior on the meaning of
-% ``some'' and a Dirichlet(1, 10) prior on the meaning of
-% ``some'', giving speaker probabilities of
-%   $\left[
-%     \begin{array}{cc}
-%       1.00 & 0.00 \\
-%       0.00 & 1.00
-%     \end{array}
-%   \right]$
-% and listener probabilities of
-% $\left[
-%     \begin{array}{cc}
-%       .86 & .14 \\
-%       .03 & .97
-%     \end{array}
-%   \right]$. These revised interpretation probabilities minimize
-%   ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.
+  rational listener will place on ``some'' as well. To simulate this
+  situation, we place a Dirichlet(10, 10) prior on the meaning of
+``some'' and a Dirichlet(1, 10) prior on the meaning of
+``some'', giving speaker probabilities of
+  $\left[
+    \begin{array}{cc}
+      1.00 & 0.00 \\
+      0.00 & 1.00
+    \end{array}
+  \right]$
+and listener probabilities of
+$\left[
+    \begin{array}{cc}
+      .86 & .14 \\
+      .03 & .97
+    \end{array}
+  \right]$. These revised interpretation probabilities minimize
+  ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.
 
@@
 
-Language learners and language users must consider word meanings both
-within and across contexts. A critical part of this process is
-reasoning pragmatically about agents' goals in individual
-situations. In the current work we assume that agents communicating
-with one another assume that there is a shared conventional lexicon
-which they both rely on. They then
-reason recursively about how this lexicon should be used to convey
-particular meanings in context. This pair of assumptions allows us to
-create a model that unifies two previously separate strands
-of modeling work on language usage and acquisition, and account for a
-variety of new phenomena. In particular, we consider new
-explanations of disambiguation in early word learning and the acquisition of
-scalar adjectives, and demonstrate that our model is capable of developing
+If each agent in a multi-agent setting assumes the existence of a
+shared, conventional norm underlying everyone's behaviour, and further
+assumes that other agents are knowledgeable about this norm while they
+themselves are not, then we are able to simultaneously preserve most
+of the normative Bayesian framework, explicitly model learning and
+recursive social reasoning within the same framework, and avoid
+paradox. The resulting model unifies two previously separate strands
+of modelling work on language usage and acquisition, and -- having
+done so -- accounts for not just the phenomena these previous models
+accounted for, but a variety of new ones as well. We propose new
+explanations of the child language acquisition phenomena of
+disambiguation without mapping \cite{horst2008} and the acquisition of
+scalar adjectives (XX ask Mike what the levinson cite is really, see
+also table), and demonstrate that our model is capable of developing
 novel and efficient communicative systems through iterated learning
@@
 
-Our work identifies two biases in communicators' behavior: a tendency for pragmatic speakers
+Accomplishing this requires limited deviations from rationality, and
+as such necessarily introduces biases in agent behaviour. We've
+identified two such biases here: the tendency for pragmatic speakers
 and listeners to accentuate useful, sparse patterns in their
-communicative systems (\S\ref{sec:emergence}), and for short,
-high-frequency (``low cost'') expressions to
+communicative systems (\S\ref{sec:emergence}), and for cheap words to
 be assigned to common objects (\S\ref{sec:horn-emergence}).
-Remarkably, while both of these biases are locally sub-optimal for individual agents,
+Remarkably, while both are locally sub-optimal for individual agents,
 they both systematically drive the overall communicative system
@@
 should leave their mark on the structure of the language itself, which
-may contribute to explaining how languages become 
+may provide a partial explanation for how languages become so tightly
 optimized for effective communication \cite{zipf1936,piantadosi2011}.

### 2013-05-31 praglearn.tex
clarify wording in scalar implicature section

@@
 The recursive pragmatic reasoning portions of our model capture
-findings on specificity implicature. Consider a representation of
-scalar implicature as a matrix $\left[
+findings on specificity implicature. Consider a lexicon in which the
+rows represent the meaning of ``some'' and ``all'', and the columns
+indicate worlds where \textsc{some-but-not-all} and \textsc{all} are
+true, respectively: $\left[
     \begin{array}{cc}
@@
     \end{array}
-  \right]$, where rows show the words ``some'' and ``all'' and columns
-  indicate worlds where {\sc some but not all} and {\sc all} are true,
-  respectively, capturing the regularity that ``some'' can be said of
+  \right]$. This lexicon captures the regularity that ``some'' can be said of
   situations where {\sc all} is actually the case. Assuming uniform
@@
     but not all}, and this is therefore the interpretation that a
-  rational listener will place on ``some'' as well.
+  rational listener -- and our model -- will place on ``some'' as well.
 

### 2013-05-31 praglearn.tex
intro tweak

--- 2013-05-31 praglearn.tex
@@
 language learning, with the hope of capturing phenomena in both
-domains. We begin by describing previous models in each domain as well
-as proposals for bringing the two together. We next simulate findings
+domains.
+We begin by describing a proposal for bringing the two together, noting several issues in applying a purely normative framework to recursive reasoning under uncertainty.
+We next simulate findings
 on pragmatic inference in one-shot games (replicating previous
@@
 uncertain about the lexicon and in iterated communication games where both
-communicators are uncertain about the lexicon. We end by discussing the implications of our results for
-work on language change.
+communicators are uncertain about the lexicon.
 
@@
 
-This resolves one paradox, but as soon as we attempt to add uncertainty about the meanings of words
+As soon as we attempt to add uncertainty about the meanings of words
 to such a model, new paradoxes arise. Suppose the listener is a young
@@
 \end{align*}
-The top-level speaker attempts to select the word which soft-maximizes
-their utility, with utility now being defined in terms of the
-informativity of the expectation (over lexicons) that the listener
-will have for the right referent\footnote{An alternative model would
-  have the speaker take the expectation over informativity, instead of
-  the informativity of the expectation.}:
+The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
 \begin{align*}
@@
 principle be compatible with eating {\sc all} of them, the listener is lead
-to believe that {\sc some-but-not-all} is the likely state of
+to believe that {\sc some-but-not-all}' is the likely state of
 affairs. We use the term {\it specificity implicature} here (following

### 2013-05-31 praglearn.tex
remove a line from 3.2, putting us back under the page limit

--- 2013-05-31 praglearn.tex
@@
 is to know about each other's behavior. However, the core result
-carries over to our model. For example, with a uniform prior over
+carries over to our model. E.g., with a uniform prior over
 lexicons and $P_\text{prior}(\text{\textsc{common}}) = 0.8$,
@@
 0.77, P_L(\text{\textsc{common}} | \text{``expensive''}) =
-0.65$). This preference is somewhat weak; the critical observation is
+0.65$). This preference is weak; the critical point is
 that it exists at all, given the unbiased priors. We return to this
-point in \S\ref{sec:horn-emergence}.
+\S\ref{sec:horn-emergence}.
 

### 2013-05-31 praglearn.tex
edit

--- 2013-05-31 praglearn.tex
@@
 
-\title{Language learning in context via recursive pragmatic reasoning}
+\title{Language learning via recursive pragmatic reasoning}
 % \title{Learning words via recursive pragmatic reasoning about other agents}
@@
 language learning, with the hope of capturing phenomena in both
-domains.
-We begin by describing a proposal for bringing the two together, noting several issues in applying a purely normative framework to recursive reasoning under uncertainty.
-We next simulate findings
+domains. We begin by describing previous models in each domain as well
+as proposals for bringing the two together. We next simulate findings
 on pragmatic inference in one-shot games (replicating previous
@@
 uncertain about the lexicon and in iterated communication games where both
-communicators are uncertain about the lexicon.
+communicators are uncertain about the lexicon. We end by discussing the implications of our results for
+work on language change.
 
@@
 however: Since both agents know perfectly well that there is no
-underlying convention to discover, there is nothing to learn from the
+existing convention to discover, there is nothing to learn from the
 other's behavior. Furthermore, even if only one partner is producing
@@
 fact using the good system all along. As a result, dyads in this game
-not only end up converging onto a stable system at a rate far above
-chance (Fig.~\ref{fig:horn}.
+ end up converging onto a stable system at a rate far above
+chance (Fig.~\ref{fig:horn}).
 
@@
 be assigned to common objects (\S\ref{sec:horn-emergence}).
-Both of these biases systematically drive the overall communicative system
+Remarkably, while both of these biases are locally sub-optimal for individual agents,
+they both systematically drive the overall communicative system
 towards greater global efficiency. In the long term, these processes
@@
 optimized for effective communication \cite{zipf1936,piantadosi2011}.
-
-More generally, considering the interaction between pragmatics and learning may be a critical part of understanding human language.
-Our work here takes a first step towards unifying disparate strands of research that have treated language acquisition and language use as distinct.
 

### 2013-05-31 praglearn.tex
tweak

@@
 \begin{align*}
-  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
+  P_S(\word & | \obj, \text{$S$'s data})  \propto \\
+ &\exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
 \end{align*}
@@
     but not all}, and this is therefore the interpretation that a
-  rational listener -- and our model -- will place on ``some'' as well.
+  rational listener---and our model---will place on ``some'' as well.
 
@@
 %        [ 0.03151434,  0.96848566]])
+
+\vspace{-2ex}
 
@@
 however: Since both agents know perfectly well that there is no
-existing convention to discover, there is nothing to learn from the
+underlying convention to discover, there is nothing to learn from the
 other's behavior. Furthermore, even if only one partner is producing
@@
 fact using the good system all along. As a result, dyads in this game
- end up converging onto a stable system at a rate far above
-chance (Fig.~\ref{fig:horn}).
+not only end up converging onto a stable system at a rate far above
+chance (Fig.~\ref{fig:horn}.
 

### 2013-05-31 praglearn.tex
tweak fig. 2 caption

--- 2013-05-31 praglearn.tex
@@
   above. In the first run, speaker and listener converge on a sparse
-  and  efficient communicative equilibrium, while in the second they reach
-  a sub-optimal equilibrium.}
+  and  efficient communicative equilibrium, in which ``cheap'' means
+  \textsc{COMMON} and ``expensive'' means \textsc{RARE}, while in the
+  second they reach a sub-optimal equilibrium. As shown in
+  Fig.~\ref{fig:emergence-average}, the former is more typical.}
 \end{SCfigure}

### 2013-05-31 figures/non-emergence.py
missing space

@@ -51,11 +51,13 @@ def simulate_non_pragmatic(r, turns, prior=None):
     for i in xrange(turns):
         for S_alpha, L_alpha in [(a1_alpha, a2_alpha),
                                  (a2_alpha, a1_alpha)]:
-        target = r.randint(SIZE)
-        S_dist = marginal_S_dist(speaker_posterior)
-        L_dist = marginal_S_dist(listener_posterior)
-        L_dist -= np.logaddexp.reduce(L_dist, axis=1)[:, np.newaxis]
+            S_dist = marginal_S_dist(S_alpha)
+            L_dist = L_dist_for(marginal_S_dist(L_alpha))
+            target = r.randint(SIZE)
+            word = conpact2.weighted_choice(r, np.exp(S_dist[:, target]))
+            L_alpha[word, target] += 1
 
+            P_understood =
 
     # Want out:
     #   P(understood)

### 2013-05-31 praglearn.tex
missing space
tewaks

--- 2013-05-31 praglearn.tex
@@
 Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
-Cross-situational learning & \cite{smith2008} & x& & & x \\
 Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
@@
 (less costly) expression, ``Lee stopped the car''
-\cite{horn1984}. Surprisingly, Bergen et al \cite{bergen2012} show
+\cite{horn1984}. Surprisingly, Bergen et al. \cite{bergen2012} show
 that the key to achieving this favorable result is ignorance. If a
@@
 %        [ 0.64702551,  0.35297449]])
+
+%% NOTE FOR REVISION: THIS BIT HERE ABOUT BERGEN MAKES NOAH SAD AND SHOULD BE RECONSIDERED.
 
@@
 
-\section{The role of pragmatics in language acquisition}
+\section{Pragmatics in learning from a knowledgable speaker}
 
@@
 
-\section{Mutual learning situations}
+\section{Pragmatic reasoning in the absence of conventional meanings}
 
@@
 reasoning pragmatically about agents' goals in individual
-situations. In the current work we assume that agents communicating
-with one another assume that there is a shared conventional lexicon
+situations. In the current work we treat agents communicating
+with one another as assuming that there is a shared conventional lexicon
 which they both rely on. They then
@@
 create a model that unifies two previously separate strands
-of modeling work on language usage and acquisition, and account for a
+of modeling work on language usage and acquisition and account for a
 variety of new phenomena. In particular, we consider new

### 2013-05-31 praglearn.tex

@@
 observed by considering the Horn implicatures introduced in \S\ref{sec:horn-implicature}. 
-In these games, pragmatic reasoning causes each agent to start out weakly biasedtowards the associations ``cheap'' $\leftrightarrow$
+In these games, pragmatic reasoning causes each agent to start out
+weakly biased towards the associations ``cheap'' $\leftrightarrow$
 \textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}; this bias arises because their prior on literal meaning

### 2013-05-31 praglearn.tex
tiny tweaks to 5.2

--- 2013-05-31 praglearn.tex
@@
 the implicature disappears, to be replaced by a belief that
-``cheap'' \textit{literally} means \textsc{common}. To demonstrate
+``cheap'' \textit{literally} means \textsc{common} (and did all along). To demonstrate
 this phenomenon, we queried each agent in each dyad about how they would refer to
@@
 this implicature-free case, while less than 20\% believed in the
-``bad'' mapping (with the rest being inconsistent). This makes the
+``bad'' mapping (with the rest being inconsistent). Our model therefore makes the
 interesting prediction that while specificity implicatures can survive

### 2017-01-12 paper/NIPS reviewer response.txt
final cleanup

We see three main critiques raised by the reviewers:

1. Use of linguistic terminology was excessive or unclear.

These comments are extremely helpful for us in reaching the broad audience that we aim for in this work. If accepted we will be sure to take them into account when revising.

2. It would be more realistic to allow a richer model of literal meaning, allowing for multi-word phrases (Reviewer 5) and unbounded numbers of words and objects via non-parametric methods (Reviewer 3).

We agree completely with this comment. Both of these extensions are part of our larger research agenda (and unpublished work has investigated both). With respect to the current paper, however, we were attempting to strike a balance between clarity and sophistication. We see the heart of our work as the discussion of the contradictions between learning models and pragmatic reasoning models (sec. 2.1), and our proposal for resolving them (sec. 2.2). Right now the core literal learning part of our model (line 90) uses simple word/object associations, but in principle it would be straightforward to replace this component with any standard Bayesian language learning model (e.g. ref [25] or Xu & Tenenbaum 2007; see also our footnote 2). When integrated into the full model, though, even the current word/object associations end up producing rich and counter-intuitive behavior -- enough so that the bulk of the paper is taken up with explaining and demonstrating this behavior through simulations (sec. 3-5).

Using a more realistic model would add more complexity to be explained -- in fact, our current implementation does model multi-word utterances, but we disabled this for the current report because it didn't change any qualitative results. In addition, a more complex learning model would make it more difficult to isolate the effects of what we view as the primary contribution, the pragmatics/learning combination. For example, more realistic language learning models generally incorporate some sort of sparseness prior. We find that our model produces sparse lexicons (sec 5.1), and that it does so even when using our simple model which has no sparseness prior. This demonstrates that this bias is a (non-obvious) side-effect of the pragmatics/learning interplay. So we feel that establishing the behavior of this simple model is a necessary precondition to understanding or explaining the behavior of the more realistic models that will follow.

3. Limited number of simulations in the manuscript (Reviewer 5).

Again, we feel that there is a tradeoff between complexity and clarity in our current work. While it would be possible to run simulations with more words and objects, these simulations would go beyond the human experimental literature and might not provide insights into the behavior of the model with respect to the phenomena of interest. (We recognize that this may be a disciplinary issue, given that it is unusual in machine learning to see so much attention given to such small scale problems).

We would also like to clarify the simulation content of the current manuscript. Sections 3-5 present the results of six qualitatively different simulation studies. Figs. 1 & 2 present individual simulation runs for illustration, but population statistics for batch runs are given in Fig. 3. In future we plan to run human experiments to test some of the hypotheses generated by these simulations (e.g., the explanation of
disambiguation without mapping given in lines 292-305, and the discrepancy between Horn and specificity implicatures with regard to iterated learning, lines 393-404). We also plan to expand the model to handle multi-word sentences. Both of these moves will provide new datasets for future work.

Again, we appreciate the thoughtful reviews. Responses to some specific queries:

Reviewer 3: What we mean by a "novel" lexicon is that our agents start jointly using lexicons that previously, no agent was using. What we mean by "efficient" is that this lexicon selection process is systematically biased towards choosing "good" ones (e.g. in the Horn implicature case, Fig. 3 right-hand side). Ref [11], by comparison, considers how to use an existing lexicon efficiently, which is a different thing. (Our model reduces to the model of [11] in the case where the lexicon is shared between communicators and known a priori.) Thank you for the suggestion about Table 1, we will revise accordingly.

Reviewer 5: In all cases, we represent the lexicon as a matrix of numbers between 0 and 1, with each row constrained to sum to 1. In the Horn implicature cases, our prior is simply a uniform distribution over such matrices. In the specificity implicature case, we use a prior which encodes the knowledge that the word "all" refers to the ALL object but not the SOME-BUT-NOT-ALL object (a Dirichlet with pseudocounts favoring ALL), and place a uniform prior on the meanings of "some." Thank you for the suggestion of a heuristic comparison model--we will investigate this in our future work.

Reviewer 7: There is substantial empirical evidence that speakers adjust what words they're using based on feedback from their listeners (e.g. ref. [8]). On the other hand, it is unknown whether this behavior is optimal because ours is (to our knowledge) the first quantitative investigation of optimal behavior, so there hasn't been any way to check! Regarding the meaning of "some", there is a large body of research on children's quantifier use (reviewed in ref. [2]). To summarize: children begin by making systematic failures to reason that "some" implies SOME-BUT-NOT-ALL. (In the words of one author, they seem "more literal" than adults). Only around age five or older do they make this and other scalar implicatures correctly. So our simulations suggesting maintenance of the literal meaning (sec. 4.2) are at least prima facie consistent with the developmental literature.
